---
title: "All nPOD Models"
output: html_document
date: "2026-03-28"
---

Family choices informed by `model_diagnostics.Rmd` and `count_nb_comparison.Rmd`:

- **Density**: Gamma(log) — dispersion 1.03, no issues
- **Count.ins**: NB1 — AIC winner by 80K over NB2; DHARMa underdispersion (0.014)
  is conservative (SEs too wide), not a misspecification
- **Count.glu**: ZINB1 (zi ~ Diagnosis) — significant ZI, ΔAIC = -5170 vs NB1
- **Count.soma**: NB1 — dispersion 0.88, fine
- **Count.PP**: NB1 — dispersion 1.05, fine
- **Islet.Cells**: NB2 — AIC winner; mild overdispersion (1.36); ZINB worse by +4 AIC
- **All total area**: Gaussian on log(area) — LogNorm dispersion 0.84-1.01 everywhere;
  Gamma catastrophically underdispersed for ins (0.005-0.11) and overdispersed
  for pp (2.8-29); raw Gaussian sometimes fails to converge
- **All per-cell area**: Gaussian on log(area/count) — AIC winner by 250K–5M over
  Gamma and lognormal in all 12 cell × subset combinations. Insulin multi-cell/islet
  subsets show conservative underdispersion (0.45–0.55), same pattern as insulin
  count models. Informed by `per_cell_area_diagnostics.Rmd`.

```{r setup, include=FALSE}
library(tidyverse)
library(glmmTMB)
library(splines)

for (d in c("Density/Models", "Count/Models", "Area/Models/Total_Area/", "Area/Models/Per_Cell_Area/")) {
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
}
```

```{r data-loading}
# ==========================================================================
# Load ALL pre-built datasets from Create_unified_dataset.Rmd
# Each dataset already has correct centering for its subset
# ==========================================================================

# --- Core datasets ---
quad_data        <- readRDS("Data/quad_data.rds")
quad_t1d         <- readRDS("Data/quad_data_t1d.rds")
cutoff_data      <- readRDS("Data/cutoff_data.rds")
density_dataset  <- readRDS("Data/density_data.rds")
density_t1d      <- readRDS("Data/density_data_t1d.rds")

# --- Ratio datasets (already filtered: no Single stain, no Donor 6473, no zero denom) ---
ratio_data       <- readRDS("Data/ratio_data.rds")
ratio_data_t1d   <- readRDS("Data/ratio_data_t1d.rds")

# --- Area datasets by size category ---
area_data_joined <- readRDS("Data/area_data.rds")
area_t1d         <- readRDS("Data/area_data_t1d.rds")

area_singlets      <- readRDS("Data/area_data_singlets.rds")
area_singlets_t1d  <- readRDS("Data/area_data_singlets_t1d.rds")
area_endobs        <- readRDS("Data/area_data_endobs.rds")
area_endobs_t1d    <- readRDS("Data/area_data_endobs_t1d.rds")
area_islets        <- readRDS("Data/area_data_islets.rds")
area_islets_t1d    <- readRDS("Data/area_data_islets_t1d.rds")

# --- Area combined datasets (logIC zeroed for singletons, per cell type) ---
beta_combined      <- readRDS("Data/area_data_beta_combined.rds")
beta_combined_t1d  <- readRDS("Data/area_data_beta_combined_t1d.rds")
alpha_combined     <- readRDS("Data/area_data_alpha_combined.rds")
alpha_combined_t1d <- readRDS("Data/area_data_alpha_combined_t1d.rds")
delta_combined     <- readRDS("Data/area_data_delta_combined.rds")
delta_combined_t1d <- readRDS("Data/area_data_delta_combined_t1d.rds")
pp_combined        <- readRDS("Data/area_data_pp_combined.rds")
pp_combined_t1d    <- readRDS("Data/area_data_pp_combined_t1d.rds")
```

```{r data-prep}
# ==========================================================================
# Factor levels and object_type — centering already done in data creation
# ==========================================================================

set_factors <- function(df) {
  df %>% mutate(
    Diagnosis = factor(Diagnosis, levels = c("ND", "T1D")),
    Region    = factor(Region, levels = c("Head", "Body", "Tail")),
    Sex       = factor(Sex, levels = c("Female", "Male")),
    Donor     = factor(Donor),
    ImageID   = factor(ImageID))
}
set_factors_no_image <- function(df) {
  df %>% mutate(
    Diagnosis = factor(Diagnosis, levels = c("ND", "T1D")),
    Region    = factor(Region, levels = c("Head", "Body", "Tail")),
    Sex       = factor(Sex, levels = c("Female", "Male")),
    Donor     = factor(Donor))
}

quad_data <- set_factors(quad_data)
quad_t1d  <- set_factors(quad_t1d) %>% droplevels()
cutoff_data <- set_factors(cutoff_data)
ratio_data <- set_factors(ratio_data)
ratio_data_t1d <- set_factors(ratio_data_t1d) %>% droplevels()
density_dataset <- set_factors_no_image(density_dataset)
density_t1d <- set_factors_no_image(density_t1d) %>% droplevels()

area_data_joined <- set_factors(area_data_joined)
area_t1d <- set_factors(area_t1d) %>% droplevels()
area_singlets <- set_factors(area_singlets)
area_singlets_t1d <- set_factors(area_singlets_t1d) %>% droplevels()
area_endobs <- set_factors(area_endobs)
area_endobs_t1d <- set_factors(area_endobs_t1d) %>% droplevels()
area_islets <- set_factors(area_islets)
area_islets_t1d <- set_factors(area_islets_t1d) %>% droplevels()
beta_combined <- set_factors(beta_combined)
beta_combined_t1d <- set_factors(beta_combined_t1d) %>% droplevels()
alpha_combined <- set_factors(alpha_combined)
alpha_combined_t1d <- set_factors(alpha_combined_t1d) %>% droplevels()
delta_combined <- set_factors(delta_combined)
delta_combined_t1d <- set_factors(delta_combined_t1d) %>% droplevels()
pp_combined <- set_factors(pp_combined)
pp_combined_t1d <- set_factors(pp_combined_t1d) %>% droplevels()

add_object_type <- function(df) {
  df %>% mutate(
    object_type = factor(case_when(
      Islet.Cells == 1 ~ "1 cell",
      Islet.Cells >= 2 & Islet.Cells <= 14 ~ "2-14 Cells",
      Islet.Cells >= 15 ~ "15+ Cells"),
      levels = c("1 cell", "2-14 Cells", "15+ Cells")))
}
area_data_joined   <- add_object_type(area_data_joined)
area_t1d           <- add_object_type(area_t1d)
beta_combined      <- add_object_type(beta_combined)
beta_combined_t1d  <- add_object_type(beta_combined_t1d)
alpha_combined     <- add_object_type(alpha_combined)
alpha_combined_t1d <- add_object_type(alpha_combined_t1d)
delta_combined     <- add_object_type(delta_combined)
delta_combined_t1d <- add_object_type(delta_combined_t1d)
pp_combined        <- add_object_type(pp_combined)
pp_combined_t1d    <- add_object_type(pp_combined_t1d)

# --- Centering verification ---
cat("=== Centering verification (all should be ~0) ===\n")
check_centering <- function(nm, df) {
  cat(sprintf("%-22s  Age_c=%+.4f  logIC_c=%+.4f  DD_c=%+.4f  onset_c=%+.4f\n",
    nm, mean(df$Age_c, na.rm=TRUE),
    if("log_Islet.Cells_c" %in% names(df)) mean(df$log_Islet.Cells_c, na.rm=TRUE) else NA,
    mean(df$Disease.Duration_c, na.rm=TRUE), mean(df$age_at_onset_c, na.rm=TRUE)))
}
for (nm in c("quad_data","quad_t1d","ratio_data","ratio_data_t1d","density_dataset","density_t1d",
             "area_singlets","area_singlets_t1d","area_endobs","area_endobs_t1d",
             "area_islets","area_islets_t1d","beta_combined","beta_combined_t1d")) {
  check_centering(nm, get(nm))
}
```

# Density

## Diagnosis

```{r density-diag}
density_diag <- glmmTMB(islet_density ~ (Diagnosis + Region + Sex + Age_c)^2 +
                          Diagnosis:Region:Age_c + (1|Donor), 
                        data = density_dataset, 
                        family = Gamma(link = "log"))
saveRDS(density_diag, "Density/Models/density_diag.rds"); rm(density_diag); gc()
```

## Age at Onset

```{r density-ao}
density_mod_t1d <- glmmTMB(islet_density ~ (Disease.Duration_c + Region + Sex + age_at_onset_c)^2 + 
                           Disease.Duration_c:Region:age_at_onset_c + (1|Donor), 
                         data = density_t1d, 
                         family = Gamma(link = "log"))
saveRDS(density_mod_t1d, "Density/Models/density_t1d.rds"); rm(density_mod_t1d); gc()
```

## Plotting Helper

```{r density-plot-helpers}
density_spline <- glmmTMB(islet_density ~ ns(Disease.Duration_c, df = 3) + Age_c + Sex + Region + 
                          (1|Donor), 
                          family = Gamma(link = "log"), 
                          data = density_t1d)
saveRDS(density_spline, "Density/Models/density_spline.rds"); rm(density_spline); gc()
```

# Proportion & Ratio Models

Run on HTC cluster:

- `proportion_diagnosis_HTC.R` (12 ordbetareg: 4 cell types × powered/mid/full)
- `proportion_t1d_HTC.R` (8 ordbetareg: 4 cell types × DD/AO)
- `ratio_diagnosis_HTC.R` (3 ordbetareg: powered/mid/full)
- `ratio_t1d_HTC.R` (2 ordbetareg: DD/AO)

# Count Models

## Diagnosis

```{r count-diag}
# Count.ins — NB1 (AIC winner by 80K; underdispersion = conservative SEs)
count_ins <- glmmTMB(Count.ins ~ (Diagnosis + Region + Sex + Age_c + log_Islet.Cells_c)^2 +
                       Diagnosis:Region:log_Islet.Cells_c + Diagnosis:Region:Age_c +
                       (1|Donor) + (1|Donor:ImageID), 
                     data = quad_data, 
                     family = nbinom1)
saveRDS(count_ins, "Count/Models/count_ins.rds"); rm(count_ins); gc()

# Count.glu — ZINB1, zi ~ Diagnosis (ΔAIC = -5170 vs NB1)
count_glu <- glmmTMB(Count.glu ~ (Diagnosis + Region + Sex + Age_c + log_Islet.Cells_c)^2 +
                       Diagnosis:Region:log_Islet.Cells_c + Diagnosis:Region:Age_c +
                       (1|Donor) + (1|Donor:ImageID), 
                     data = quad_data, 
                     family = nbinom1, 
                     ziformula = ~ Diagnosis)
saveRDS(count_glu, "Count/Models/count_glu.rds"); rm(count_glu); gc()

# Count.soma — NB1 (disp = 0.88)
count_soma <- glmmTMB(Count.soma ~ (Diagnosis + Region + Sex + Age_c + log_Islet.Cells_c)^2 +
                        Diagnosis:Region:log_Islet.Cells_c + Diagnosis:Region:Age_c +
                        (1|Donor) + (1|Donor:ImageID), 
                      data = quad_data, 
                      family = nbinom1)
saveRDS(count_soma, "Count/Models/count_soma.rds"); rm(count_soma); gc()

# Count.PP — NB1 (disp = 1.05)
count_pp <- glmmTMB(Count.PP ~ (Diagnosis + Region + Sex + Age_c + log_Islet.Cells_c)^2 +
                      Diagnosis:Region:log_Islet.Cells_c + Diagnosis:Region:Age_c +
                      (1|Donor) + (1|Donor:ImageID), 
                    data = quad_data, 
                    family = nbinom1)
saveRDS(count_pp, "Count/Models/count_pp.rds"); rm(count_pp); gc()

# Islet.Cells — NB2 (AIC winner; disp = 1.36; ZINB worse by +4)
count_islet <- glmmTMB(Islet.Cells ~ (Diagnosis + Region + Sex + Age_c)^2 +
                         Diagnosis:Region:Age_c + (1|Donor) + (1|Donor:ImageID),
                       data = quad_data, 
                       family = nbinom2)
saveRDS(count_islet, "Count/Models/count_islet.rds"); rm(count_islet); gc()
```

## T1D

```{r count-t1d}
count_ins_t1d <- glmmTMB(Count.ins ~ (Disease.Duration_c + Region + Sex + age_at_onset_c + log_Islet.Cells_c)^2 +
                             Disease.Duration_c:Region:log_Islet.Cells_c + Disease.Duration_c:Region:age_at_onset_c +
                             (1|Donor) + (1|Donor:ImageID), 
                           data = quad_t1d, 
                           family = nbinom1)
saveRDS(count_ins_t1d, "Count/Models/count_ins_t1d.rds"); rm(count_ins_t1d); gc()

count_glu_t1d <- glmmTMB(Count.glu ~ (Disease.Duration_c + Region + Sex + age_at_onset_c + log_Islet.Cells_c)^2 +
                             Disease.Duration_c:Region:log_Islet.Cells_c + Disease.Duration_c:Region:age_at_onset_c +
                             (1|Donor) + (1|Donor:ImageID), 
                           data = quad_t1d, 
                           family = nbinom1, 
                           ziformula = ~ 1)
saveRDS(count_glu_t1d, "Count/Models/count_glu_t1d.rds"); rm(count_glu_t1d); gc()

count_soma_t1d <- glmmTMB(Count.soma ~ (Disease.Duration_c + Region + Sex + age_at_onset_c + log_Islet.Cells_c)^2 +
                              Disease.Duration_c:Region:log_Islet.Cells_c + Disease.Duration_c:Region:age_at_onset_c +
                              (1|Donor) + (1|Donor:ImageID), 
                            data = quad_t1d, 
                            family = nbinom1)
saveRDS(count_soma_t1d, "Count/Models/count_soma_t1d.rds"); rm(count_soma_t1d); gc()

count_pp_t1d <- glmmTMB(Count.PP ~ (Disease.Duration_c + Region + Sex + age_at_onset_c + log_Islet.Cells_c)^2 +
                            Disease.Duration_c:Region:log_Islet.Cells_c + Disease.Duration_c:Region:age_at_onset_c +
                            (1|Donor) + (1|Donor:ImageID), 
                          data = quad_t1d, 
                          family = nbinom1)
saveRDS(count_pp_t1d, "Count/Models/count_pp_t1d.rds"); rm(count_pp_t1d); gc()

count_islet_t1d <- glmmTMB(Islet.Cells ~ (Disease.Duration_c + Region + Sex + age_at_onset_c)^2 +
                               Disease.Duration_c:Region:age_at_onset_c + (1|Donor) + (1|Donor:ImageID),
                             data = quad_t1d, 
                             family = nbinom2)
saveRDS(count_islet_t1d, "Count/Models/count_islet_t1d.rds"); rm(count_islet_t1d); gc()
```

# Total Area Models — All Gaussian on log(area)

```{r area-helpers}
# In all_npod_models_c.Rmd area-helpers chunk
area_to_count <- c(ins_area = "Count.ins", glu_area = "Count.glu",
                   soma_area = "Count.soma", pp_area = "Count.PP")

split_area <- function(data, cell_col) {
  log_col   <- paste0("log_", cell_col)
  count_col <- area_to_count[[cell_col]]
  data %>%
    filter(!is.na(!!sym(cell_col)) & !!sym(cell_col) > 0 &
           !!sym(count_col) > 0) %>%
    mutate(!!log_col := log(!!sym(cell_col)))
}

# --- Diagnosis formulas ---
area_full_formula <- function(log_out) {
  as.formula(paste0(log_out, " ~ (Diagnosis + Region + Sex + Age_c + log_Islet.Cells_c)^2 +
    Diagnosis:Region:log_Islet.Cells_c + Diagnosis:Region:Age_c +
    (1|Donor) + (1|Donor:ImageID)"))
}
area_reduced_formula <- function(log_out) {
  as.formula(paste0(log_out, " ~ (Diagnosis + Region + Sex + Age_c)^2 +
    Diagnosis:Region:Age_c + (1|Donor) + (1|Donor:ImageID)"))
}
area_seo_formula <- function(log_out) {
  as.formula(paste0(log_out, " ~ (Diagnosis + Region + Sex + Age_c)^2 +
    Diagnosis:Region:Age_c + log_Islet.Cells_c +
    (1|Donor) + (1|Donor:ImageID)"))
}

# --- T1D formulas ---
area_t1d_full_formula <- function(log_out) {
  as.formula(paste0(log_out, " ~ (Disease.Duration_c + Region + Sex + age_at_onset_c + log_Islet.Cells_c)^2 +
    Disease.Duration_c:Region:log_Islet.Cells_c + Disease.Duration_c:Region:age_at_onset_c +
    (1|Donor) + (1|Donor:ImageID)"))
}
area_t1d_reduced_formula <- function(log_out) {
  as.formula(paste0(log_out, " ~ (Disease.Duration_c + Region + Sex + age_at_onset_c)^2 +
    Disease.Duration_c:Region:age_at_onset_c + (1|Donor) + (1|Donor:ImageID)"))
}
area_t1d_seo_formula <- function(log_out) {
  as.formula(paste0(log_out, " ~ (Disease.Duration_c + Region + Sex + age_at_onset_c)^2 +
    Disease.Duration_c:Region:age_at_onset_c + log_Islet.Cells_c +
    (1|Donor) + (1|Donor:ImageID)"))
}

# --- Per-cell area helper ---
# Divides total cell-type area by cell-type count, then log-transforms
split_per_cell <- function(data, cell_col) {
  count_col <- area_to_count[[cell_col]]
  log_col   <- paste0("log_", cell_col, "_pc")
  data %>%
    filter(!is.na(!!sym(cell_col)) & !!sym(cell_col) > 0 &
           !!sym(count_col) > 0) %>%
    mutate(!!log_col := log(!!sym(cell_col) / !!sym(count_col)))
}

# Per-cell area formulas (identical structure to total area formulas)
pc_full_formula <- function(log_out) {
  as.formula(paste0(log_out, " ~ (Diagnosis + Region + Sex + Age_c + log_Islet.Cells_c)^2 +
    Diagnosis:Region:log_Islet.Cells_c + Diagnosis:Region:Age_c +
    (1|Donor) + (1|Donor:ImageID)"))
}
pc_reduced_formula <- function(log_out) {
  as.formula(paste0(log_out, " ~ (Diagnosis + Region + Sex + Age_c)^2 +
    Diagnosis:Region:Age_c + (1|Donor) + (1|Donor:ImageID)"))
}
pc_seo_formula <- function(log_out) {
  as.formula(paste0(log_out, " ~ (Diagnosis + Region + Sex + Age_c)^2 +
    Diagnosis:Region:Age_c + log_Islet.Cells_c +
    (1|Donor) + (1|Donor:ImageID)"))
}
pc_t1d_full_formula <- function(log_out) {
  as.formula(paste0(log_out, " ~ (Disease.Duration_c + Region + Sex + age_at_onset_c + log_Islet.Cells_c)^2 +
    Disease.Duration_c:Region:log_Islet.Cells_c + Disease.Duration_c:Region:age_at_onset_c +
    (1|Donor) + (1|Donor:ImageID)"))
}
pc_t1d_reduced_formula <- function(log_out) {
  as.formula(paste0(log_out, " ~ (Disease.Duration_c + Region + Sex + age_at_onset_c)^2 +
    Disease.Duration_c:Region:age_at_onset_c + (1|Donor) + (1|Donor:ImageID)"))
}
pc_t1d_seo_formula <- function(log_out) {
  as.formula(paste0(log_out, " ~ (Disease.Duration_c + Region + Sex + age_at_onset_c)^2 +
    Disease.Duration_c:Region:age_at_onset_c + log_Islet.Cells_c +
    (1|Donor) + (1|Donor:ImageID)"))
}
```

## Diagnosis — Singlets (Total Area)

```{r area-diag-singlets}
beta_1cell_data  <- split_area(area_singlets, "ins_area")
alpha_1cell_data <- split_area(area_singlets, "glu_area")
delta_1cell_data <- split_area(area_singlets, "soma_area")
pp_1cell_data    <- split_area(area_singlets, "pp_area")

area_beta_1cell <- glmmTMB(area_reduced_formula("log_ins_area"), 
                           data = beta_1cell_data, 
                           family = gaussian())
saveRDS(area_beta_1cell, "Area/Models/Total_Area/area_beta_1cell.rds"); rm(area_beta_1cell); gc()

area_alpha_1cell <- glmmTMB(area_reduced_formula("log_glu_area"), 
                            data = alpha_1cell_data, 
                            family = gaussian())
saveRDS(area_alpha_1cell, "Area/Models/Total_Area/area_alpha_1cell.rds"); rm(area_alpha_1cell); gc()

area_delta_1cell <- glmmTMB(area_reduced_formula("log_soma_area"), 
                            data = delta_1cell_data, 
                            family = gaussian())
saveRDS(area_delta_1cell, "Area/Models/Total_Area/area_delta_1cell.rds"); rm(area_delta_1cell); gc()

area_pp_1cell <- glmmTMB(area_reduced_formula("log_pp_area"), 
                         data = pp_1cell_data, 
                         family = gaussian())
saveRDS(area_pp_1cell, "Area/Models/Total_Area/area_pp_1cell.rds"); rm(area_pp_1cell); gc()
```

## Diagnosis — Small EOs (Total Area)

```{r area-diag-smallEO}
beta_2to14_data  <- split_area(area_endobs, "ins_area")
alpha_2to14_data <- split_area(area_endobs, "glu_area")
delta_2to14_data <- split_area(area_endobs, "soma_area")
pp_2to14_data    <- split_area(area_endobs, "pp_area")

area_beta_2to14 <- glmmTMB(area_seo_formula("log_ins_area"), 
                           data = beta_2to14_data, 
                           family = gaussian())
saveRDS(area_beta_2to14, "Area/Models/Total_Area/area_beta_2to14.rds"); rm(area_beta_2to14); gc()

area_alpha_2to14 <- glmmTMB(area_seo_formula("log_glu_area"), 
                            data = alpha_2to14_data, 
                            family = gaussian())
saveRDS(area_alpha_2to14, "Area/Models/Total_Area/area_alpha_2to14.rds"); rm(area_alpha_2to14); gc()

area_delta_2to14 <- glmmTMB(area_seo_formula("log_soma_area"), 
                            data = delta_2to14_data, 
                            family = gaussian())
saveRDS(area_delta_2to14, "Area/Models/Total_Area/area_delta_2to14.rds"); rm(area_delta_2to14); gc()

area_pp_2to14 <- glmmTMB(area_seo_formula("log_pp_area"), 
                         data = pp_2to14_data, 
                         family = gaussian())
saveRDS(area_pp_2to14, "Area/Models/Total_Area/area_pp_2to14.rds"); rm(area_pp_2to14); gc()
```

## Diagnosis — Islets (Total Area)

```{r area-diag-islets}
beta_15plus_data  <- split_area(area_islets, "ins_area")
alpha_15plus_data <- split_area(area_islets, "glu_area")
delta_15plus_data <- split_area(area_islets, "soma_area")
pp_15plus_data    <- split_area(area_islets, "pp_area")

area_beta_15plus <- glmmTMB(area_full_formula("log_ins_area"), 
                            data = beta_15plus_data, 
                            family = gaussian())
saveRDS(area_beta_15plus, "Area/Models/Total_Area/area_beta_15plus.rds"); rm(area_beta_15plus); gc()

area_alpha_15plus <- glmmTMB(area_full_formula("log_glu_area"), 
                             data = alpha_15plus_data, 
                             family = gaussian())
saveRDS(area_alpha_15plus, "Area/Models/Total_Area/area_alpha_15plus.rds"); rm(area_alpha_15plus); gc()

area_delta_15plus <- glmmTMB(area_full_formula("log_soma_area"), 
                             data = delta_15plus_data, 
                             family = gaussian())
saveRDS(area_delta_15plus, "Area/Models/Total_Area/area_delta_15plus.rds"); rm(area_delta_15plus); gc()

area_pp_15plus <- glmmTMB(area_full_formula("log_pp_area"), 
                          data = pp_15plus_data, 
                          family = gaussian())
saveRDS(area_pp_15plus, "Area/Models/Total_Area/area_pp_15plus.rds"); rm(area_pp_15plus); gc()
```

## Diagnosis — Combined (Total Area)

```{r area-diag-combined}
beta_combined_model <- glmmTMB(log_ins_area ~ (Diagnosis + Region + Sex + Age_c)^2 + 
                                 Diagnosis:Region:Age_c + log_Islet.Cells_c + 
                                 object_type * Diagnosis + object_type * Region +
                                 (1|Donor) + (1|Donor:ImageID),
                               data = beta_combined, 
                               family = gaussian())
saveRDS(beta_combined_model, "Area/Models/Total_Area/beta_combined_model.rds"); rm(beta_combined_model); gc()

alpha_combined_model <- glmmTMB(log_glu_area ~ (Diagnosis + Region + Sex + Age_c)^2 + 
                                  Diagnosis:Region:Age_c + log_Islet.Cells_c + 
                                  object_type * Diagnosis + object_type * Region +
                                  (1|Donor) + (1|Donor:ImageID),
                                data = alpha_combined, 
                                family = gaussian())
saveRDS(alpha_combined_model, "Area/Models/Total_Area/alpha_combined_model.rds"); rm(alpha_combined_model); gc()

delta_combined_model <- glmmTMB(log_soma_area ~ (Diagnosis + Region + Sex + Age_c)^2 + 
                                  Diagnosis:Region:Age_c + log_Islet.Cells_c + 
                                  object_type * Diagnosis + object_type * Region +
                                  (1|Donor) + (1|Donor:ImageID),
                                data = delta_combined, 
                                family = gaussian())
saveRDS(delta_combined_model, "Area/Models/Total_Area/delta_combined_model.rds"); rm(delta_combined_model); gc()

pp_combined_model <- glmmTMB(log_pp_area ~ (Diagnosis + Region + Sex + Age_c)^2 + 
                               Diagnosis:Region:Age_c + log_Islet.Cells_c + 
                               object_type * Diagnosis + object_type * Region +
                               (1|Donor) + (1|Donor:ImageID),
                             data = pp_combined, 
                             family = gaussian())
saveRDS(pp_combined_model, "Area/Models/Total_Area/pp_combined_model.rds"); rm(pp_combined_model); gc()
```

## T1D — Singlets (Total Area)

```{r area-t1d-singlets}
beta_1cell_data_t1d  <- split_area(area_singlets_t1d, "ins_area")
alpha_1cell_data_t1d <- split_area(area_singlets_t1d, "glu_area")
delta_1cell_data_t1d <- split_area(area_singlets_t1d, "soma_area")
pp_1cell_data_t1d    <- split_area(area_singlets_t1d, "pp_area")

area_beta_1cell_t1d <- glmmTMB(area_t1d_reduced_formula("log_ins_area"), 
                                 data = beta_1cell_data_t1d, 
                                 family = gaussian())
saveRDS(area_beta_1cell_t1d, "Area/Models/Total_Area/area_beta_1cell_t1d.rds"); rm(area_beta_1cell_t1d); gc()

area_alpha_1cell_t1d <- glmmTMB(area_t1d_reduced_formula("log_glu_area"), 
                                  data = alpha_1cell_data_t1d, 
                                  family = gaussian())
saveRDS(area_alpha_1cell_t1d, "Area/Models/Total_Area/area_alpha_1cell_t1d.rds"); rm(area_alpha_1cell_t1d); gc()

area_delta_1cell_t1d <- glmmTMB(area_t1d_reduced_formula("log_soma_area"), 
                                  data = delta_1cell_data_t1d, 
                                  family = gaussian())
saveRDS(area_delta_1cell_t1d, "Area/Models/Total_Area/area_delta_1cell_t1d.rds"); rm(area_delta_1cell_t1d); gc()

area_pp_1cell_t1d <- glmmTMB(area_t1d_reduced_formula("log_pp_area"), 
                               data = pp_1cell_data_t1d, 
                               family = gaussian())
saveRDS(area_pp_1cell_t1d, "Area/Models/Total_Area/area_pp_1cell_t1d.rds"); rm(area_pp_1cell_t1d); gc()
```

## T1D — Small EOs (Total Area)

```{r area-t1d-smallEO}
beta_2to14_data_t1d  <- split_area(area_endobs_t1d, "ins_area")
alpha_2to14_data_t1d <- split_area(area_endobs_t1d, "glu_area")
delta_2to14_data_t1d <- split_area(area_endobs_t1d, "soma_area")
pp_2to14_data_t1d    <- split_area(area_endobs_t1d, "pp_area")

area_beta_2to14_t1d <- glmmTMB(area_t1d_seo_formula("log_ins_area"), 
                                 data = beta_2to14_data_t1d, 
                                 family = gaussian())
saveRDS(area_beta_2to14_t1d, "Area/Models/Total_Area/area_beta_2to14_t1d.rds"); rm(area_beta_2to14_t1d); gc()

area_alpha_2to14_t1d <- glmmTMB(area_t1d_seo_formula("log_glu_area"), 
                                  data = alpha_2to14_data_t1d, 
                                  family = gaussian())
saveRDS(area_alpha_2to14_t1d, "Area/Models/Total_Area/area_alpha_2to14_t1d.rds"); rm(area_alpha_2to14_t1d); gc()

area_delta_2to14_t1d <- glmmTMB(area_t1d_seo_formula("log_soma_area"), 
                                  data = delta_2to14_data_t1d, 
                                  family = gaussian())
saveRDS(area_delta_2to14_t1d, "Area/Models/Total_Area/area_delta_2to14_t1d.rds"); rm(area_delta_2to14_t1d); gc()

area_pp_2to14_t1d <- glmmTMB(area_t1d_seo_formula("log_pp_area"), 
                               data = pp_2to14_data_t1d, 
                               family = gaussian())
saveRDS(area_pp_2to14_t1d, "Area/Models/Total_Area/area_pp_2to14_t1d.rds"); rm(area_pp_2to14_t1d); gc()
```

## T1D — Islets (Total Area)

```{r area-t1d-islets}
beta_15plus_data_t1d  <- split_area(area_islets_t1d, "ins_area")
alpha_15plus_data_t1d <- split_area(area_islets_t1d, "glu_area")
delta_15plus_data_t1d <- split_area(area_islets_t1d, "soma_area")
pp_15plus_data_t1d    <- split_area(area_islets_t1d, "pp_area")

area_beta_15plus_t1d <- glmmTMB(area_t1d_full_formula("log_ins_area"), 
                                  data = beta_15plus_data_t1d, 
                                  family = gaussian())
saveRDS(area_beta_15plus_t1d, "Area/Models/Total_Area/area_beta_15plus_t1d.rds"); rm(area_beta_15plus_t1d); gc()

area_alpha_15plus_t1d <- glmmTMB(area_t1d_full_formula("log_glu_area"), 
                                   data = alpha_15plus_data_t1d, 
                                   family = gaussian())
saveRDS(area_alpha_15plus_t1d, "Area/Models/Total_Area/area_alpha_15plus_t1d.rds"); rm(area_alpha_15plus_t1d); gc()

area_delta_15plus_t1d <- glmmTMB(area_t1d_full_formula("log_soma_area"), 
                                   data = delta_15plus_data_t1d, 
                                   family = gaussian())
saveRDS(area_delta_15plus_t1d, "Area/Models/Total_Area/area_delta_15plus_t1d.rds"); rm(area_delta_15plus_t1d); gc()

area_pp_15plus_t1d <- glmmTMB(area_t1d_full_formula("log_pp_area"), 
                                data = pp_15plus_data_t1d, 
                                family = gaussian())
saveRDS(area_pp_15plus_t1d, "Area/Models/Total_Area/area_pp_15plus_t1d.rds"); rm(area_pp_15plus_t1d); gc()
```

## T1D — Combined (Total Area)

```{r area-t1d-combined}
beta_combined_model_t1d <- glmmTMB(log_ins_area ~ (Disease.Duration_c + Region + Sex + age_at_onset_c)^2 +
                                 Disease.Duration_c:Region:age_at_onset_c + log_Islet.Cells_c + 
                                 object_type * Disease.Duration_c + object_type * Region +
                                 (1|Donor) + (1|Donor:ImageID),
                               data = beta_combined_t1d, 
                               family = gaussian())
saveRDS(beta_combined_model_t1d, "Area/Models/Total_Area/beta_combined_t1d.rds"); rm(beta_combined_model_t1d); gc()

alpha_combined_model_t1d <- glmmTMB(log_glu_area ~ (Disease.Duration_c + Region + Sex + age_at_onset_c)^2 + 
                                  Disease.Duration_c:Region:age_at_onset_c + log_Islet.Cells_c + 
                                  object_type * Disease.Duration_c + object_type * Region + 
                                  (1|Donor) + (1|Donor:ImageID),
                                data = alpha_combined_t1d, 
                                family = gaussian())
saveRDS(alpha_combined_model_t1d, "Area/Models/Total_Area/alpha_combined_t1d.rds"); rm(alpha_combined_model_t1d); gc()

delta_combined_model_t1d <- glmmTMB(log_soma_area ~ (Disease.Duration_c + Region + Sex + age_at_onset_c)^2 + 
                                  Disease.Duration_c:Region:age_at_onset_c + log_Islet.Cells_c + 
                                  object_type * Disease.Duration_c + object_type * Region + 
                                  (1|Donor) + (1|Donor:ImageID), 
                                data = delta_combined_t1d, 
                                family = gaussian())
saveRDS(delta_combined_model_t1d, "Area/Models/Total_Area/delta_combined_t1d.rds"); rm(delta_combined_model_t1d); gc()

pp_combined_model_t1d <- glmmTMB(log_pp_area ~ (Disease.Duration_c + Region + Sex + age_at_onset_c)^2 + 
                               Disease.Duration_c:Region:age_at_onset_c + log_Islet.Cells_c + 
                               object_type * Disease.Duration_c + object_type * Region + 
                               (1|Donor) + (1|Donor:ImageID), 
                             data = pp_combined_t1d, 
                             family = gaussian())
saveRDS(pp_combined_model_t1d, "Area/Models/Total_Area/pp_combined_t1d.rds"); rm(pp_combined_model_t1d); gc()
```

# Per-Cell Area Models — All Gaussian on log(area/count)

Per-cell area = total cell-type area / cell-type count within each islet.
Isolates individual cell size from cell number. Family choice informed by
`per_cell_area_diagnostics.Rmd`: Gaussian on log(per-cell area) wins by
ΔAIC = 250K–5M over Gamma and lognormal in all 12 combinations. Insulin
multi-cell/islet subsets have conservative underdispersion (0.45–0.55).

**Singlet models are not built here.** For singlets (Islet.Cells == 1),
per-cell area ≡ total area (dividing by 1 changes nothing). Reference the
existing total area singlet models in `Area/Models/area_*_1cell*.rds` instead.

## Diagnosis — Small EOs (Per-Cell Area)

```{r pc-diag-smallEO}
beta_2to14_pc  <- split_per_cell(area_endobs, "ins_area")
alpha_2to14_pc <- split_per_cell(area_endobs, "glu_area")
delta_2to14_pc <- split_per_cell(area_endobs, "soma_area")
pp_2to14_pc    <- split_per_cell(area_endobs, "pp_area")

pc_beta_2to14 <- glmmTMB(pc_seo_formula("log_ins_area_pc"),
                         data = beta_2to14_pc, family = gaussian())
saveRDS(pc_beta_2to14, "Area/Models/Per_Cell_Area/pc_beta_2to14.rds"); rm(pc_beta_2to14); gc()

pc_alpha_2to14 <- glmmTMB(pc_seo_formula("log_glu_area_pc"),
                          data = alpha_2to14_pc, family = gaussian())
saveRDS(pc_alpha_2to14, "Area/Models/Per_Cell_Area/pc_alpha_2to14.rds"); rm(pc_alpha_2to14); gc()

pc_delta_2to14 <- glmmTMB(pc_seo_formula("log_soma_area_pc"),
                          data = delta_2to14_pc, family = gaussian())
saveRDS(pc_delta_2to14, "Area/Models/Per_Cell_Area/pc_delta_2to14.rds"); rm(pc_delta_2to14); gc()

pc_pp_2to14 <- glmmTMB(pc_seo_formula("log_pp_area_pc"),
                       data = pp_2to14_pc, family = gaussian())
saveRDS(pc_pp_2to14, "Area/Models/Per_Cell_Area/pc_pp_2to14.rds"); rm(pc_pp_2to14); gc()
```

## Diagnosis — Islets (Per-Cell Area)

```{r pc-diag-islets}
beta_15plus_pc  <- split_per_cell(area_islets, "ins_area")
alpha_15plus_pc <- split_per_cell(area_islets, "glu_area")
delta_15plus_pc <- split_per_cell(area_islets, "soma_area")
pp_15plus_pc    <- split_per_cell(area_islets, "pp_area")

pc_beta_15plus <- glmmTMB(pc_full_formula("log_ins_area_pc"),
                          data = beta_15plus_pc, family = gaussian())
saveRDS(pc_beta_15plus, "Area/Models/Per_Cell_Area/pc_beta_15plus.rds"); rm(pc_beta_15plus); gc()

pc_alpha_15plus <- glmmTMB(pc_full_formula("log_glu_area_pc"),
                           data = alpha_15plus_pc, family = gaussian())
saveRDS(pc_alpha_15plus, "Area/Models/Per_Cell_Area/pc_alpha_15plus.rds"); rm(pc_alpha_15plus); gc()

pc_delta_15plus <- glmmTMB(pc_full_formula("log_soma_area_pc"),
                           data = delta_15plus_pc, family = gaussian())
saveRDS(pc_delta_15plus, "Area/Models/Per_Cell_Area/pc_delta_15plus.rds"); rm(pc_delta_15plus); gc()

pc_pp_15plus <- glmmTMB(pc_full_formula("log_pp_area_pc"),
                        data = pp_15plus_pc, family = gaussian())
saveRDS(pc_pp_15plus, "Area/Models/Per_Cell_Area/pc_pp_15plus.rds"); rm(pc_pp_15plus); gc()
```

## Diagnosis — Combined (Per-Cell Area)

```{r pc-diag-combined}
beta_comb_pc <- split_per_cell(beta_combined, "ins_area")
pc_beta_combined <- glmmTMB(log_ins_area_pc ~ (Diagnosis + Region + Sex + Age_c)^2 +
                              Diagnosis:Region:Age_c + log_Islet.Cells_c +
                              object_type * Diagnosis + object_type * Region +
                              (1|Donor) + (1|Donor:ImageID),
                            data = beta_comb_pc, family = gaussian())
saveRDS(pc_beta_combined, "Area/Models/Per_Cell_Area/pc_beta_combined.rds"); rm(pc_beta_combined); gc()

alpha_comb_pc <- split_per_cell(alpha_combined, "glu_area")
pc_alpha_combined <- glmmTMB(log_glu_area_pc ~ (Diagnosis + Region + Sex + Age_c)^2 +
                               Diagnosis:Region:Age_c + log_Islet.Cells_c +
                               object_type * Diagnosis + object_type * Region +
                               (1|Donor) + (1|Donor:ImageID),
                             data = alpha_comb_pc, family = gaussian())
saveRDS(pc_alpha_combined, "Area/Models/Per_Cell_Area/pc_alpha_combined.rds"); rm(pc_alpha_combined); gc()

delta_comb_pc <- split_per_cell(delta_combined, "soma_area")
pc_delta_combined <- glmmTMB(log_soma_area_pc ~ (Diagnosis + Region + Sex + Age_c)^2 +
                               Diagnosis:Region:Age_c + log_Islet.Cells_c +
                               object_type * Diagnosis + object_type * Region +
                               (1|Donor) + (1|Donor:ImageID),
                             data = delta_comb_pc, family = gaussian())
saveRDS(pc_delta_combined, "Area/Models/Per_Cell_Area/pc_delta_combined.rds"); rm(pc_delta_combined); gc()

pp_comb_pc <- split_per_cell(pp_combined, "pp_area")
pc_pp_combined <- glmmTMB(log_pp_area_pc ~ (Diagnosis + Region + Sex + Age_c)^2 +
                            Diagnosis:Region:Age_c + log_Islet.Cells_c +
                            object_type * Diagnosis + object_type * Region +
                            (1|Donor) + (1|Donor:ImageID),
                          data = pp_comb_pc, family = gaussian())
saveRDS(pc_pp_combined, "Area/Models/Per_Cell_Area/pc_pp_combined.rds"); rm(pc_pp_combined); gc()
```

## T1D — Small EOs (Per-Cell Area)

```{r pc-t1d-smallEO}
beta_2to14_pc_t1d  <- split_per_cell(area_endobs_t1d, "ins_area")
alpha_2to14_pc_t1d <- split_per_cell(area_endobs_t1d, "glu_area")
delta_2to14_pc_t1d <- split_per_cell(area_endobs_t1d, "soma_area")
pp_2to14_pc_t1d    <- split_per_cell(area_endobs_t1d, "pp_area")

pc_beta_2to14_t1d <- glmmTMB(pc_t1d_seo_formula("log_ins_area_pc"),
                             data = beta_2to14_pc_t1d, family = gaussian())
saveRDS(pc_beta_2to14_t1d, "Area/Models/Per_Cell_Area/pc_beta_2to14_t1d.rds"); rm(pc_beta_2to14_t1d); gc()

pc_alpha_2to14_t1d <- glmmTMB(pc_t1d_seo_formula("log_glu_area_pc"),
                              data = alpha_2to14_pc_t1d, family = gaussian())
saveRDS(pc_alpha_2to14_t1d, "Area/Models/Per_Cell_Area/pc_alpha_2to14_t1d.rds"); rm(pc_alpha_2to14_t1d); gc()

pc_delta_2to14_t1d <- glmmTMB(pc_t1d_seo_formula("log_soma_area_pc"),
                              data = delta_2to14_pc_t1d, family = gaussian())
saveRDS(pc_delta_2to14_t1d, "Area/Models/Per_Cell_Area/pc_delta_2to14_t1d.rds"); rm(pc_delta_2to14_t1d); gc()

pc_pp_2to14_t1d <- glmmTMB(pc_t1d_seo_formula("log_pp_area_pc"),
                           data = pp_2to14_pc_t1d, family = gaussian())
saveRDS(pc_pp_2to14_t1d, "Area/Models/Per_Cell_Area/pc_pp_2to14_t1d.rds"); rm(pc_pp_2to14_t1d); gc()
```

## T1D — Islets (Per-Cell Area)

```{r pc-t1d-islets}
beta_15plus_pc_t1d  <- split_per_cell(area_islets_t1d, "ins_area")
alpha_15plus_pc_t1d <- split_per_cell(area_islets_t1d, "glu_area")
delta_15plus_pc_t1d <- split_per_cell(area_islets_t1d, "soma_area")
pp_15plus_pc_t1d    <- split_per_cell(area_islets_t1d, "pp_area")

pc_beta_15plus_t1d <- glmmTMB(pc_t1d_full_formula("log_ins_area_pc"),
                              data = beta_15plus_pc_t1d, family = gaussian())
saveRDS(pc_beta_15plus_t1d, "Area/Models/Per_Cell_Area/pc_beta_15plus_t1d.rds"); rm(pc_beta_15plus_t1d); gc()

pc_alpha_15plus_t1d <- glmmTMB(pc_t1d_full_formula("log_glu_area_pc"),
                               data = alpha_15plus_pc_t1d, family = gaussian())
saveRDS(pc_alpha_15plus_t1d, "Area/Models/Per_Cell_Area/pc_alpha_15plus_t1d.rds"); rm(pc_alpha_15plus_t1d); gc()

pc_delta_15plus_t1d <- glmmTMB(pc_t1d_full_formula("log_soma_area_pc"),
                               data = delta_15plus_pc_t1d, family = gaussian())
saveRDS(pc_delta_15plus_t1d, "Area/Models/Per_Cell_Area/pc_delta_15plus_t1d.rds"); rm(pc_delta_15plus_t1d); gc()

pc_pp_15plus_t1d <- glmmTMB(pc_t1d_full_formula("log_pp_area_pc"),
                            data = pp_15plus_pc_t1d, family = gaussian())
saveRDS(pc_pp_15plus_t1d, "Area/Models/Per_Cell_Area/pc_pp_15plus_t1d.rds"); rm(pc_pp_15plus_t1d); gc()
```

## T1D — Combined (Per-Cell Area)

```{r pc-t1d-combined}
beta_comb_pc_t1d <- split_per_cell(beta_combined_t1d, "ins_area")
pc_beta_combined_t1d <- glmmTMB(log_ins_area_pc ~ (Disease.Duration_c + Region + Sex + age_at_onset_c)^2 +
                                  Disease.Duration_c:Region:age_at_onset_c + log_Islet.Cells_c +
                                  object_type * Disease.Duration_c + object_type * Region +
                                  (1|Donor) + (1|Donor:ImageID),
                                data = beta_comb_pc_t1d, family = gaussian())
saveRDS(pc_beta_combined_t1d, "Area/Models/Per_Cell_Area/pc_beta_combined_t1d.rds"); rm(pc_beta_combined_t1d); gc()

alpha_comb_pc_t1d <- split_per_cell(alpha_combined_t1d, "glu_area")
pc_alpha_combined_t1d <- glmmTMB(log_glu_area_pc ~ (Disease.Duration_c + Region + Sex + age_at_onset_c)^2 +
                                   Disease.Duration_c:Region:age_at_onset_c + log_Islet.Cells_c +
                                   object_type * Disease.Duration_c + object_type * Region +
                                   (1|Donor) + (1|Donor:ImageID),
                                 data = alpha_comb_pc_t1d, family = gaussian())
saveRDS(pc_alpha_combined_t1d, "Area/Models/Per_Cell_Area/pc_alpha_combined_t1d.rds"); rm(pc_alpha_combined_t1d); gc()

delta_comb_pc_t1d <- split_per_cell(delta_combined_t1d, "soma_area")
pc_delta_combined_t1d <- glmmTMB(log_soma_area_pc ~ (Disease.Duration_c + Region + Sex + age_at_onset_c)^2 +
                                   Disease.Duration_c:Region:age_at_onset_c + log_Islet.Cells_c +
                                   object_type * Disease.Duration_c + object_type * Region +
                                   (1|Donor) + (1|Donor:ImageID),
                                 data = delta_comb_pc_t1d, family = gaussian())
saveRDS(pc_delta_combined_t1d, "Area/Models/Per_Cell_Area/pc_delta_combined_t1d.rds"); rm(pc_delta_combined_t1d); gc()

pp_comb_pc_t1d <- split_per_cell(pp_combined_t1d, "pp_area")
pc_pp_combined_t1d <- glmmTMB(log_pp_area_pc ~ (Disease.Duration_c + Region + Sex + age_at_onset_c)^2 +
                                Disease.Duration_c:Region:age_at_onset_c + log_Islet.Cells_c +
                                object_type * Disease.Duration_c + object_type * Region +
                                (1|Donor) + (1|Donor:ImageID),
                              data = pp_comb_pc_t1d, family = gaussian())
saveRDS(pc_pp_combined_t1d, "Area/Models/Per_Cell_Area/pc_pp_combined_t1d.rds"); rm(pc_pp_combined_t1d); gc()
```

```{r}
# ============================================================
# Option 2: Total area models with cell-count adjustment
# ============================================================
# Asks whether the cross-sectional Diagnosis effect survives
# once you condition on cell count. If the per-cell effect was
# driven by Halo's count-vs-area decoupling in disease tissue,
# the Diagnosis coefficient here should collapse toward zero.

# area_to_count is the same mapping you already use in split_per_cell:
area_to_count <- list(ins_area = "Count.ins", glu_area = "Count.glu",
                      soma_area = "Count.soma", pp_area = "Count.PP")

# ---- Helper: filter to cell-type-positive objects, add log_count_c ----
split_total_area <- function(data, cell_col) {
  count_col    <- area_to_count[[cell_col]]
  log_area_col <- paste0("log_", cell_col)

  data %>%
    filter(!is.na(!!sym(cell_col)) & !!sym(cell_col) > 0 &
           !!sym(count_col) > 0) %>%
    mutate(
      !!log_area_col := log(!!sym(cell_col)),
      log_count      = log(!!sym(count_col)),
      log_count_c    = log_count - mean(log_count)
    )
}

# ---- Formulas: identical to pc_* formulas, but outcome is log(area)
#      not log(area/count), and log_count_c is added as a covariate ----

ta_count_seo_formula <- function(log_out) {
  as.formula(paste0(log_out, " ~ (Diagnosis + Region + Sex + Age_c)^2 +
    Diagnosis:Region:Age_c + log_count_c + log_count_c +
    (1|Donor) + (1|Donor:ImageID)"))
}

ta_count_full_formula <- function(log_out) {
  as.formula(paste0(log_out, " ~ (Diagnosis + Region + Sex + Age_c + log_count_c)^2 +
    Diagnosis:Region:log_count_c + Diagnosis:Region:Age_c + log_count_c +
    (1|Donor) + (1|Donor:ImageID)"))
}

ta_count_t1d_seo_formula <- function(log_out) {
  as.formula(paste0(log_out, " ~ (Disease.Duration_c + Region + Sex + age_at_onset_c)^2 +
    Disease.Duration_c:Region:age_at_onset_c + log_count_c + log_count_c +
    (1|Donor) + (1|Donor:ImageID)"))
}

ta_count_t1d_full_formula <- function(log_out) {
  as.formula(paste0(log_out, " ~ (Disease.Duration_c + Region + Sex + age_at_onset_c + log_count_c)^2 +
    Disease.Duration_c:Region:log_count_c + Disease.Duration_c:Region:age_at_onset_c + log_count_c +
    (1|Donor) + (1|Donor:ImageID)"))
}

# ---- Diagnosis — Small EOs (2-14 cells) ----
beta_2to14_ta  <- split_total_area(area_endobs, "ins_area")
alpha_2to14_ta <- split_total_area(area_endobs, "glu_area")
delta_2to14_ta <- split_total_area(area_endobs, "soma_area")
pp_2to14_ta    <- split_total_area(area_endobs, "pp_area")

ta_beta_2to14  <- glmmTMB(ta_count_seo_formula("log_ins_area"),
                          data = beta_2to14_ta,  family = gaussian())
ta_alpha_2to14 <- glmmTMB(ta_count_seo_formula("log_glu_area"),
                          data = alpha_2to14_ta, family = gaussian())
ta_delta_2to14 <- glmmTMB(ta_count_seo_formula("log_soma_area"),
                          data = delta_2to14_ta, family = gaussian())
ta_pp_2to14    <- glmmTMB(ta_count_seo_formula("log_pp_area"),
                          data = pp_2to14_ta,    family = gaussian())

# ---- Diagnosis — Islets (15+ cells) ----
beta_15plus_ta  <- split_total_area(area_islets, "ins_area")
alpha_15plus_ta <- split_total_area(area_islets, "glu_area")
delta_15plus_ta <- split_total_area(area_islets, "soma_area")
pp_15plus_ta    <- split_total_area(area_islets, "pp_area")

ta_beta_15plus  <- glmmTMB(ta_count_full_formula("log_ins_area"),
                           data = beta_15plus_ta,  family = gaussian())
ta_alpha_15plus <- glmmTMB(ta_count_full_formula("log_glu_area"),
                           data = alpha_15plus_ta, family = gaussian())
ta_delta_15plus <- glmmTMB(ta_count_full_formula("log_soma_area"),
                           data = delta_15plus_ta, family = gaussian())
ta_pp_15plus    <- glmmTMB(ta_count_full_formula("log_pp_area"),
                           data = pp_15plus_ta,    family = gaussian())

# (T1D-only versions follow the same pattern with the _t1d data and
# ta_count_t1d_seo_formula / ta_count_t1d_full_formula. Add saveRDS calls
# wherever you want them in your existing pipeline.)

summary(ta_beta_2to14)
summary(ta_beta_15plus)

summary(ta_alpha_2to14)
summary(ta_alpha_15plus)

summary(ta_delta_2to14)
summary(ta_delta_15plus)

summary(ta_pp_2to14)
summary(ta_pp_15plus)
```


```{r final-summary}
cat("\n================================================================\n")
cat("  ALL MODELS COMPLETE\n")
cat("================================================================\n\n")
cat("Family choices (from diagnostics):\n")
cat("  Density:         Gamma(log)\n")
cat("  Count.ins:       NB1 (AIC winner by 80K; underdispersion = conservative SEs)\n")
cat("  Count.glu:       ZINB1, zi ~ Diagnosis (ΔAIC = -5170 vs NB1)\n")
cat("  Count.soma:      NB1 (disp = 0.88)\n")
cat("  Count.PP:        NB1 (disp = 1.05)\n")
cat("  Islet.Cells:     NB2 (AIC winner; disp = 1.36; ZINB worse by +4)\n")
cat("  All total area:  Gaussian on log(area) (LogNorm disp 0.84-1.01)\n")
cat("  All per-cell:    Gaussian on log(area/count) (ΔAIC 250K-5M vs Gamma/LN)\n")
cat("                   Insulin multi-cell/islet underdispersed (0.45-0.55)\n\n")

cat("Per-cell area note: Singlet models not built — for singlets,\n")
cat("per-cell area ≡ total area (count = 1). Reference Area/Models/area_*_1cell*.rds.\n\n")

for (d in c("Density/Models","Count/Models","Area/Models/Total_Area/", "Area/Models/Per_Cell_Area/")) {
  files <- list.files(d, pattern = "\\.rds$")
  cat(d, ":", length(files), "models\n")
  for (f in files) cat("  ", f, "\n")
}

cat(sprintf("\nTotal models in this file: %d\n",
  sum(sapply(c("Density/Models","Count/Models","Area/Models/Total_Area/", "Area/Models/Per_Cell_Area/"),
             function(d) length(list.files(d, pattern = "\\.rds$"))))))
cat("  + HTC cluster models (proportion & ratio ordbetareg)\n")
```

# Session info

```{r}
sessionInfo()
```
