---
title: "Unified Dataset Pipeline"
output: html_document
date: "2026-03-22"
---

This single Rmd produces **all** project datasets in one run:

| Output file | Description |
|---|---|
| `no_cutoff_data.csv` | All islets (Islet.Cells > 0), no size filter, with location |
| `full_data.csv` | Size-filtered (15 ≤ Islet.Cells, log ≤ 5.5), with location |
| `quad_data.csv` | Quad-stain subset of full_data, with location |
| `area_data_joined.csv` | Cell-level area measurements joined to count data |
| `density_data.csv` | Image-level islet density (islets / tissue area) |
| `ki67_data.csv` | Ki-67 proliferation data matched to count islets, with demographics |
| `ki67_data_t1d.csv` | Ki-67 data, T1D donors only |

**Data sources required in working directory or specified paths:**

- `corrected nPOD Data/` — folder of per-donor HALO CSVs (count data)
- `F:/Work/nPOD_final_v3/area_data/` — folder of per-donor HALO area CSVs
- `matched_tissue_classification_images.csv` — verified image list
- `No_stain_data_stain_added.xlsx` — stain fix lookup
- `remove_list.xlsx` — manual islet exclusion list
- `01282025_Full.nPOD.csv` — donor demographics

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 12, fig.height = 8)

packages <- c("tidyverse", "readr", "stringr", "readxl", "purrr", "writexl",
              "janitor")

for (pkg in packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
  library(pkg, character.only = TRUE)
}

dir.create("Data", showWarnings = FALSE)
```

---

# Part A: Core Count Pipeline (produces no_cutoff, full_data, quad_data)

## A1. Read and combine all count CSVs

```{r a1_read_count_csvs}
count_base_path <- "corrected nPOD Data"

csv_files <- list.files(count_base_path, pattern = "\\.csv$",
                        recursive = TRUE, full.names = TRUE)
csv_files <- csv_files[!grepl("Total_Summary_Results\\.csv$", csv_files)]

cat("Found", length(csv_files), "count CSV files\n")

Data_raw <- csv_files %>%
  lapply(function(file) {
    tryCatch({
      df <- read.csv(file, stringsAsFactors = FALSE)
      df$source_file <- basename(file)
      df$folder_id <- basename(dirname(dirname(dirname(dirname(file)))))
      df$anatomical_location <- basename(dirname(dirname(dirname(file))))
      return(df)
    }, error = function(e) {
      cat("Error reading file:", file, "\n", e$message, "\n")
      return(NULL)
    })
  }) %>%
  bind_rows()

# Fix HPAP folder IDs
Data_raw <- Data_raw %>%
  mutate(folder_id = case_when(
    folder_id == "030" ~ "HPAP030",
    folder_id == "041" ~ "HPAP041",
    folder_id == "048" ~ "HPAP048",
    folder_id == "068" ~ "HPAP068",
    folder_id == "078" ~ "HPAP078",
    TRUE ~ folder_id
  ))

length(unique(Data_raw$folder_id))
```

```{r}
Data_raw$pct_sum <- Data_raw$X..T1D.AI.Phenotyper...Beta.Cell.Positive.Cells +
  Data_raw$X..T1D.AI.Phenotyper...Alpha.Cell.Positive.Cells +
  Data_raw$X..T1D.AI.Phenotyper...Delta.Cell.Positive.Cells +
  Data_raw$X..T1D.AI.Phenotyper...PP.Cell.Positive.Cells +
  Data_raw$X..T1D.AI.Phenotyper...Acinar.Cell.Positive.Cells

summary(Data_raw$pct_sum)
```


## A2. Save location columns before column-dropping

```{r a2_location_lookup}
location_lookup <- Data_raw %>%
  mutate(ImageID = str_extract(source_file, "\\d+")) %>%
  dplyr::select(
    source_file,
    Object.Id,
    ImageID,
    Region.Area..μm..,
    Avg.Cell.Area..um.2.,
    Region.Perimeter..μm.
  )

cat("Location lookup:", nrow(location_lookup), "rows\n")
```

## A3. Drop junk columns

```{r a3_drop_columns}
phenotyper_raw <- Data_raw %>%
  mutate(ImageID = str_extract(source_file, "\\d+"))

Data <- Data_raw %>%
  dplyr::select(-matches(
    "hema|area|negative|nucleus|weak|moderate|strong|score|membrane|cytoplasm|nuclei|positive",
    ignore.case = TRUE
  )) %>%
  dplyr::select(-starts_with("X"))

rm(Data_raw)

length(unique(Data$folder_id))
```

## A4. Select and summarize cell counts

```{r a4_cell_counts}
Data2 <- Data %>%
  mutate(ImageID = str_extract(source_file, "\\d+")) %>%
  dplyr::select(
    source_file,
    Object.Id,
    Donor = folder_id,
    Region = anatomical_location,
    Total.Cells,
    Algorithm.Name,
    matches("Phenotyper", ignore.case = TRUE),
    ImageID
  )

rm(Data)

Data_summed <- Data2 %>%
  mutate(across(matches("Phenotyper", ignore.case = TRUE), as.numeric)) %>%
  mutate(
    Count.ins    = rowSums(dplyr::select(., matches("Beta(\\.Cell\\.Cells|\\.Cells)$",  ignore.case = TRUE)), na.rm = TRUE),
    Count.glu    = rowSums(dplyr::select(., matches("Alpha(\\.Cell\\.Cells|\\.Cells)$", ignore.case = TRUE)), na.rm = TRUE),
    Count.soma   = rowSums(dplyr::select(., matches("Delta(\\.Cell\\.Cells|\\.Cells)$", ignore.case = TRUE)), na.rm = TRUE),
    Count.PP     = rowSums(dplyr::select(., matches("PP(\\.Cell\\.Cells|\\.Cells)$",    ignore.case = TRUE)), na.rm = TRUE),
    Count.Acinar = rowSums(dplyr::select(., matches("Acinar(\\.Cell\\.Cells|\\.Cells)$", ignore.case = TRUE)), na.rm = TRUE)
  ) %>%
  mutate(Stain = case_when(
    str_detect(Algorithm.Name, regex("Single", ignore_case = TRUE)) ~ "Single",
    str_detect(Algorithm.Name, regex("Double", ignore_case = TRUE)) ~ "Double",
    str_detect(Algorithm.Name, regex("Quad",   ignore_case = TRUE)) ~ "Quad",
    TRUE ~ NA_character_
  )) %>%
  dplyr::select(
    Donor, ImageID, Object.Id, Region, Stain, Algorithm.Name,
    source_file, Total.Cells, matches("Count", ignore.case = TRUE)
  ) %>%
  mutate(
    Islet.Cells        = Count.ins + Count.glu + Count.soma + Count.PP,
    non_primary_cells   = Total.Cells - Islet.Cells - Count.Acinar,
    Percent.ins    = Count.ins  / Islet.Cells,
    Percent.glu    = Count.glu  / Islet.Cells,
    Percent.soma   = Count.soma / Islet.Cells,
    Percent.PP     = Count.PP   / Islet.Cells,
    Percent.other  = non_primary_cells / (Islet.Cells + non_primary_cells)
  )

length(unique(Data_summed$Donor))

rm(Data2)
```

## A5. Fix stain NAs

```{r a5_fix_stain}
Stain_info <- read_xlsx("No_stain_data_stain_added.xlsx")

Data_summed <- Data_summed %>%
  left_join(Stain_info %>% dplyr::select(ImageID, Stain_new = Stain), by = "ImageID") %>%
  mutate(Stain = if_else(is.na(Stain), Stain_new, Stain)) %>%
  dplyr::select(-Stain_new)

rm(Stain_info)

length(unique(Data_summed$Donor))
```

## A6. Apply remove_list

```{r a6_remove_list}
remove_list <- read_excel("remove_list.xlsx")

remove_list <- remove_list %>%
  fill(ImageID) %>%
  mutate(Object.Id = as.numeric(Object.Id),
         ImageID   = as.character(ImageID))

full_data_clean <- Data_summed %>%
  anti_join(remove_list, by = c("ImageID", "Object.Id"))

cat("Rows before remove_list:", nrow(Data_summed), "\n")
cat("Rows after  remove_list:", nrow(full_data_clean), "\n")

length(unique(full_data_clean$Donor))

rm(remove_list, Data_summed)
```

## A7. Join demographics

```{r a7_demographics}
original_data <- as.data.frame(read_csv("01282025_Full.nPOD.csv"))

donor_info <- original_data %>%
  dplyr::select(Donor, Gender, Age, AgeCat, Diagnosis, Disease.Duration) %>%
  distinct(Donor, .keep_all = TRUE) %>%
  mutate(Donor = as.character(Donor))

full_data_clean <- full_data_clean %>%
  mutate(Donor = as.character(Donor)) %>%
  left_join(donor_info, by = "Donor")

length(unique(full_data_clean$Donor))

rm(donor_info)
```

## A8. Fix demographics, filter donors, recode

```{r a8_fix_demographics}
full_data_clean <- full_data_clean %>%
  # Fix donor 6250
  mutate(
    Diagnosis = if_else(Donor == "6250", "No.Diagnosis", Diagnosis),
    AgeCat    = if_else(Donor == "6250", "20-40_years",  AgeCat),
    Gender    = if_else(Donor == "6250", "Male",         Gender),
    Age       = if_else(Donor == "6250", 40,             Age)
  ) %>%
  rename(Sex = Gender) %>%
  # Remove excluded donors
  filter(!Donor %in% c("6021", "6102", "6127")) %>% # Removed because they are older than 40
  # Recode Diagnosis to ND/T1D
  mutate(Diagnosis = if_else(Diagnosis == "T1D", "T1D", "ND")) %>%
  # Fix AgeCat typos
  mutate(AgeCat = dplyr::recode(AgeCat,
    "20-40years"   = "20-40_years",
    "20-40_yeasrs" = "20-40_years",
    "20-40"        = "20-40_years"
  ))

# Disease duration: ND = 0, T1D with 0 → 0.1
full_data_clean$Disease.Duration[full_data_clean$Diagnosis == "ND"] <- 0
full_data_clean$Disease.Duration[
  full_data_clean$Diagnosis == "T1D" & full_data_clean$Disease.Duration == 0
] <- 0.1

length(unique(full_data_clean$Donor))
length(unique(full_data_clean$ImageID))
nrow(full_data_clean)

full_data_clean %>%
  group_by(Stain) %>%
  summarise(n_donors = n_distinct(Donor))

donors_with_na <- unique(full_data_clean$Donor[is.na(full_data_clean$Stain)])

full_data_clean %>%
  filter(Donor %in% donors_with_na) %>%
  group_by(Donor) %>%
  summarise(
    n_NA      = sum(is.na(Stain)),
    n_nonNA   = sum(!is.na(Stain)),
    stains    = paste(sort(unique(Stain[!is.na(Stain)])), collapse = ", ")
  )

full_data_clean %>%
  group_by(Donor) %>%
  filter(any(is.na(Stain)) & any(!is.na(Stain))) %>%
  arrange(Donor)
```

## A9. Apply cutoffs, join location, and save datasets

```{r a9_cutoffs_and_save}
# --- no_cutoff_data: all islets with Islet.Cells > 0 ---
no_cutoff_data <- full_data_clean %>%
  filter(Islet.Cells > 0) %>%
  mutate(log_Islet.Cells = log(as.numeric(Islet.Cells))) %>%
  mutate(
    Diagnosis       = factor(Diagnosis, levels = c("ND", "T1D")),
    Region          = factor(Region, levels = c("Head", "Body", "Tail")),
    Sex             = factor(Sex, levels = c("Female", "Male")),
    Donor           = factor(Donor),
    ImageID         = factor(ImageID)
  ) %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    log_Islet.Cells_c  = scale(log_Islet.Cells, center = TRUE, scale = FALSE)[, 1],
    Disease.Duration_c  = scale(Disease.Duration, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    age_at_onset   = Age - Disease.Duration,
    age_at_onset_c = scale(age_at_onset, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  # Join location variables
  left_join(
    location_lookup %>% dplyr::select(source_file, Object.Id,
                                       Region.Area..μm..,
                                       Avg.Cell.Area..um.2.,
                                       Region.Perimeter..μm.),
    by = c("source_file", "Object.Id")
  )

write_csv(no_cutoff_data, "Data/no_cutoff_data.csv")
saveRDS(no_cutoff_data, "Data/no_cutoff_data.rds")
cat("Saved: no_cutoff_data.csv —", nrow(no_cutoff_data), "rows,",
    length(unique(no_cutoff_data$Donor)), "donors\n")

no_cutoff_data_t1d <- no_cutoff_data %>%
  filter(Diagnosis == "T1D") %>% 
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    log_Islet.Cells_c  = scale(log_Islet.Cells, center = TRUE, scale = FALSE)[, 1],
    Disease.Duration_c  = scale(Disease.Duration, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    age_at_onset   = Age - Disease.Duration,
    age_at_onset_c = scale(age_at_onset, center = TRUE, scale = FALSE)[, 1]
  )

write_csv(no_cutoff_data_t1d, "Data/no_cutoff_data_t1d.csv")
saveRDS(no_cutoff_data_t1d, "Data/no_cutoff_data_t1d.rds")
cat("Saved: no_cutoff_data_t1d.csv —", nrow(no_cutoff_data_t1d), "rows,",
    length(unique(no_cutoff_data_t1d$Donor)), "donors\n")

# --- full_data: size-filtered ---
full_data <- no_cutoff_data %>%
  filter(log_Islet.Cells <= 5.5) %>%
  filter(Islet.Cells >= 15) %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    log_Islet.Cells_c  = scale(log_Islet.Cells, center = TRUE, scale = FALSE)[, 1],
    Disease.Duration_c  = scale(Disease.Duration, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    age_at_onset   = Age - Disease.Duration,
    age_at_onset_c = scale(age_at_onset, center = TRUE, scale = FALSE)[, 1]
  ) 

write_csv(full_data, "Data/cutoff_data.csv")
saveRDS(full_data, "Data/cutoff_data.rds")
cat("Saved: cutoff_data.csv —", nrow(full_data), "rows,",
    length(unique(full_data$Donor)), "donors\n")

full_data_t1d <- full_data %>%
  filter(Diagnosis == "T1D") %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    log_Islet.Cells_c  = scale(log_Islet.Cells, center = TRUE, scale = FALSE)[, 1],
    Disease.Duration_c  = scale(Disease.Duration, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    age_at_onset   = Age - Disease.Duration,
    age_at_onset_c = scale(age_at_onset, center = TRUE, scale = FALSE)[, 1]
  ) 

write_csv(full_data_t1d, "Data/cutoff_data_t1d.csv")
saveRDS(full_data_t1d, "Data/cutoff_data_t1d.rds")
cat("Saved: cutoff_data_t1d.csv —", nrow(full_data_t1d), "rows,",
    length(unique(full_data_t1d$Donor)), "donors\n")

ratio_data <- full_data %>%
  filter(Stain != "Single") %>%
  filter(Donor != "6473") %>%
  filter(Count.ins > 0 | Count.glu > 0) %>%
  mutate(beta_alpha_ratio = Count.ins / (Count.ins + Count.glu)) %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    log_Islet.Cells_c  = scale(log_Islet.Cells, center = TRUE, scale = FALSE)[, 1],
    Disease.Duration_c  = scale(Disease.Duration, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    age_at_onset   = Age - Disease.Duration,
    age_at_onset_c = scale(age_at_onset, center = TRUE, scale = FALSE)[, 1]
  ) 

write_csv(ratio_data, "Data/ratio_data.csv")
saveRDS(ratio_data, "Data/ratio_data.rds")
cat("Saved: ratio_data.csv —", nrow(ratio_data), "rows,",
    length(unique(ratio_data$Donor)), "donors\n")

ratio_data_t1d <- ratio_data %>%
  filter(Diagnosis == "T1D") %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    log_Islet.Cells_c  = scale(log_Islet.Cells, center = TRUE, scale = FALSE)[, 1],
    Disease.Duration_c  = scale(Disease.Duration, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    age_at_onset   = Age - Disease.Duration,
    age_at_onset_c = scale(age_at_onset, center = TRUE, scale = FALSE)[, 1]
  ) 

write_csv(ratio_data_t1d, "Data/ratio_data_t1d.csv")
saveRDS(ratio_data_t1d, "Data/ratio_data_t1d.rds")
cat("Saved: ratio_data_t1d.csv —", nrow(ratio_data_t1d), "rows,",
    length(unique(ratio_data_t1d$Donor)), "donors\n")

# --- quad_data: Quad stain only ---
quad_data <- full_data %>%
  filter(Stain == "Quad") %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    log_Islet.Cells_c  = scale(log_Islet.Cells, center = TRUE, scale = FALSE)[, 1],
    Disease.Duration_c  = scale(Disease.Duration, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    age_at_onset   = Age - Disease.Duration,
    age_at_onset_c = scale(age_at_onset, center = TRUE, scale = FALSE)[, 1]
  ) 

write_csv(quad_data, "Data/quad_data.csv")
saveRDS(quad_data, "Data/quad_data.rds")
cat("Saved: quad_data.csv —", nrow(quad_data), "rows,",
    length(unique(quad_data$Donor)), "donors\n")

quad_data_t1d <- quad_data %>%
  filter(Diagnosis == "T1D") %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    log_Islet.Cells_c  = scale(log_Islet.Cells, center = TRUE, scale = FALSE)[, 1],
    Disease.Duration_c  = scale(Disease.Duration, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    age_at_onset   = Age - Disease.Duration,
    age_at_onset_c = scale(age_at_onset, center = TRUE, scale = FALSE)[, 1]
  ) 

write_csv(quad_data_t1d, "Data/quad_data_t1d.csv")
saveRDS(quad_data_t1d, "Data/quad_data_t1d.rds")
cat("Saved: quad_data_t1d.csv —", nrow(quad_data_t1d), "rows,",
    length(unique(quad_data_t1d$Donor)), "donors\n")


```
---

# Part B: Area Data (cell-level area measurements)

## B1. Read and harmonize area CSVs

```{r b1_read_area_data}
area_base_path <- "F:/Work/nPOD_final_v4/area_data"

area_files <- Sys.glob(file.path(area_base_path, "*/Halo archive*/ObjectData/*.csv"))
area_files <- area_files[!grepl("Total_Summary_Results", area_files, ignore.case = TRUE)]

cat("Found", length(area_files), "area data files\n")

read_and_harmonize <- function(filepath) {
  df <- read_csv(filepath, show_col_types = FALSE)

  colnames(df) <- gsub("^\uFEFF", "", colnames(df))
  colnames(df) <- make.names(colnames(df))

  is_quad   <- grepl("Quad",   filepath, ignore.case = TRUE)
  is_double <- grepl("Double", filepath, ignore.case = TRUE)

  df$ImageID <- as.integer(str_extract(df$Image.Location, "[0-9]+(?=\\.svs)"))
  df$Donor   <- str_extract(filepath, "[^/\\\\]+(?=[/\\\\]Halo archive)")

  if ("Object.ID" %in% names(df) && !"Object.Id" %in% names(df)) {
    df$Object.Id <- df$Object.ID
  }

  cols <- names(df)
  ins_col  <- cols[grepl("^Ins.*Area.*um",  cols, ignore.case = TRUE) & !grepl("Weak|Moderate|Strong", cols, ignore.case = TRUE)]
  glu_col  <- cols[grepl("^Glu.*Area.*um",  cols, ignore.case = TRUE) & !grepl("Weak|Moderate|Strong", cols, ignore.case = TRUE)]
  soma_col <- cols[grepl("^Soma.*Area.*um", cols, ignore.case = TRUE) & !grepl("Weak|Moderate|Strong", cols, ignore.case = TRUE)]
  pp_col   <- cols[grepl("^PP.*Area.*um",   cols, ignore.case = TRUE) & !grepl("Weak|Moderate|Strong", cols, ignore.case = TRUE)]

  result <- data.frame(
    Donor      = df$Donor,
    ImageID    = df$ImageID,
    Object.Id  = df$Object.Id,
    stain_type = ifelse(is_quad, "Quad", ifelse(is_double, "Double", "Unknown")),
    stringsAsFactors = FALSE
  )

  result$ins_area  <- if (length(ins_col)  == 1) df[[ins_col]]  else NA_real_
  result$glu_area  <- if (length(glu_col)  == 1) df[[glu_col]]  else NA_real_
  result$soma_area <- if (length(soma_col) == 1) df[[soma_col]] else NA_real_
  result$pp_area   <- if (length(pp_col)   == 1) df[[pp_col]]   else NA_real_

  if (is_double) {
    result$soma_area <- NA_real_
    result$pp_area   <- NA_real_
  }

  result$source_file <- basename(filepath)
  return(result)
}

all_area_data <- map_dfr(area_files, function(f) {
  tryCatch(read_and_harmonize(f),
           error = function(e) {
             warning(paste("Error reading:", basename(f), "-", e$message))
             return(NULL)
           })
})

cat("Area data:", nrow(all_area_data), "rows from",
    length(unique(all_area_data$source_file)), "files\n")
```


## B2. Fix donor IDs and filter

```{r b2_fix_area_donors}
all_area_data <- all_area_data %>%
  mutate(Donor = case_when(
    Donor == "6343" ~ "6434",
    TRUE ~ Donor
  )) %>%
  filter(!Donor %in% c("6021", "6102", "6127"))
```

## B3. Filter to verified images

```{r b3_verified_images}
# Read area summary files
area_summary_files <- Sys.glob(file.path(area_base_path, "*/Halo archive*/Total_Summary_Results.csv"))

summary_area <- map_dfr(area_summary_files, function(f) {
  tryCatch({
    df <- read_csv(f, show_col_types = FALSE)
    df$source_file <- basename(dirname(f))
    return(df)
  }, error = function(e) {
    warning(paste("Error reading:", f))
    return(NULL)
  })
})

# Read count summary files
count_summary_files <- list.files(count_base_path, pattern = "Total_Summary_Results\\.csv$",
                                   recursive = TRUE, full.names = TRUE)

summary_count <- map_dfr(count_summary_files, function(f) {
  tryCatch({
    df <- read_csv(f, show_col_types = FALSE)
    df$source_file <- basename(dirname(f))
    return(df)
  }, error = function(e) {
    warning(paste("Error reading:", f))
    return(NULL)
  })
})

cat("Area summaries:", nrow(summary_area), "rows\n")
cat("Count summaries:", nrow(summary_count), "rows\n")

# Join on Image Tag and keep only images where tissue classification matches exactly
summary_compare <- summary_area %>%
  inner_join(summary_count, by = "Image Tag", suffix = c(".area", ".count"))

matched_tags <- summary_compare %>%
  filter(
    `Classified Area (mm²).area` == `Classified Area (mm²).count`,
    `Acinar Area (mm²).area`     == `Acinar Area (mm²).count`,
    `Glass Area (mm²).area`      == `Glass Area (mm²).count`
  ) %>%
  pull(`Image Tag`)

matched_image_ids <- as.integer(str_extract(matched_tags, "\\d+"))

cat("Images with matching tissue classification:", length(matched_image_ids),
    "out of", nrow(summary_compare), "compared\n")

# Filter area data to verified images only
n_before <- nrow(all_area_data)
all_area_data <- all_area_data %>%
  filter(ImageID %in% matched_image_ids)

cat("Filtered to verified images:", nrow(all_area_data), "rows\n")
cat("Removed", n_before - nrow(all_area_data), "rows from unverified images\n")

# Clean up
rm(summary_area, summary_count, summary_compare, matched_tags, matched_image_ids); gc()
```

## B4. Join area data to count data

```{r b4_join_area_to_counts}
# Use no_cutoff_data as the count backbone (already has demographics, location, etc.)
count_for_area <- no_cutoff_data %>%
  mutate(
    Donor    = as.character(Donor),
    ImageID  = as.integer(as.character(ImageID)),
    Object.Id = as.integer(Object.Id)
  ) %>%
  dplyr::select(
    Donor, Diagnosis, Disease.Duration, Disease.Duration_c, Sex, Age, 
    age_at_onset, log_Islet.Cells, ImageID, Object.Id, Region,
    Count.ins, Count.glu, Count.soma, Count.PP, Islet.Cells,
    Percent.ins, Percent.glu, Percent.soma, Percent.PP
  )

all_area_data <- all_area_data %>%
  mutate(
    Donor     = as.character(Donor),
    ImageID   = as.integer(ImageID),
    Object.Id = as.integer(Object.Id)
  )

area_data_joined <- count_for_area %>%
  inner_join(all_area_data, by = c("Donor", "ImageID", "Object.Id")) %>%
  mutate(stain_rank = case_when(
    stain_type == "Quad" ~ 1,
    stain_type == "Double" ~ 2,
    TRUE ~ 3
  )) %>%
  arrange(stain_rank) %>%
  distinct(ImageID, Object.Id, .keep_all = TRUE) %>%
  select(-stain_rank) %>%
  mutate(
    ins_area  = if_else(Count.ins  > 0 & ins_area  == 0, NA_real_, ins_area),
    glu_area  = if_else(Count.glu  > 0 & glu_area  == 0, NA_real_, glu_area),
    soma_area = if_else(Count.soma > 0 & soma_area == 0, NA_real_, soma_area),
    pp_area   = if_else(Count.PP   > 0 & pp_area   == 0, NA_real_, pp_area)
  ) %>%
  mutate(
    ins_area_per_cell  = ifelse(Count.ins  > 0, ins_area  / Count.ins,  NA_real_),
    glu_area_per_cell  = ifelse(Count.glu  > 0, glu_area  / Count.glu,  NA_real_),
    soma_area_per_cell = ifelse(Count.soma > 0, soma_area / Count.soma, NA_real_),
    pp_area_per_cell   = ifelse(Count.PP   > 0, pp_area   / Count.PP,   NA_real_)
  ) %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    log_Islet.Cells_c  = scale(log_Islet.Cells, center = TRUE, scale = FALSE)[, 1],
    Disease.Duration_c  = scale(Disease.Duration, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    age_at_onset   = Age - Disease.Duration,
    age_at_onset_c = scale(age_at_onset, center = TRUE, scale = FALSE)[, 1]
  ) 

write_csv(area_data_joined, "Data/area_data.csv")
saveRDS(area_data_joined, "Data/area_data.rds")
cat("Saved: area_data_joined.csv —", nrow(area_data_joined), "rows,",
    length(unique(area_data_joined$Donor)), "donors,",
    length(unique(area_data_joined$ImageID)), "images\n")

area_data_joined_t1d <- area_data_joined %>%
  filter(Diagnosis == "T1D") %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    log_Islet.Cells_c  = scale(log_Islet.Cells, center = TRUE, scale = FALSE)[, 1],
    Disease.Duration_c  = scale(Disease.Duration, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    age_at_onset   = Age - Disease.Duration,
    age_at_onset_c = scale(age_at_onset, center = TRUE, scale = FALSE)[, 1]
  ) 

write_csv(area_data_joined_t1d, "Data/area_data_t1d.csv")
saveRDS(area_data_joined_t1d, "Data/area_data_t1d.rds")
cat("Saved: area_data_joined_t1d.csv —", nrow(area_data_joined_t1d), "rows,",
    length(unique(area_data_joined_t1d$Donor)), "donors,",
    length(unique(area_data_joined_t1d$ImageID)), "images\n")
```

## B5 Area subsets

### Singlets

```{r}
area_data_joined_singlets <- area_data_joined %>%
  filter(Islet.Cells == 1) %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    Disease.Duration_c  = scale(Disease.Duration, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    age_at_onset   = Age - Disease.Duration,
    age_at_onset_c = scale(age_at_onset, center = TRUE, scale = FALSE)[, 1]
  ) 

write_csv(area_data_joined_singlets, "Data/area_data_singlets.csv")
saveRDS(area_data_joined_singlets, "Data/area_data_singlets.rds")
cat("Saved: area_data_joined_singlets.csv —", nrow(area_data_joined_singlets), "rows,",
    length(unique(area_data_joined_singlets$Donor)), "donors,",
    length(unique(area_data_joined_singlets$ImageID)), "images\n")

area_data_joined_singlets_t1d <- area_data_joined_singlets %>%
  filter(Diagnosis == "T1D") %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    Disease.Duration_c  = scale(Disease.Duration, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    age_at_onset   = Age - Disease.Duration,
    age_at_onset_c = scale(age_at_onset, center = TRUE, scale = FALSE)[, 1]
  ) 

write_csv(area_data_joined_singlets_t1d, "Data/area_data_singlets_t1d.csv")
saveRDS(area_data_joined_singlets_t1d, "Data/area_data_singlets_t1d.rds")
cat("Saved: area_data_joined_singlets_t1d.csv —", nrow(area_data_joined_singlets_t1d), "rows,",
    length(unique(area_data_joined_singlets_t1d$Donor)), "donors,",
    length(unique(area_data_joined_singlets_t1d$ImageID)), "images\n")
```

### Endocrine Objects

```{r}
area_data_joined_endobs <- area_data_joined %>%
  filter(Islet.Cells >= 2 & Islet.Cells <= 14) %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    log_Islet.Cells_c  = scale(log_Islet.Cells, center = TRUE, scale = FALSE)[, 1],
    Disease.Duration_c  = scale(Disease.Duration, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    age_at_onset   = Age - Disease.Duration,
    age_at_onset_c = scale(age_at_onset, center = TRUE, scale = FALSE)[, 1]
  ) 

write_csv(area_data_joined_endobs, "Data/area_data_endobs.csv")
saveRDS(area_data_joined_endobs, "Data/area_data_endobs.rds")
cat("Saved: area_data_joined_endobs.csv —", nrow(area_data_joined_endobs), "rows,",
    length(unique(area_data_joined_endobs$Donor)), "donors,",
    length(unique(area_data_joined_endobs$ImageID)), "images\n")

area_data_joined_endobs_t1d <- area_data_joined_endobs %>%
  filter(Diagnosis == "T1D") %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    log_Islet.Cells_c  = scale(log_Islet.Cells, center = TRUE, scale = FALSE)[, 1],
    Disease.Duration_c  = scale(Disease.Duration, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    age_at_onset   = Age - Disease.Duration,
    age_at_onset_c = scale(age_at_onset, center = TRUE, scale = FALSE)[, 1]
  ) 

write_csv(area_data_joined_endobs_t1d, "Data/area_data_endobs_t1d.csv")
saveRDS(area_data_joined_endobs_t1d, "Data/area_data_endobs_t1d.rds")
cat("Saved: area_data_joined_endobs_t1d.csv —", nrow(area_data_joined_endobs_t1d), "rows,",
    length(unique(area_data_joined_endobs_t1d$Donor)), "donors,",
    length(unique(area_data_joined_endobs_t1d$ImageID)), "images\n")
```

### Islets

```{r}
area_data_joined_islets <- area_data_joined %>%
  filter(Islet.Cells >= 15 & log_Islet.Cells <= 5.5) %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    log_Islet.Cells_c  = scale(log_Islet.Cells, center = TRUE, scale = FALSE)[, 1],
    Disease.Duration_c  = scale(Disease.Duration, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    age_at_onset   = Age - Disease.Duration,
    age_at_onset_c = scale(age_at_onset, center = TRUE, scale = FALSE)[, 1]
  ) 

write_csv(area_data_joined_islets, "Data/area_data_islets.csv")
saveRDS(area_data_joined_islets, "Data/area_data_islets.rds")
cat("Saved: area_data_joined_islets.csv —", nrow(area_data_joined_islets), "rows,",
    length(unique(area_data_joined_islets$Donor)), "donors,",
    length(unique(area_data_joined_islets$ImageID)), "images\n")

area_data_joined_islets_t1d <- area_data_joined_islets %>%
  filter(Diagnosis == "T1D") %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    log_Islet.Cells_c  = scale(log_Islet.Cells, center = TRUE, scale = FALSE)[, 1],
    Disease.Duration_c  = scale(Disease.Duration, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    age_at_onset   = Age - Disease.Duration,
    age_at_onset_c = scale(age_at_onset, center = TRUE, scale = FALSE)[, 1]
  ) 

write_csv(area_data_joined_islets_t1d, "Data/area_data_islets_t1d.csv")
saveRDS(area_data_joined_islets_t1d, "Data/area_data_islets_t1d.rds")
cat("Saved: area_data_joined_islets_t1d.csv —", nrow(area_data_joined_islets_t1d), "rows,",
    length(unique(area_data_joined_islets_t1d$Donor)), "donors,",
    length(unique(area_data_joined_islets_t1d$ImageID)), "images\n")
```

### Combined

```{r}
beta_combined <- area_data_joined %>%
  filter(!is.na(ins_area) & ins_area > 0 & Count.ins > 0) %>%
  mutate(log_ins_area = log(ins_area)) %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    Disease.Duration_c = scale(Disease.Duration, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    age_at_onset   = Age - Disease.Duration,
    age_at_onset_c = scale(age_at_onset, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    log_Islet.Cells_c = if_else(
      Islet.Cells == 1,
      0,
      log_Islet.Cells - mean(log_Islet.Cells[Islet.Cells > 1])
    )
  )

write_csv(beta_combined, "Data/area_data_beta_combined.csv")
saveRDS(beta_combined, "Data/area_data_beta_combined.rds")
cat("Saved: area_data_beta_combined.csv —", nrow(beta_combined), "rows,",
    length(unique(beta_combined$Donor)), "donors,",
    length(unique(beta_combined$ImageID)), "images\n")

beta_combined_t1d <- beta_combined %>%
  filter(Diagnosis == "T1D") %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    Disease.Duration_c = scale(Disease.Duration, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    age_at_onset   = Age - Disease.Duration,
    age_at_onset_c = scale(age_at_onset, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    log_Islet.Cells_c = if_else(
      Islet.Cells == 1,
      0,
      log_Islet.Cells - mean(log_Islet.Cells[Islet.Cells > 1])
    )
  )

write_csv(beta_combined_t1d, "Data/area_data_beta_combined_t1d.csv")
saveRDS(beta_combined_t1d, "Data/area_data_beta_combined_t1d.rds")
cat("Saved: area_data_beta_combined_t1d.csv —", nrow(beta_combined_t1d), "rows,",
    length(unique(beta_combined_t1d$Donor)), "donors,",
    length(unique(beta_combined_t1d$ImageID)), "images\n")




alpha_combined <- area_data_joined %>%
  filter(!is.na(glu_area) & glu_area > 0 & Count.glu > 0) %>%
  mutate(log_glu_area = log(glu_area)) %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    Disease.Duration_c = scale(Disease.Duration, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    age_at_onset   = Age - Disease.Duration,
    age_at_onset_c = scale(age_at_onset, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    log_Islet.Cells_c = if_else(
      Islet.Cells == 1,
      0,
      log_Islet.Cells - mean(log_Islet.Cells[Islet.Cells > 1])
    )
  )

write_csv(alpha_combined, "Data/area_data_alpha_combined.csv")
saveRDS(alpha_combined, "Data/area_data_alpha_combined.rds")
cat("Saved: area_data_alpha_combined.csv —", nrow(alpha_combined), "rows,",
    length(unique(alpha_combined$Donor)), "donors,",
    length(unique(alpha_combined$ImageID)), "images\n")

alpha_combined_t1d <- alpha_combined %>%
  filter(Diagnosis == "T1D") %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    Disease.Duration_c = scale(Disease.Duration, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    age_at_onset   = Age - Disease.Duration,
    age_at_onset_c = scale(age_at_onset, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    log_Islet.Cells_c = if_else(
      Islet.Cells == 1,
      0,
      log_Islet.Cells - mean(log_Islet.Cells[Islet.Cells > 1])
    )
  )

write_csv(alpha_combined_t1d, "Data/area_data_alpha_combined_t1d.csv")
saveRDS(alpha_combined_t1d, "Data/area_data_alpha_combined_t1d.rds")
cat("Saved: area_data_alpha_combined_t1d.csv —", nrow(alpha_combined_t1d), "rows,",
    length(unique(alpha_combined_t1d$Donor)), "donors,",
    length(unique(alpha_combined_t1d$ImageID)), "images\n")




delta_combined <- area_data_joined %>%
  filter(!is.na(soma_area) & soma_area > 0 & Count.soma > 0) %>%
  mutate(log_soma_area = log(soma_area)) %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    Disease.Duration_c = scale(Disease.Duration, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    age_at_onset   = Age - Disease.Duration,
    age_at_onset_c = scale(age_at_onset, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    log_Islet.Cells_c = if_else(
      Islet.Cells == 1,
      0,
      log_Islet.Cells - mean(log_Islet.Cells[Islet.Cells > 1])
    )
  )

write_csv(delta_combined, "Data/area_data_delta_combined.csv")
saveRDS(delta_combined, "Data/area_data_delta_combined.rds")
cat("Saved: area_data_delta_combined.csv —", nrow(delta_combined), "rows,",
    length(unique(delta_combined$Donor)), "donors,",
    length(unique(delta_combined$ImageID)), "images\n")

delta_combined_t1d <- delta_combined %>%
  filter(Diagnosis == "T1D") %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    Disease.Duration_c = scale(Disease.Duration, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    age_at_onset   = Age - Disease.Duration,
    age_at_onset_c = scale(age_at_onset, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    log_Islet.Cells_c = if_else(
      Islet.Cells == 1,
      0,
      log_Islet.Cells - mean(log_Islet.Cells[Islet.Cells > 1])
    )
  )

write_csv(delta_combined_t1d, "Data/area_data_delta_combined_t1d.csv")
saveRDS(delta_combined_t1d, "Data/area_data_delta_combined_t1d.rds")
cat("Saved: area_data_delta_combined_t1d.csv —", nrow(delta_combined_t1d), "rows,",
    length(unique(delta_combined_t1d$Donor)), "donors,",
    length(unique(delta_combined_t1d$ImageID)), "images\n")




pp_combined <- area_data_joined %>%
  filter(!is.na(pp_area) & pp_area > 0 & Count.PP > 0) %>%
  mutate(log_pp_area = log(pp_area)) %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    Disease.Duration_c = scale(Disease.Duration, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    age_at_onset   = Age - Disease.Duration,
    age_at_onset_c = scale(age_at_onset, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    log_Islet.Cells_c = if_else(
      Islet.Cells == 1,
      0,
      log_Islet.Cells - mean(log_Islet.Cells[Islet.Cells > 1])
    )
  )

write_csv(pp_combined, "Data/area_data_pp_combined.csv")
saveRDS(pp_combined, "Data/area_data_pp_combined.rds")
cat("Saved: area_data_pp_combined.csv —", nrow(pp_combined), "rows,",
    length(unique(pp_combined$Donor)), "donors,",
    length(unique(pp_combined$ImageID)), "images\n")

pp_combined_t1d <- pp_combined %>%
  filter(Diagnosis == "T1D") %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    Disease.Duration_c = scale(Disease.Duration, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    age_at_onset   = Age - Disease.Duration,
    age_at_onset_c = scale(age_at_onset, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    log_Islet.Cells_c = if_else(
      Islet.Cells == 1,
      0,
      log_Islet.Cells - mean(log_Islet.Cells[Islet.Cells > 1])
    )
  )

write_csv(pp_combined_t1d, "Data/area_data_pp_combined_t1d.csv")
saveRDS(pp_combined_t1d, "Data/area_data_pp_combined_t1d.rds")
cat("Saved: area_data_pp_combined_t1d.csv —", nrow(pp_combined_t1d), "rows,",
    length(unique(pp_combined_t1d$Donor)), "donors,",
    length(unique(pp_combined_t1d$ImageID)), "images\n")
```
---

# Part C: Density Data (image-level islet density)

## C1. Read Total_Summary_Results files

```{r c1_read_summary_files}
count_summary_files <- list.files(count_base_path,
                                  pattern = "Total_Summary_Results\\.csv$",
                                  recursive = TRUE, full.names = TRUE)

summary_count <- map_dfr(count_summary_files, function(f) {
  tryCatch({
    df <- read_csv(f, show_col_types = FALSE)
    df$source_file <- basename(dirname(f))
    df$full_path <- f
    return(df)
  }, error = function(e) {
    warning(paste("Error reading:", f))
    return(NULL)
  })
})

cat("Summary files loaded:", nrow(summary_count), "rows\n")
```

## C2. Coalesce column name variants and compute tissue area

```{r c2_compute_tissue_area}
density <- summary_count %>%
  mutate(
    `Duct Area (mm²)` = coalesce(
      `Ducts Area (mm²)`, `Duct Area (mm²)`,
      `Ducts/Arteries Area (mm²)`, `duct Area (mm²)`
    ),
    `Glass Area (mm²)` = coalesce(`glass Area (mm²)`, `Glass Area (mm²)`),
    `Tissue Area (μm²)` = coalesce(`Tissue Area (μm²)`, `Tissue Area Analyzed (μm²)`)
  ) %>%
  mutate(
    `Duct Area (mm²)` = replace_na(`Duct Area (mm²)`, 0),
    full_tissue_area   = `Classified Area (mm²)` - `Duct Area (mm²)` - `Glass Area (mm²)`,
    ImageID            = as.character(str_remove(`Image Tag`, "\\.svs$"))
  )
```

## C3. Count islets per image and compute density

```{r c3_islet_density}
# Use full_data (size-filtered) to count islets per image
islet_counts <- full_data %>%
  mutate(ImageID = as.character(ImageID)) %>%
  group_by(ImageID) %>%
  summarise(
    islet_count = n(),
    Donor       = first(as.character(Donor)),
    .groups     = "drop"
  )

density_result <- density %>%
  left_join(islet_counts, by = "ImageID") %>%
  mutate(islet_density = islet_count / full_tissue_area) %>%
  dplyr::select(ImageID, full_tissue_area, islet_density, islet_count, Donor) %>%
  filter(!is.na(islet_density)) 

cat("Density computed for", nrow(density_result), "images,",
    length(unique(density_result$Donor)), "donors\n")
```

## C4. Join demographics to density and save

```{r c4_save_density}
density_result$ImageID <- as.numeric(density_result$ImageID)

density_dataset <- density_result %>%
  left_join(
    full_data %>%
      mutate(ImageID = as.numeric(as.character(ImageID)),
             Donor   = as.character(Donor)) %>%
      dplyr::select(ImageID, Donor, Region, Age, age_at_onset, Sex, Diagnosis, Disease.Duration) %>%
      distinct(),
    by = c("ImageID", "Donor")
  ) %>%
  # Drop images whose islets were all size-filtered out (so the join produced NA demographics)
  { 
    dropped <- sum(is.na(.$Diagnosis))
    if (dropped > 0) cat("Dropping", dropped, "images with no islets surviving the size filter\n")
    filter(., !is.na(Diagnosis))
  } %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    Disease.Duration_c = scale(Disease.Duration, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    age_at_onset   = Age - Disease.Duration,
    age_at_onset_c = scale(age_at_onset, center = TRUE, scale = FALSE)[, 1]
  )

write_csv(density_dataset, "Data/density_data.csv")
saveRDS(density_dataset, "Data/density_data.rds")
cat("Saved: density_data.csv —", nrow(density_dataset), "rows,",
    length(unique(density_dataset$Donor)), "donors\n")

density_dataset_t1d <- density_dataset %>%
  filter(Diagnosis == "T1D") %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    Disease.Duration_c  = scale(Disease.Duration, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    age_at_onset   = Age - Disease.Duration,
    age_at_onset_c = scale(age_at_onset, center = TRUE, scale = FALSE)[, 1]
  ) 

write_csv(density_dataset_t1d, "Data/density_data_t1d.csv")
saveRDS(density_dataset_t1d, "Data/density_data_t1d.rds")
cat("Saved: density_data_t1d.csv —", nrow(density_dataset_t1d), "rows,",
    length(unique(density_dataset_t1d$Donor)), "donors\n")

unique(density_dataset$Donor)
```
---
# Part D: Ki-67 Proliferation Data

## D1. Read and combine all Ki-67 CSVs

```{r d1_read_ki67_csvs}
ki67_base_path <- "ki67_data"

ki67_files <- list.files(ki67_base_path, pattern = "\\.csv$",
                         recursive = TRUE, full.names = TRUE)
ki67_files <- ki67_files[!grepl("Total_Summary_Results\\.csv$", ki67_files)]

cat("Found", length(ki67_files), "ki67 CSV files\n")

ki67_raw <- ki67_files %>%
  lapply(function(file) {
    tryCatch({
      df <- read.csv(file, stringsAsFactors = FALSE)
      df$source_file <- basename(file)
      df$folder_id <- basename(dirname(dirname(dirname(dirname(file)))))
      df$anatomical_location <- basename(dirname(dirname(dirname(file))))
      return(df)
    }, error = function(e) {
      cat("Error reading file:", file, "\n", e$message, "\n")
      return(NULL)
    })
  }) %>%
  bind_rows() %>%
  mutate(
    folder_id = case_when(
      folder_id == "030" ~ "HPAP030",
      folder_id == "041" ~ "HPAP041",
      folder_id == "048" ~ "HPAP048",
      folder_id == "068" ~ "HPAP068",
      folder_id == "078" ~ "HPAP078",
      TRUE ~ folder_id
    ),
    ImageID = str_extract(source_file, "\\d+")
  )

cat("Ki-67 raw rows:", nrow(ki67_raw), "\n")
```

## D2. Dedupe phenotyper data to physical islets

```{r d2_dedupe_physical_islets}
# Each physical islet appears once per algorithm run (Single/Double/Quad) in
# the phenotyper data. For spatial matching we need one canonical row per
# physical islet, keyed by bounding-box coordinates.
phenotyper_islets <- phenotyper_raw %>%
  distinct(ImageID, XMin, XMax, YMin, YMax, .keep_all = TRUE)

cat("Phenotyper rows (all algorithms):", nrow(phenotyper_raw), "\n")
cat("Phenotyper distinct physical islets:", nrow(phenotyper_islets), "\n")
```

## D3. Spatially match ki67 islets to phenotyper islets

```{r d3_spatial_match}
# Greedy nearest-centroid matcher: for each image, pairs islets across the two
# datasets by bounding-box centroid distance, enforcing 1-to-1 pairings.
match_islets_one_image <- function(df1, df2, max_dist = 20) {
  df1 <- df1 %>% mutate(cx = (XMin + XMax)/2, cy = (YMin + YMax)/2, .i = row_number())
  df2 <- df2 %>% mutate(cx = (XMin + XMax)/2, cy = (YMin + YMax)/2, .j = row_number())

  pairs <- expand.grid(.i = df1$.i, .j = df2$.j) %>%
    mutate(dist = sqrt((df1$cx[.i] - df2$cx[.j])^2 +
                       (df1$cy[.i] - df2$cy[.j])^2)) %>%
    filter(dist <= max_dist) %>%
    arrange(dist)

  used1 <- logical(nrow(df1))
  used2 <- logical(nrow(df2))
  keep  <- logical(nrow(pairs))
  for (k in seq_len(nrow(pairs))) {
    if (!used1[pairs$.i[k]] && !used2[pairs$.j[k]]) {
      keep[k] <- TRUE
      used1[pairs$.i[k]] <- TRUE
      used2[pairs$.j[k]] <- TRUE
    }
  }
  pairs <- pairs[keep, ]

  tibble(
    ki67_Object.Id  = df1$Object.Id[pairs$.i],
    count_Object.Id = df2$Object.Id[pairs$.j],
    dist_px         = pairs$dist
  )
}

# Apply per image
all_matches <- phenotyper_islets %>%
  distinct(ImageID) %>%
  pull(ImageID) %>%
  map_dfr(function(img) {
    d1 <- ki67_raw          %>% filter(ImageID == img)
    d2 <- phenotyper_islets %>% filter(ImageID == img)
    if (nrow(d1) == 0 || nrow(d2) == 0) return(NULL)
    match_islets_one_image(d1, d2, max_dist = 20) %>%
      mutate(ImageID = img, .before = 1)
  })

cat("Raw matches:", nrow(all_matches), "\n")
```

## D4. Filter matches by distance and bounding-box size ratio

```{r d4_filter_matches}
# True matches should be both spatially close AND similar in size.
# Attach bounding-box areas from each dataset, compute size ratio, filter.
all_matches_clean <- all_matches %>%
  left_join(
    ki67_raw %>%
      mutate(area1 = (XMax - XMin) * (YMax - YMin)) %>%
      dplyr::select(ImageID, Object.Id, area1,
                    ki67_XMin = XMin, ki67_XMax = XMax,
                    ki67_YMin = YMin, ki67_YMax = YMax),
    by = c("ImageID", "ki67_Object.Id" = "Object.Id")
  ) %>%
  left_join(
    phenotyper_islets %>%
      mutate(area2 = (XMax - XMin) * (YMax - YMin)) %>%
      dplyr::select(ImageID, Object.Id, area2,
                    ph_XMin = XMin, ph_XMax = XMax,
                    ph_YMin = YMin, ph_YMax = YMax),
    by = c("ImageID", "count_Object.Id" = "Object.Id")
  ) %>%
  mutate(area_ratio = pmin(area1, area2) / pmax(area1, area2)) %>%
  filter(dist_px <= 15, area_ratio >= 0.5)

cat("Matched islets after filtering:", nrow(all_matches_clean),
    sprintf("(%.1f%% of ki67 islets)\n",
            100 * nrow(all_matches_clean) / nrow(ki67_raw)))
```

## D5. Build ki67_data and join demographics

```{r d5_build_ki67_data}
# Pull all Ki-67 columns from raw ki67 data (regex handles the "Ki.67" dot).
ki67_cols <- ki67_raw %>%
  dplyr::select(
    ImageID, XMin, XMax, YMin, YMax,
    matches("Ki\\.?_?67", ignore.case = TRUE)
  )

# Join ki67 values onto the FULL phenotyper_raw (not the deduped version)
# using phenotyper coordinates, so every algorithm row for a matched islet
# receives the same ki67 values.
match_link <- all_matches_clean %>%
  dplyr::select(
    ImageID,
    XMin = ph_XMin, XMax = ph_XMax, YMin = ph_YMin, YMax = ph_YMax,
    ki67_XMin, ki67_XMax, ki67_YMin, ki67_YMax,
    dist_px
  )

ki67_joined <- phenotyper_raw %>%
  inner_join(match_link,
             by = c("ImageID", "XMin", "XMax", "YMin", "YMax"),
             relationship = "many-to-one") %>%
  left_join(ki67_cols,
            by = c("ImageID",
                   "ki67_XMin" = "XMin", "ki67_XMax" = "XMax",
                   "ki67_YMin" = "YMin", "ki67_YMax" = "YMax"),
            relationship = "many-to-one")

# Select desired columns and apply basic filters
ki67_data <- ki67_joined %>%
  clean_names(replace = c("\u03bc" = "u")) %>%
  dplyr::select(
    folder_id,
    ImageID = image_id,
    object_id,
    total_cells,
    ki_67_phenotyper_insulin_cells,
    x_ki_67_phenotyper_insulin_positive_cells,
    ki_67_phenotyper_glucagon_cells,
    x_ki_67_phenotyper_glucagon_positive_cells,
    ki_67_phenotyper_acinar_cells,
    x_ki_67_phenotyper_acinar_positive_cells,
    tissue_area_um_2,
    dist_px
  ) %>%
  filter(total_cells > 0)

# Join demographics from no_cutoff_data (already has all donor-level fixes applied)
ki67_data <- ki67_data %>%
  mutate(folder_id = as.character(folder_id),
         ImageID   = as.character(ImageID),
         object_id = as.integer(object_id)) %>%
  left_join(
    no_cutoff_data %>%
      mutate(Donor     = as.character(Donor),
             ImageID   = as.character(ImageID),
             Object.Id = as.integer(Object.Id),
             Region    = as.character(Region),
             log_islet_cells = as.numeric(log_Islet.Cells)) %>%
      dplyr::select(Donor, ImageID, Object.Id, Region,
                    Sex, Age, AgeCat, Diagnosis,
                    Disease.Duration, age_at_onset,
                    Count.ins, Count.glu, log_Islet.Cells) %>%
      distinct(),
    by = c("folder_id" = "Donor", "ImageID", "object_id" = "Object.Id")
  ) %>%
  filter(!is.na(Diagnosis)) %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    log_Islet.Cells_c  = scale(log_Islet.Cells, center = TRUE, scale = FALSE)[, 1]
  )

# Derive proliferation rates using the phenotyper's hormone-cell counts as denominators
ki67_data <- ki67_data %>%
  mutate(
    # Reconstruct absolute counts of Ki-67+ hormone cells from the percent columns
    ki67pos_glu = round(x_ki_67_phenotyper_glucagon_positive_cells *
                        ki_67_phenotyper_glucagon_cells / 100),
    ki67pos_ins = round(x_ki_67_phenotyper_insulin_positive_cells *
                        ki_67_phenotyper_insulin_cells / 100),
    
    # Proliferation rate using the more reliable phenotyper Count.glu / Count.ins
    # as the hormone-cell denominator
    glu_prolif_rate = if_else(Count.glu > 0, ki67pos_glu / Count.glu, NA_real_),
    ins_prolif_rate = if_else(Count.ins > 0, ki67pos_ins / Count.ins, NA_real_),
    
    # Flag biologically-impossible values (cross-algorithm disagreement on small islets)
    glu_prolif_impossible = !is.na(glu_prolif_rate) & glu_prolif_rate > 1,
    ins_prolif_impossible = !is.na(ins_prolif_rate) & ins_prolif_rate > 1,
    
    # NA out the impossible values so they don't contaminate downstream stats
    glu_prolif_rate = if_else(glu_prolif_impossible, NA_real_, glu_prolif_rate),
    ins_prolif_rate = if_else(ins_prolif_impossible, NA_real_, ins_prolif_rate)
  ) %>%
  rename(Donor = folder_id)

rm(ki67_joined, ki67_cols, match_link, all_matches, all_matches_clean); gc()
```

## D6. Sanity checks

```{r d6_ki67_sanity}
cat("=== Ki-67 dataset sanity checks ===\n\n")

cat("Distance distribution (should be tight near 0, max near 15):\n")
print(summary(ki67_data$dist_px))

cat("\nMissing demographics (should be 0):\n")
print(ki67_data %>%
  summarise(
    missing_sex       = sum(is.na(Sex)),
    missing_age       = sum(is.na(Age)),
    missing_diagnosis = sum(is.na(Diagnosis))
  ))

cat("\nAlgorithm disagreements (ki67+ insulin cells > total cells):\n")
cat("  count:", sum(ki67_data$ki67_gt_total, na.rm = TRUE),
    sprintf("(%.3f%% of rows)\n",
            100 * mean(ki67_data$ki67_gt_total, na.rm = TRUE)))
```

## D7. Save Ki-67 datasets

```{r d7_save_ki67}
write_csv(ki67_data, "Data/ki67_data.csv")
saveRDS(ki67_data, "Data/ki67_data.rds")
cat("Saved: ki67_data.csv —", nrow(ki67_data), "rows,",
    length(unique(ki67_data$Donor)), "donors,",
    length(unique(ki67_data$ImageID)), "images\n")

ki67_data_t1d <- ki67_data %>% filter(Diagnosis == "T1D") %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    log_Islet.Cells_c  = scale(log_Islet.Cells, center = TRUE, scale = FALSE)[, 1],
    Disease.Duration_c  = scale(Disease.Duration, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    age_at_onset   = Age - Disease.Duration,
    age_at_onset_c = scale(age_at_onset, center = TRUE, scale = FALSE)[, 1]
  )

write_csv(ki67_data_t1d, "Data/ki67_data_t1d.csv")
saveRDS(ki67_data_t1d, "Data/ki67_data_t1d.rds")
cat("Saved: ki67_data_t1d.csv —", nrow(ki67_data_t1d), "rows,",
    length(unique(ki67_data_t1d$Donor)), "donors\n")
```
---

# Part E: CD3 Data

## E1. Read and combine all CD3 CSVs

```{r d1_read_ki67_csvs}
cd3_base_path <- "cd3_data"

cd3_files <- list.files(cd3_base_path, pattern = "\\.csv$",
                         recursive = TRUE, full.names = TRUE)
cd3_files <- cd3_files[!grepl("Total_Summary_Results\\.csv$", cd3_files)]

cat("Found", length(cd3_files), "CD3 CSV files\n")

cd3_raw <- cd3_files %>%
  lapply(function(file) {
    tryCatch({
      df <- read.csv(file, stringsAsFactors = FALSE)
      df$source_file <- basename(file)
      df$folder_id <- basename(dirname(dirname(dirname(dirname(file)))))
      df$anatomical_location <- basename(dirname(dirname(dirname(file))))
      return(df)
    }, error = function(e) {
      cat("Error reading file:", file, "\n", e$message, "\n")
      return(NULL)
    })
  }) %>%
  bind_rows() %>%
  mutate(
    folder_id = case_when(
      folder_id == "030" ~ "HPAP030",
      folder_id == "041" ~ "HPAP041",
      folder_id == "048" ~ "HPAP048",
      folder_id == "078" ~ "HPAP078",
      TRUE ~ folder_id
    ),
    ImageID = str_extract(source_file, "\\d+")
  )

cat("CD3 raw rows:", nrow(cd3_raw), "\n")
```

## E2. Dedupe phenotyper data to physical islets

```{r d2_dedupe_physical_islets}
# Each physical islet appears once per algorithm run (Single/Double/Quad) in
# the phenotyper data. For spatial matching we need one canonical row per
# physical islet, keyed by bounding-box coordinates.
phenotyper_islets <- phenotyper_raw %>%
  distinct(ImageID, XMin, XMax, YMin, YMax, .keep_all = TRUE)

cat("Phenotyper rows (all algorithms):", nrow(phenotyper_raw), "\n")
cat("Phenotyper distinct physical islets:", nrow(phenotyper_islets), "\n")
```

## E3. Spatially match CD3 islets to phenotyper islets

```{r d3_spatial_match}
# Greedy nearest-centroid matcher: for each image, pairs islets across the two
# datasets by bounding-box centroid distance, enforcing 1-to-1 pairings.
match_islets_one_image <- function(df1, df2, max_dist = 20) {
  df1 <- df1 %>% mutate(cx = (XMin + XMax)/2, cy = (YMin + YMax)/2, .i = row_number())
  df2 <- df2 %>% mutate(cx = (XMin + XMax)/2, cy = (YMin + YMax)/2, .j = row_number())

  pairs <- expand.grid(.i = df1$.i, .j = df2$.j) %>%
    mutate(dist = sqrt((df1$cx[.i] - df2$cx[.j])^2 +
                       (df1$cy[.i] - df2$cy[.j])^2)) %>%
    filter(dist <= max_dist) %>%
    arrange(dist)

  used1 <- logical(nrow(df1))
  used2 <- logical(nrow(df2))
  keep  <- logical(nrow(pairs))
  for (k in seq_len(nrow(pairs))) {
    if (!used1[pairs$.i[k]] && !used2[pairs$.j[k]]) {
      keep[k] <- TRUE
      used1[pairs$.i[k]] <- TRUE
      used2[pairs$.j[k]] <- TRUE
    }
  }
  pairs <- pairs[keep, ]

  tibble(
    cd3_Object.Id  = df1$Object.Id[pairs$.i],
    count_Object.Id = df2$Object.Id[pairs$.j],
    dist_px         = pairs$dist
  )
}

# Apply per image
all_matches <- phenotyper_islets %>%
  distinct(ImageID) %>%
  pull(ImageID) %>%
  map_dfr(function(img) {
    d1 <- cd3_raw          %>% filter(ImageID == img)
    d2 <- phenotyper_islets %>% filter(ImageID == img)
    if (nrow(d1) == 0 || nrow(d2) == 0) return(NULL)
    match_islets_one_image(d1, d2, max_dist = 20) %>%
      mutate(ImageID = img, .before = 1)
  })

cat("Raw matches:", nrow(all_matches), "\n")
```

## E4. Filter matches by distance and bounding-box size ratio

```{r d4_filter_matches}
# True matches should be both spatially close AND similar in size.
# Attach bounding-box areas from each dataset, compute size ratio, filter.
all_matches_clean <- all_matches %>%
  left_join(
    cd3_raw %>%
      mutate(area1 = (XMax - XMin) * (YMax - YMin)) %>%
      dplyr::select(ImageID, Object.Id, area1,
                    cd3_XMin = XMin, cd3_XMax = XMax,
                    cd3_YMin = YMin, cd3_YMax = YMax),
    by = c("ImageID", "cd3_Object.Id" = "Object.Id")
  ) %>%
  left_join(
    phenotyper_islets %>%
      mutate(area2 = (XMax - XMin) * (YMax - YMin)) %>%
      dplyr::select(ImageID, Object.Id, area2,
                    ph_XMin = XMin, ph_XMax = XMax,
                    ph_YMin = YMin, ph_YMax = YMax),
    by = c("ImageID", "count_Object.Id" = "Object.Id")
  ) %>%
  mutate(area_ratio = pmin(area1, area2) / pmax(area1, area2)) %>%
  filter(dist_px <= 15, area_ratio >= 0.5)

cat("Matched islets after filtering:", nrow(all_matches_clean),
    sprintf("(%.1f%% of CD3 islets)\n",
            100 * nrow(all_matches_clean) / nrow(cd3_raw)))
```

## E5. Build cd3_data and join demographics

```{r d5_build_cd3_data}
# Pull all Ki-67 columns from raw CD3 data (regex handles the "Ki.67" dot).
cd3_cols <- cd3_raw %>%
  dplyr::select(
    ImageID, XMin, XMax, YMin, YMax,
    matches("Ki\\.?_?67", ignore.case = TRUE)
  )

# Join CD3 values onto the FULL phenotyper_raw (not the deduped version)
# using phenotyper coordinates, so every algorithm row for a matched islet
# receives the same CD3 values.
match_link <- all_matches_clean %>%
  dplyr::select(
    ImageID,
    XMin = ph_XMin, XMax = ph_XMax, YMin = ph_YMin, YMax = ph_YMax,
    cd3_XMin, cd3_XMax, cd3_YMin, cd3_YMax,
    dist_px
  )

cd3_joined <- phenotyper_raw %>%
  inner_join(match_link,
             by = c("ImageID", "XMin", "XMax", "YMin", "YMax"),
             relationship = "many-to-one") %>%
  left_join(cd3_cols,
            by = c("ImageID",
                   "cd3_XMin" = "XMin", "cd3_XMax" = "XMax",
                   "cd3_YMin" = "YMin", "cd3_YMax" = "YMax"),
            relationship = "many-to-one")

# Select desired columns and apply basic filters
cd3_data <- cd3_joined %>%
  clean_names(replace = c("\u03bc" = "u")) %>%
  dplyr::select(
    folder_id,
    ImageID = image_id,
    object_id,
    total_cells,
    ki_67_phenotyper_insulin_cells,
    x_ki_67_phenotyper_insulin_positive_cells,
    ki_67_phenotyper_glucagon_cells,
    x_ki_67_phenotyper_glucagon_positive_cells,
    ki_67_phenotyper_acinar_cells,
    x_ki_67_phenotyper_acinar_positive_cells,
    tissue_area_um_2,
    dist_px
  ) %>%
  filter(total_cells > 0)

# Join demographics from no_cutoff_data (already has all donor-level fixes applied)
cd3_data <- cd3_data %>%
  mutate(folder_id = as.character(folder_id),
         ImageID   = as.character(ImageID),
         object_id = as.integer(object_id)) %>%
  left_join(
    no_cutoff_data %>%
      mutate(Donor     = as.character(Donor),
             ImageID   = as.character(ImageID),
             Object.Id = as.integer(Object.Id),
             Region    = as.character(Region),
             log_islet_cells = as.numeric(log_Islet.Cells)) %>%
      dplyr::select(Donor, ImageID, Object.Id, Region,
                    Sex, Age, AgeCat, Diagnosis,
                    Disease.Duration, age_at_onset,
                    Count.ins, Count.glu, log_Islet.Cells) %>%
      distinct(),
    by = c("folder_id" = "Donor", "ImageID", "object_id" = "Object.Id")
  ) %>%
  filter(!is.na(Diagnosis)) %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    log_Islet.Cells_c  = scale(log_Islet.Cells, center = TRUE, scale = FALSE)[, 1]
  )

# Derive proliferation rates using the phenotyper's hormone-cell counts as denominators
cd3_data <- cd3_data %>%
  mutate(
    # Reconstruct absolute counts of Ki-67+ hormone cells from the percent columns
    cd3pos_glu = round(x_ki_67_phenotyper_glucagon_positive_cells *
                        ki_67_phenotyper_glucagon_cells / 100),
    cd3pos_ins = round(x_ki_67_phenotyper_insulin_positive_cells *
                        ki_67_phenotyper_insulin_cells / 100),
    
    # Proliferation rate using the more reliable phenotyper Count.glu / Count.ins
    # as the hormone-cell denominator
    glu_prolif_rate = if_else(Count.glu > 0, cd3pos_glu / Count.glu, NA_real_),
    ins_prolif_rate = if_else(Count.ins > 0, cd3pos_ins / Count.ins, NA_real_),
    
    # Flag biologically-impossible values (cross-algorithm disagreement on small islets)
    glu_prolif_impossible = !is.na(glu_prolif_rate) & glu_prolif_rate > 1,
    ins_prolif_impossible = !is.na(ins_prolif_rate) & ins_prolif_rate > 1,
    
    # NA out the impossible values so they don't contaminate downstream stats
    glu_prolif_rate = if_else(glu_prolif_impossible, NA_real_, glu_prolif_rate),
    ins_prolif_rate = if_else(ins_prolif_impossible, NA_real_, ins_prolif_rate)
  ) %>%
  rename(Donor = folder_id)

rm(cd3_joined, cd3_cols, match_link, all_matches, all_matches_clean); gc()
```

## E6. Sanity checks

```{r d6_cd3_sanity}
cat("=== Ki-67 dataset sanity checks ===\n\n")

cat("Distance distribution (should be tight near 0, max near 15):\n")
print(summary(cd3_data$dist_px))

cat("\nMissing demographics (should be 0):\n")
print(cd3_data %>%
  summarise(
    missing_sex       = sum(is.na(Sex)),
    missing_age       = sum(is.na(Age)),
    missing_diagnosis = sum(is.na(Diagnosis))
  ))

cat("\nAlgorithm disagreements (CD3+ insulin cells > total cells):\n")
cat("  count:", sum(cd3_data$cd3_gt_total, na.rm = TRUE),
    sprintf("(%.3f%% of rows)\n",
            100 * mean(cd3_data$cd3_gt_total, na.rm = TRUE)))
```

## E7. Save CD3 datasets

```{r d7_save_cd3}
write_csv(cd3_data, "Data/cd3_data.csv")
saveRDS(cd3_data, "Data/cd3_data.rds")
cat("Saved: cd3_data.csv —", nrow(cd3_data), "rows,",
    length(unique(cd3_data$Donor)), "donors,",
    length(unique(cd3_data$ImageID)), "images\n")

cd3_data_t1d <- cd3_data %>% filter(Diagnosis == "T1D") %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    log_Islet.Cells_c  = scale(log_Islet.Cells, center = TRUE, scale = FALSE)[, 1],
    Disease.Duration_c  = scale(Disease.Duration, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    age_at_onset   = Age - Disease.Duration,
    age_at_onset_c = scale(age_at_onset, center = TRUE, scale = FALSE)[, 1]
  )

write_csv(cd3_data_t1d, "Data/cd3_data_t1d.csv")
saveRDS(cd3_data_t1d, "Data/cd3_data_t1d.rds")
cat("Saved: cd3_data_t1d.csv —", nrow(cd3_data_t1d), "rows,",
    length(unique(cd3_data_t1d$Donor)), "donors\n")
```
---

# Part F: CD45 Data

## F1. Read and combine all CD45 CSVs

```{r d1_read_ki67_csvs}
cd45_base_path <- "cd45_data"

cd45_files <- list.files(cd45_base_path, pattern = "\\.csv$",
                         recursive = TRUE, full.names = TRUE)
cd45_files <- cd45_files[!grepl("Total_Summary_Results\\.csv$", cd45_files)]

cat("Found", length(cd45_files), "CD3 CSV files\n")

cd45_raw <- cd45_files %>%
  lapply(function(file) {
    tryCatch({
      df <- read.csv(file, stringsAsFactors = FALSE)
      df$source_file <- basename(file)
      df$folder_id <- basename(dirname(dirname(dirname(dirname(file)))))
      df$anatomical_location <- basename(dirname(dirname(dirname(file))))
      return(df)
    }, error = function(e) {
      cat("Error reading file:", file, "\n", e$message, "\n")
      return(NULL)
    })
  }) %>%
  bind_rows() %>%
  mutate(
    folder_id = case_when(
      folder_id == "030" ~ "HPAP030",
      TRUE ~ folder_id
    ),
    ImageID = str_extract(source_file, "\\d+")
  )

cat("CD3 raw rows:", nrow(cd45_raw), "\n")
```

## F2. Dedupe phenotyper data to physical islets

```{r d2_dedupe_physical_islets}
# Each physical islet appears once per algorithm run (Single/Double/Quad) in
# the phenotyper data. For spatial matching we need one canonical row per
# physical islet, keyed by bounding-box coordinates.
phenotyper_islets <- phenotyper_raw %>%
  distinct(ImageID, XMin, XMax, YMin, YMax, .keep_all = TRUE)

cat("Phenotyper rows (all algorithms):", nrow(phenotyper_raw), "\n")
cat("Phenotyper distinct physical islets:", nrow(phenotyper_islets), "\n")
```

## F3. Spatially match CD45 islets to phenotyper islets

```{r d3_spatial_match}
# Greedy nearest-centroid matcher: for each image, pairs islets across the two
# datasets by bounding-box centroid distance, enforcing 1-to-1 pairings.
match_islets_one_image <- function(df1, df2, max_dist = 20) {
  df1 <- df1 %>% mutate(cx = (XMin + XMax)/2, cy = (YMin + YMax)/2, .i = row_number())
  df2 <- df2 %>% mutate(cx = (XMin + XMax)/2, cy = (YMin + YMax)/2, .j = row_number())

  pairs <- expand.grid(.i = df1$.i, .j = df2$.j) %>%
    mutate(dist = sqrt((df1$cx[.i] - df2$cx[.j])^2 +
                       (df1$cy[.i] - df2$cy[.j])^2)) %>%
    filter(dist <= max_dist) %>%
    arrange(dist)

  used1 <- logical(nrow(df1))
  used2 <- logical(nrow(df2))
  keep  <- logical(nrow(pairs))
  for (k in seq_len(nrow(pairs))) {
    if (!used1[pairs$.i[k]] && !used2[pairs$.j[k]]) {
      keep[k] <- TRUE
      used1[pairs$.i[k]] <- TRUE
      used2[pairs$.j[k]] <- TRUE
    }
  }
  pairs <- pairs[keep, ]

  tibble(
    cd45_Object.Id  = df1$Object.Id[pairs$.i],
    count_Object.Id = df2$Object.Id[pairs$.j],
    dist_px         = pairs$dist
  )
}

# Apply per image
all_matches <- phenotyper_islets %>%
  distinct(ImageID) %>%
  pull(ImageID) %>%
  map_dfr(function(img) {
    d1 <- cd45_raw          %>% filter(ImageID == img)
    d2 <- phenotyper_islets %>% filter(ImageID == img)
    if (nrow(d1) == 0 || nrow(d2) == 0) return(NULL)
    match_islets_one_image(d1, d2, max_dist = 20) %>%
      mutate(ImageID = img, .before = 1)
  })

cat("Raw matches:", nrow(all_matches), "\n")
```

## F4. Filter matches by distance and bounding-box size ratio

```{r d4_filter_matches}
# True matches should be both spatially close AND similar in size.
# Attach bounding-box areas from each dataset, compute size ratio, filter.
all_matches_clean <- all_matches %>%
  left_join(
    cd45_raw %>%
      mutate(area1 = (XMax - XMin) * (YMax - YMin)) %>%
      dplyr::select(ImageID, Object.Id, area1,
                    cd45_XMin = XMin, cd45_XMax = XMax,
                    cd45_YMin = YMin, cd45_YMax = YMax),
    by = c("ImageID", "cd45_Object.Id" = "Object.Id")
  ) %>%
  left_join(
    phenotyper_islets %>%
      mutate(area2 = (XMax - XMin) * (YMax - YMin)) %>%
      dplyr::select(ImageID, Object.Id, area2,
                    ph_XMin = XMin, ph_XMax = XMax,
                    ph_YMin = YMin, ph_YMax = YMax),
    by = c("ImageID", "count_Object.Id" = "Object.Id")
  ) %>%
  mutate(area_ratio = pmin(area1, area2) / pmax(area1, area2)) %>%
  filter(dist_px <= 15, area_ratio >= 0.5)

cat("Matched islets after filtering:", nrow(all_matches_clean),
    sprintf("(%.1f%% of CD3 islets)\n",
            100 * nrow(all_matches_clean) / nrow(cd45_raw)))
```

## F5. Build cd45_data and join demographics

```{r d5_build_cd45_data}
# Pull all Ki-67 columns from raw CD3 data (regex handles the "Ki.67" dot).
cd45_cols <- cd45_raw %>%
  dplyr::select(
    ImageID, XMin, XMax, YMin, YMax,
    matches("Ki\\.?_?67", ignore.case = TRUE)
  )

# Join CD3 values onto the FULL phenotyper_raw (not the deduped version)
# using phenotyper coordinates, so every algorithm row for a matched islet
# receives the same CD3 values.
match_link <- all_matches_clean %>%
  dplyr::select(
    ImageID,
    XMin = ph_XMin, XMax = ph_XMax, YMin = ph_YMin, YMax = ph_YMax,
    cd45_XMin, cd45_XMax, cd45_YMin, cd45_YMax,
    dist_px
  )

cd45_joined <- phenotyper_raw %>%
  inner_join(match_link,
             by = c("ImageID", "XMin", "XMax", "YMin", "YMax"),
             relationship = "many-to-one") %>%
  left_join(cd45_cols,
            by = c("ImageID",
                   "cd45_XMin" = "XMin", "cd45_XMax" = "XMax",
                   "cd45_YMin" = "YMin", "cd45_YMax" = "YMax"),
            relationship = "many-to-one")

# Select desired columns and apply basic filters
cd45_data <- cd45_joined %>%
  clean_names(replace = c("\u03bc" = "u")) %>%
  dplyr::select(
    folder_id,
    ImageID = image_id,
    object_id,
    total_cells,
    ki_67_phenotyper_insulin_cells,
    x_ki_67_phenotyper_insulin_positive_cells,
    ki_67_phenotyper_glucagon_cells,
    x_ki_67_phenotyper_glucagon_positive_cells,
    ki_67_phenotyper_acinar_cells,
    x_ki_67_phenotyper_acinar_positive_cells,
    tissue_area_um_2,
    dist_px
  ) %>%
  filter(total_cells > 0)

# Join demographics from no_cutoff_data (already has all donor-level fixes applied)
cd45_data <- cd45_data %>%
  mutate(folder_id = as.character(folder_id),
         ImageID   = as.character(ImageID),
         object_id = as.integer(object_id)) %>%
  left_join(
    no_cutoff_data %>%
      mutate(Donor     = as.character(Donor),
             ImageID   = as.character(ImageID),
             Object.Id = as.integer(Object.Id),
             Region    = as.character(Region),
             log_islet_cells = as.numeric(log_Islet.Cells)) %>%
      dplyr::select(Donor, ImageID, Object.Id, Region,
                    Sex, Age, AgeCat, Diagnosis,
                    Disease.Duration, age_at_onset,
                    Count.ins, Count.glu, log_Islet.Cells) %>%
      distinct(),
    by = c("folder_id" = "Donor", "ImageID", "object_id" = "Object.Id")
  ) %>%
  filter(!is.na(Diagnosis)) %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    log_Islet.Cells_c  = scale(log_Islet.Cells, center = TRUE, scale = FALSE)[, 1]
  )

# Derive proliferation rates using the phenotyper's hormone-cell counts as denominators
cd45_data <- cd45_data %>%
  mutate(
    # Reconstruct absolute counts of Ki-67+ hormone cells from the percent columns
    cd45pos_glu = round(x_ki_67_phenotyper_glucagon_positive_cells *
                        ki_67_phenotyper_glucagon_cells / 100),
    cd45pos_ins = round(x_ki_67_phenotyper_insulin_positive_cells *
                        ki_67_phenotyper_insulin_cells / 100),
    
    # Proliferation rate using the more reliable phenotyper Count.glu / Count.ins
    # as the hormone-cell denominator
    glu_prolif_rate = if_else(Count.glu > 0, cd45pos_glu / Count.glu, NA_real_),
    ins_prolif_rate = if_else(Count.ins > 0, cd45pos_ins / Count.ins, NA_real_),
    
    # Flag biologically-impossible values (cross-algorithm disagreement on small islets)
    glu_prolif_impossible = !is.na(glu_prolif_rate) & glu_prolif_rate > 1,
    ins_prolif_impossible = !is.na(ins_prolif_rate) & ins_prolif_rate > 1,
    
    # NA out the impossible values so they don't contaminate downstream stats
    glu_prolif_rate = if_else(glu_prolif_impossible, NA_real_, glu_prolif_rate),
    ins_prolif_rate = if_else(ins_prolif_impossible, NA_real_, ins_prolif_rate)
  ) %>%
  rename(Donor = folder_id)

rm(cd45_joined, cd45_cols, match_link, all_matches, all_matches_clean); gc()
```

## F6. Sanity checks

```{r d6_cd45_sanity}
cat("=== Ki-67 dataset sanity checks ===\n\n")

cat("Distance distribution (should be tight near 0, max near 15):\n")
print(summary(cd45_data$dist_px))

cat("\nMissing demographics (should be 0):\n")
print(cd45_data %>%
  summarise(
    missing_sex       = sum(is.na(Sex)),
    missing_age       = sum(is.na(Age)),
    missing_diagnosis = sum(is.na(Diagnosis))
  ))

cat("\nAlgorithm disagreements (CD3+ insulin cells > total cells):\n")
cat("  count:", sum(cd45_data$cd45_gt_total, na.rm = TRUE),
    sprintf("(%.3f%% of rows)\n",
            100 * mean(cd45_data$cd45_gt_total, na.rm = TRUE)))
```

## F7. Save CD45 datasets

```{r d7_save_cd45}
write_csv(cd45_data, "Data/cd45_data.csv")
saveRDS(cd45_data, "Data/cd45_data.rds")
cat("Saved: cd45_data.csv —", nrow(cd45_data), "rows,",
    length(unique(cd45_data$Donor)), "donors,",
    length(unique(cd45_data$ImageID)), "images\n")

cd45_data_t1d <- cd45_data %>% filter(Diagnosis == "T1D") %>%
  mutate(
    Age_c              = scale(Age, center = TRUE, scale = FALSE)[, 1],
    log_Islet.Cells_c  = scale(log_Islet.Cells, center = TRUE, scale = FALSE)[, 1],
    Disease.Duration_c  = scale(Disease.Duration, center = TRUE, scale = FALSE)[, 1]
  ) %>%
  mutate(
    age_at_onset   = Age - Disease.Duration,
    age_at_onset_c = scale(age_at_onset, center = TRUE, scale = FALSE)[, 1]
  )

write_csv(cd45_data_t1d, "Data/cd45_data_t1d.csv")
saveRDS(cd45_data_t1d, "Data/cd45_data_t1d.rds")
cat("Saved: cd45_data_t1d.csv —", nrow(cd45_data_t1d), "rows,",
    length(unique(cd45_data_t1d$Donor)), "donors\n")
```
---

# Summary

```{r summary}
cat("=== Pipeline Complete ===\n\n")

datasets <- list(
  "no_cutoff_data.csv"   = no_cutoff_data,
  "no_cutoff_data_t1d.csv"   = no_cutoff_data_t1d,
  "full_data.csv"        = full_data,
  "full_data_t1d.csv"        = full_data_t1d,
  "quad_data.csv"        = quad_data,
  "quad_data_t1d.csv"        = quad_data_t1d,
  "area_data_joined.csv" = area_data_joined,
  "area_data_joined_t1d.csv" = area_data_joined_t1d,
  "density_data.csv"     = density_dataset,
  "density_data_t1d.csv"     = density_dataset_t1d,
  "ki67_data.csv"           = ki67_data,
  "ki67_data_t1d.csv"       = ki67_data_t1d,
  "cd3_data.csv"           = cd3_data,
  "cd3_data_t1d.csv"       = cd3_data_t1d,
  "cd45_data.csv"           = cd45_data,
  "cd45_data_t1d.csv"       = cd45_data_t1d
)

for (nm in names(datasets)) {
  d <- datasets[[nm]]
  cat(sprintf("%-25s  %7d rows  %3d donors  %4d images\n",
              nm, nrow(d),
              length(unique(d$Donor)),
              length(unique(d$ImageID))))
}
```

```{r}
sessionInfo()
```

