---
title: "Per-cell rate models (negative binomial with denominator offset)"
output: html_document
date: "`r Sys.Date()`"
---

Six models. Outcome is the marker-positive count in a bin; the count of cells
that could have been positive enters as `offset(log(denominator))`, so the
coefficients are **rate ratios per cell of that type**.

| model | outcome | offset |
|---|---|---|
| `ki67_ins_rate` | `insulin_cells` | `log(Count.ins)` |
| `ki67_glu_rate` | `glucagon_cells` | `log(Count.glu)` |
| `ki67_aci_rate` | `acinar_cells` | `log(Count.Acinar)` |
| `cd3_ins_rate` | `insulin_cells` | `log(Count.ins)` |
| `cd3_glu_rate` | `glucagon_cells` | `log(Count.glu)` |
| `cd3_aci_rate` | `acinar_cells` | `log(Count.Acinar)` |

## One term is removed from the formula, deliberately

The primary models use

```
(Diagnosis + Region + Sex + Age_c)^2 + Diagnosis:Region:Age_c
```

which is your formula with `log_Islet.Cells_c` and its interactions dropped.
Two reasons, and the second is the important one.

**It is no longer the exposure.** In the count models `log_Islet.Cells_c`
carried the "bigger islets have more cells" scaling. The offset now does that
job, and does it for the *relevant* cell type rather than the total.

**Keeping it would condition away the exposure of interest.** With
`log(Count.ins)` in the offset and `log(Islet.Cells)` as a covariate, the model
conditions on both the beta-cell count and the total — that is, on the beta-cell
*fraction*. But the beta fraction is 77% in ND and 22% in T1D. Those
distributions barely overlap, so a Diagnosis contrast holding fraction fixed is
estimated almost entirely by extrapolation. That is a positivity violation, and
it is the kind that produces a confident number pointing the wrong way.

The version *with* `log_Islet.Cells_c` is still fitted, as
`*_rate_size`, so the difference is visible rather than asserted. Treat those
as diagnostics, not results.

A consequence worth noting: with `log_Islet.Cells_c` gone there is no slope to
randomise, so the primary models use `(1|Donor) + (1|Donor:ImageID)` —
**exactly the random-effect structure of your earlier scripts**. The one
departure I had introduced disappears here.

## The restriction, and why it limits the claim

`offset(log(D))` is undefined when `D = 0`, so each model drops islets with no
cells of the relevant type. That is not a nuisance — the donors it removes are
systematic. Four of 21 Ki-67 T1D donors and six of 20 CD3 T1D donors have zero
insulin+ cells in every islet, presumably the longest-duration cases. So a
beta-cell rate result applies to donors with residual beta mass and is silent
about donors with none. The restriction chunk names the dropped donors with
their disease durations so this can be stated precisely.

Centring is *not* recomputed on the restricted subsets, so coefficients stay
comparable with the count models. The subset mean of `Age_c` is printed for each
model, since the Diagnosis contrast is evaluated at `Age_c = 0` and that is no
longer exactly the subset mean.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(glmmTMB)

MODEL_DIR <- "Marker_Rate_Models"
if (!dir.exists(MODEL_DIR)) dir.create(MODEL_DIR)

REFIT_ALL      <- FALSE
STAIN_HANDLING <- "double_only"      # must match marker_models_v6.Rmd
RUN_SIZE_SENS  <- TRUE               # the *_rate_size diagnostics
FAMILIES       <- c("nb1", "nb2", "zinb1")
TMB_CTRL       <- glmmTMBControl(optCtrl = list(iter.max = 1000, eval.max = 1000))

MARKERS <- c("ki67", "cd3")
BINS <- tibble(
  bin = c("ins", "glu", "aci"),
  num = c("insulin_cells", "glucagon_cells", "acinar_cells"),
  den = c("Count.ins", "Count.glu", "Count.Acinar")
)
```

```{r load}
get_marker <- function(tag) {
  nm <- paste0(tag, "_data")
  if (exists(nm, envir = globalenv())) return(get(nm, envir = globalenv()))
  p <- sprintf("Data/%s_data.csv", tag)
  if (!file.exists(p)) stop("missing ", p)
  df <- readr::read_csv(p, show_col_types = FALSE) %>%
    mutate(Diagnosis = factor(Diagnosis, levels = c("ND", "T1D")),
           Region    = factor(Region, levels = c("Head", "Body", "Tail")),
           Sex       = factor(Sex, levels = c("Female", "Male")),
           Donor     = factor(Donor),
           ImageID   = factor(ImageID))
  if (identical(STAIN_HANDLING, "double_only")) {
    df <- df[!is.na(df$Stain) & df$Stain == "Double", , drop = FALSE]
  }
  droplevels(df)
}
dat <- lapply(setNames(MARKERS, MARKERS), get_marker)
for (tag in MARKERS) cat(sprintf("%-5s %7d islets, %2d donors\n", tag,
  nrow(dat[[tag]]), dplyr::n_distinct(dat[[tag]]$Donor)))
```

# What the denominator restriction costs

```{r restriction}
subsets <- list()
restrict_rows <- list()

for (tag in MARKERS) {
  d <- dat[[tag]]
  for (i in seq_len(nrow(BINS))) {
    id  <- paste0(tag, "_", BINS$bin[i])
    den <- BINS$den[i]
    keep <- !is.na(d[[den]]) & d[[den]] > 0
    sub  <- droplevels(d[keep, , drop = FALSE])
    subsets[[id]] <- sub

    all_don  <- unique(as.character(d$Donor))
    kept_don <- unique(as.character(sub$Donor))
    lost     <- setdiff(all_don, kept_don)

    restrict_rows[[id]] <- tibble(
      model = id, offset = den,
      islets_all = nrow(d), islets_kept = sum(keep),
      pct_islets_lost = 100 * (1 - sum(keep) / nrow(d)),
      donors_all = length(all_don), donors_kept = length(kept_don),
      donors_lost = length(lost),
      lost_ND  = sum(d$Diagnosis[match(lost, as.character(d$Donor))] == "ND"),
      lost_T1D = sum(d$Diagnosis[match(lost, as.character(d$Donor))] == "T1D"),
      mean_Age_c_subset = mean(sub$Age_c))

    if (length(lost) > 0) {
      info <- d %>%
        filter(as.character(Donor) %in% lost) %>%
        distinct(Donor, Diagnosis, Disease.Duration) %>%
        arrange(Diagnosis, Disease.Duration)
      cat(sprintf("\n%s: %d donor(s) dropped (no %s anywhere)\n", id, length(lost), den))
      print(as.data.frame(info), row.names = FALSE)
    }
  }
}

cat("\n=== restriction summary ===\n")
print(as.data.frame(bind_rows(restrict_rows)), digits = 4)
cat("\nDropped donors are the limit on the claim, not a rounding error: a rate\n")
cat("result describes donors that still have cells of that type.\n")
```

```{r betabinom-feasibility}
# Beta-binomial would give a tighter model than a count with an offset, but it
# needs numerator <= denominator, and the two come from different classifiers
# with independent segmentation. Check rather than assume.
cat("=== beta-binomial feasibility: numerator > denominator? ===\n")
print(as.data.frame(bind_rows(lapply(MARKERS, function(tag) {
  d <- dat[[tag]]
  bind_rows(lapply(seq_len(nrow(BINS)), function(i) {
    n <- d[[BINS$num[i]]]; D <- d[[BINS$den[i]]]
    tibble(marker = tag, bin = BINS$bin[i],
           n_violations = sum(n > D),
           pct = 100 * mean(n > D),
           worst_excess = if (any(n > D)) as.integer(max((n - D)[n > D])) else 0L)
  }))
}))), digits = 4)
cat("\nUnder ~1% a beta-binomial is defensible after dropping the violations.\n")
cat("Above that the offset model is the right choice: it does not require the\n")
cat("numerator to be a subset of the denominator.\n")
```

# Formulas

```{r formulas}
# Your structure, minus log_Islet.Cells_c - the offset has taken over its role.
RHS_RATE <- paste(
  "(Diagnosis + Region + Sex + Age_c)^2",
  "+ Diagnosis:Region:Age_c")

# Diagnostic only: conditions on the beta fraction, which barely overlaps
# between groups. See the header.
RHS_RATE_SIZE <- paste(
  "(Diagnosis + Region + Sex + Age_c + log_Islet.Cells_c)^2",
  "+ Diagnosis:Region:log_Islet.Cells_c",
  "+ Diagnosis:Region:Age_c")

RE_PRIMARY <- "(1 | Donor) + (1 | Donor:ImageID)"          # as in the earlier scripts
RE_SIZE    <- c("(1 + log_Islet.Cells_c | Donor) + (1 | Donor:ImageID)",
                "(1 | Donor) + (0 + log_Islet.Cells_c | Donor) + (1 | Donor:ImageID)",
                "(1 | Donor) + (1 | Donor:ImageID)")
```

# Fitting

```{r fit-helpers}
FAM_OBJ <- list(nb1 = nbinom1(), nb2 = nbinom2(), zinb1 = nbinom1())
zi_of   <- function(fam, zi_rhs) if (grepl("^zi", fam)) zi_rhs else ~0

build_rate_formula <- function(y, rhs, den, re) {
  f <- as.formula(sprintf("%s ~ %s + offset(log(%s)) + %s", y, rhs, den, re))
  environment(f) <- globalenv()
  f
}

converged_ok <- function(m) {
  if (is.null(m)) return(FALSE)
  tryCatch(isTRUE(m$sdr$pdHess) &&
             identical(as.integer(m$fit$convergence), 0L) &&
             all(is.finite(sqrt(diag(vcov(m, full = TRUE))))),
           error = function(e) FALSE)
}

degenerate_terms <- function(m, est_max = 15, se_max = 50) {
  if (is.null(m)) return("NULL")
  co <- tryCatch(summary(m)$coefficients, error = function(e) NULL)
  if (is.null(co)) return("no coefficient table")
  out <- character(0)
  for (cmp in names(co)) {
    tab <- co[[cmp]]
    if (is.null(tab) || !is.matrix(tab) || !nrow(tab)) next
    bad <- !is.finite(tab[, 1]) | !is.finite(tab[, 2]) |
      abs(tab[, 1]) > est_max | tab[, 2] > se_max
    if (any(bad)) out <- c(out, sprintf("%s: %s", cmp,
                                        paste(rownames(tab)[bad], collapse = ", ")))
  }
  out
}

fit_cached <- function(name, fn) {
  path <- file.path(MODEL_DIR, paste0(name, ".rds"))
  if (!REFIT_ALL && file.exists(path)) return(readRDS(path))
  t0 <- Sys.time(); m <- fn()
  cat(sprintf("  [fit] %-30s %5.1f min%s\n", name,
              as.numeric(difftime(Sys.time(), t0, units = "mins")),
              if (is.null(m)) "  FAILED" else ""))
  if (!is.null(m)) saveRDS(m, path)
  m
}

fit_re <- function(y, rhs, den, data, fam, zi_rhs, re_options) {
  for (re in re_options) {
    m <- tryCatch(glmmTMB(build_rate_formula(y, rhs, den, re), data = data,
                          family = FAM_OBJ[[fam]], ziformula = zi_of(fam, zi_rhs),
                          control = TMB_CTRL),
                  error = function(e) NULL)
    if (converged_ok(m)) { attr(m, "re_used") <- re; attr(m, "fam_name") <- fam; return(m) }
  }
  NULL
}

# Family by AIC on the actual formula; non-convergence and boundary fits veto.
fit_rate_model <- function(id, y, rhs, den, data, zi_rhs, re_options) {
  fits <- list(); rows <- list()
  for (f in FAMILIES) {
    m <- fit_cached(paste0(id, "_", f),
                    function() fit_re(y, rhs, den, data, f, zi_rhs, re_options))
    fits[f] <- list(m)
    deg <- degenerate_terms(m)
    rows[[f]] <- tibble(family = f, AIC = if (converged_ok(m)) AIC(m) else NA_real_,
                        ok = converged_ok(m), degenerate = length(deg) > 0,
                        note = paste(deg, collapse = "; "))
  }
  tab <- bind_rows(rows) %>% arrange(AIC) %>% mutate(dAIC = AIC - min(AIC, na.rm = TRUE))
  usable <- tab %>% filter(ok, !degenerate, is.finite(AIC))
  if (!nrow(usable)) {
    cat("  NO USABLE FAMILY for", id, "\n"); print(as.data.frame(tab), digits = 6)
    return(list(model = NULL, family = NA_character_, table = tab))
  }
  win <- usable$family[which.min(usable$AIC)]
  cat(sprintf("  %-18s family %-5s  AIC %10.1f%s\n", id, win, AIC(fits[[win]]),
              if (any(tab$degenerate)) paste0("  (vetoed: ",
                paste(tab$family[tab$degenerate], collapse = ", "), ")") else ""))
  list(model = fits[[win]], family = win, table = tab)
}
```

```{r fit-primary}
rate_models <- list(); rate_fam <- character(0); rate_tabs <- list()

cat("=== primary rate models (offset, no islet-size term) ===\n")
for (tag in MARKERS) {
  for (i in seq_len(nrow(BINS))) {
    key <- paste0(tag, "_", BINS$bin[i])
    id  <- paste0(key, "_rate")
    res <- fit_rate_model(id, BINS$num[i], RHS_RATE, BINS$den[i],
                          subsets[[key]], ~ Diagnosis, RE_PRIMARY)
    rate_models[id] <- list(res$model)
    rate_tabs[id]   <- list(res$table)
    rate_fam[[id]]  <- res$family
    assign(id, res$model, envir = globalenv())
  }
}
```

```{r fit-size-sens, eval = RUN_SIZE_SENS}
cat("=== diagnostic: same models WITH log_Islet.Cells_c ===\n")
cat("These condition on the beta fraction. Read the header before using them.\n")
size_models <- list(); size_fam <- character(0)
for (tag in MARKERS) {
  for (i in seq_len(nrow(BINS))) {
    key <- paste0(tag, "_", BINS$bin[i])
    id  <- paste0(key, "_rate_size")
    res <- fit_rate_model(id, BINS$num[i], RHS_RATE_SIZE, BINS$den[i],
                          subsets[[key]], ~ Diagnosis, RE_SIZE)
    size_models[id] <- list(res$model)
    size_fam[[id]]  <- res$family
    assign(id, res$model, envir = globalenv())
  }
}
```

# Rate ratios

```{r rate-ratios}
# With an offset, exp(DiagnosisT1D) is the rate ratio directly. For a
# zero-inflated fit it must also be scaled by the change in the non-zero
# probability, since a structural zero contributes no cells at all.
rate_ratio <- function(m) {
  if (is.null(m)) return(NULL)
  co <- summary(m)$coefficients$cond
  if (!"DiagnosisT1D" %in% rownames(co)) return(NULL)
  est <- co["DiagnosisT1D", "Estimate"]; se <- co["DiagnosisT1D", "Std. Error"]
  zi <- fixef(m)$zi
  scale <- 1; p0 <- NA_real_; p1 <- NA_real_
  if (length(zi)) {
    p0 <- plogis(zi[["(Intercept)"]])
    p1 <- if ("DiagnosisT1D" %in% names(zi))
      plogis(zi[["(Intercept)"]] + zi[["DiagnosisT1D"]]) else p0
    scale <- (1 - p1) / (1 - p0)
  }
  tibble(cond_rate_ratio = exp(est),
         lwr = exp(est - 1.96 * se), upr = exp(est + 1.96 * se),
         p = co["DiagnosisT1D", "Pr(>|z|)"],
         zi_ND = p0, zi_T1D = p1,
         marginal_rate_ratio = exp(est) * scale)
}

# Crude rates recomputed here so this file stands alone.
crude <- bind_rows(lapply(MARKERS, function(tag) {
  d <- dat[[tag]]
  bind_rows(lapply(seq_len(nrow(BINS)), function(i) {
    g <- d %>% group_by(Diagnosis) %>%
      summarise(N = mean(.data[[BINS$num[i]]]),
                D = mean(.data[[BINS$den[i]]]), .groups = "drop")
    nd <- g[g$Diagnosis == "ND", ]; t1 <- g[g$Diagnosis == "T1D", ]
    tibble(model = paste0(tag, "_", BINS$bin[i], "_rate"),
           crude_ND_pct = 100 * nd$N / nd$D, crude_T1D_pct = 100 * t1$N / t1$D,
           crude_fold_rate = (t1$N / t1$D) / (nd$N / nd$D))
  }))
}))

res <- bind_rows(lapply(names(rate_models), function(id) {
  rr <- rate_ratio(rate_models[[id]])
  if (is.null(rr)) return(NULL)
  rr %>% mutate(model = id, family = rate_fam[[id]], .before = 1)
})) %>%
  left_join(crude, by = "model") %>%
  left_join(bind_rows(restrict_rows) %>%
              mutate(model = paste0(model, "_rate")) %>%
              dplyr::select(model, islets_kept, donors_kept, donors_lost),
            by = "model")

cat("=== per-cell rate ratios, T1D vs ND ===\n")
cat("Region = Head, Sex = Female, mean Age. Adjusted for Region, Sex, Age and donor.\n\n")
print(as.data.frame(res %>% dplyr::select(model, family, cond_rate_ratio, lwr, upr, p,
                                          marginal_rate_ratio, crude_fold_rate,
                                          islets_kept, donors_kept, donors_lost)),
      digits = 4)

cat("\nzero-inflation, where selected:\n")
print(as.data.frame(res %>% filter(!is.na(zi_ND)) %>%
                      dplyr::select(model, zi_ND, zi_T1D)), digits = 3)

cat("\ncrude rates for reference (unadjusted, %):\n")
print(as.data.frame(res %>% dplyr::select(model, crude_ND_pct, crude_T1D_pct)), digits = 3)
```

```{r size-comparison, eval = RUN_SIZE_SENS}
cat("=== effect of adding log_Islet.Cells_c (diagnostic) ===\n")
cmp <- bind_rows(lapply(names(rate_models), function(id) {
  a <- rate_ratio(rate_models[[id]])
  b <- rate_ratio(size_models[[paste0(id, "_size")]])
  if (is.null(a) || is.null(b)) return(NULL)
  tibble(model = id,
         primary_ratio = a$marginal_rate_ratio, primary_p = a$p,
         with_size_ratio = b$marginal_rate_ratio, with_size_p = b$p,
         ratio_of_ratios = b$marginal_rate_ratio / a$marginal_rate_ratio,
         # Guarded: sign(NA) is NA, and `if (NA)` errors rather than skipping.
         flag = {
           ra <- a$marginal_rate_ratio; rb <- b$marginal_rate_ratio
           if (!is.finite(ra) || !is.finite(rb) || ra <= 0 || rb <= 0 ||
               is.na(a$p) || is.na(b$p)) "not comparable"
           else if (sign(log(ra)) != sign(log(rb))) "DIRECTION FLIP"
           else if ((a$p < 0.05) != (b$p < 0.05)) "sig flip"
           else ""
         })
}))
print(as.data.frame(cmp), digits = 4)
cat("\nA direction flip here is the positivity problem showing itself, not a\n")
cat("finding: conditioning on both the cell count and the total conditions on\n")
cat("the composition, which is what differs between the groups.\n")
```

```{r full-summaries}
for (id in names(rate_models)) {
  cat(sprintf("\n%s\n===== %s =====\n", strrep("=", 72), id))
  if (is.null(rate_models[[id]])) cat("  no usable fit\n") else print(summary(rate_models[[id]]))
}
```

```{r save}
saveRDS(list(families = rate_fam, family_tables = rate_tabs,
             restriction = bind_rows(restrict_rows), results = res),
        file.path(MODEL_DIR, "rate_results.rds"))
cat("\nModels cached in", MODEL_DIR, "\n")
cat("Objects: summary(ki67_glu_rate), summary(cd3_aci_rate), etc.\n")
```

```{r session-info}
sessionInfo()
```
