---
title: "Can survivor selection explain the alpha-cell shift?"
output: html_document
date: "`r Sys.Date()`"
---

## The test

If T1D islets are simply the alpha-rich survivors of a formerly ND-like
population, then selection has only removed islets. It cannot have created any,
and it certainly cannot have created alpha-rich ones. So for every level of
alpha fraction *a*:

```
islets per mm² at alpha fraction a, in T1D   <=   the same quantity in ND
```

**Any bin where T1D density exceeds ND density falsifies pure survivor
selection.** No distributional assumptions, no model. The same inequality
applies to alpha cells per mm² of tissue.

Comparing distribution *shapes* is the weaker version of this. The density
inequality is what can actually settle it, and it needs both composition
(`ratio_data`) and a tissue denominator (`density_data`) — which you already
have.

## What this can and cannot establish

It can **falsify pure survivor selection**. It cannot confirm
transdifferentiation: ruling out one non-proliferative mechanism leaves the
others (transdifferentiation, reduced alpha-cell turnover) undistinguished.
Nor can it rule out selection acting *partially* alongside another mechanism.

Four caveats that belong in any write-up of this:

1. **Islet detection is composition-dependent.** An islet that lost its beta
   cells may fall below the `Islet.Cells >= 15` threshold and vanish from the
   dataset — a selection effect created by the pipeline, not by biology. This
   works *against* finding a violation, so a violation is still informative. The
   sensitivity chunk repeats everything on `no_cutoff_data` (`Islet.Cells > 0`).
2. **The tissue denominator can move.** `full_tissue_area` is classified minus
   duct minus glass. Exocrine atrophy in T1D would shrink it and inflate every
   T1D density. Checked directly below.
3. **Two-dimensional sections**, so these are areal densities, not volumetric.
4. **Cross-sectional.** No islet is observed before and after.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)

N_BINS <- 10          # alpha-fraction bins for the density inequality
USE_NO_CUTOFF <- TRUE # also run everything without the Islet.Cells >= 15 filter
```

```{r load}
read_any <- function(stem) {
  for (p in sprintf("Data/%s.%s", stem, c("rds", "csv"))) {
    if (file.exists(p)) return(if (grepl("rds$", p)) readRDS(p)
                               else readr::read_csv(p, show_col_types = FALSE))
  }
  stop("cannot find Data/", stem, ".rds or .csv")
}

# ratio_data already excludes Single-stain images (where Count.glu is
# structurally 0) and donor 6473, so alpha fraction is well defined on it.
ratio_data <- read_any("ratio_data")
density    <- read_any("density_data")

norm_img <- function(x) as.character(suppressWarnings(as.numeric(as.character(x))))

prep_comp <- function(df) {
  df %>%
    mutate(Diagnosis  = factor(Diagnosis, levels = c("ND", "T1D")),
           Donor      = as.character(Donor),
           ImageID    = norm_img(ImageID),
           alpha_frac = Count.glu / (Count.ins + Count.glu)) %>%
    filter(is.finite(alpha_frac))
}

comp <- prep_comp(ratio_data)
dens <- density %>%
  mutate(Donor = as.character(Donor), ImageID = norm_img(ImageID)) %>%
  dplyr::select(Donor, ImageID, full_tissue_area) %>%
  distinct()

cat(sprintf("composition: %d islets, %d donors\n", nrow(comp),
            dplyr::n_distinct(comp$Donor)))
matched <- comp %>% semi_join(dens, by = c("Donor", "ImageID"))
cat(sprintf("with a tissue-area denominator: %d islets (%.1f%%), %d donors\n",
            nrow(matched), 100 * nrow(matched) / nrow(comp),
            dplyr::n_distinct(matched$Donor)))
if (nrow(matched) < 0.8 * nrow(comp)) {
  cat("WARNING: many islets have no matching image in density_data - check the\n")
  cat("ImageID join before trusting the density tests.\n")
}
```

# Caveat 2 first: does the denominator move?

```{r tissue-area}
# If full_tissue_area is systematically smaller in T1D, every T1D density is
# inflated and a violation could be an artefact. Check before anything else.
ta <- comp %>%
  distinct(Donor, Diagnosis, ImageID) %>%
  inner_join(dens, by = c("Donor", "ImageID")) %>%
  group_by(Donor, Diagnosis) %>%
  summarise(mean_area = mean(full_tissue_area), images = n(), .groups = "drop")

print(as.data.frame(ta %>% group_by(Diagnosis) %>%
  summarise(donors = n(), median_area = median(mean_area),
            q25 = quantile(mean_area, .25), q75 = quantile(mean_area, .75),
            .groups = "drop")), digits = 4)
cat(sprintf("\nWilcoxon on donor means: p = %.4f\n",
            suppressWarnings(wilcox.test(mean_area ~ Diagnosis, data = ta)$p.value)))
cat("A large difference here means the density comparisons need the areal\n")
cat("denominator reported alongside, not just the ratio.\n")
```

# Test 1: distributions and overlap

```{r overlap}
cat("=== alpha fraction by diagnosis ===\n")
print(as.data.frame(comp %>% group_by(Diagnosis) %>%
  summarise(islets = n(), donors = dplyr::n_distinct(Donor),
            mean = mean(alpha_frac), median = median(alpha_frac),
            q10 = quantile(alpha_frac, .10), q25 = quantile(alpha_frac, .25),
            q75 = quantile(alpha_frac, .75), q90 = quantile(alpha_frac, .90),
            .groups = "drop")), digits = 3)

nd <- comp$alpha_frac[comp$Diagnosis == "ND"]
t1 <- comp$alpha_frac[comp$Diagnosis == "T1D"]

thr <- median(t1)
cat(sprintf("\nmedian T1D alpha fraction: %.3f\n", thr))
cat(sprintf("ND islets at or above it: %.2f%% (%d of %d)\n",
            100 * mean(nd >= thr), sum(nd >= thr), length(nd)))
cat(sprintf("T1D islets inside the ND 10-90%% range [%.3f, %.3f]: %.1f%%\n",
            quantile(nd, .10), quantile(nd, .90),
            100 * mean(t1 >= quantile(nd, .10) & t1 <= quantile(nd, .90))))
cat("\nIf only a small fraction of ND islets are as alpha-rich as the typical\n")
cat("T1D islet, pure selection requires destroying almost everything else -\n")
cat("which Test 2 checks against the observed islet density.\n")
```

# Test 2: the density inequality

```{r density-inequality}
# Islets per mm2 in each alpha-fraction bin, by diagnosis. Under pure survivor
# selection T1D must be <= ND in EVERY bin.
bin_edges <- seq(0, 1, length.out = N_BINS + 1)

# Tissue area is recorded per image, so a donor's area is the sum over the
# images that actually contributed islets.
donor_area <- comp %>%
  distinct(Donor, Diagnosis, ImageID) %>%
  inner_join(dens, by = c("Donor", "ImageID")) %>%
  group_by(Donor, Diagnosis) %>%
  summarise(area_mm2 = sum(full_tissue_area), .groups = "drop")

counts_by_bin <- comp %>%
  inner_join(dens, by = c("Donor", "ImageID")) %>%
  mutate(bin = cut(alpha_frac, bin_edges, include.lowest = TRUE)) %>%
  count(Donor, Diagnosis, bin, name = "n_islets") %>%
  complete(nesting(Donor, Diagnosis), bin, fill = list(n_islets = 0)) %>%
  inner_join(donor_area, by = c("Donor", "Diagnosis")) %>%
  mutate(islets_per_mm2 = n_islets / area_mm2)

# Donor is the unit: mean of donor densities, not pooled islets over pooled area.
ineq <- counts_by_bin %>%
  group_by(Diagnosis, bin) %>%
  summarise(donors = n(), mean_density = mean(islets_per_mm2), .groups = "drop") %>%
  pivot_wider(names_from = Diagnosis, values_from = c(donors, mean_density)) %>%
  mutate(ratio_T1D_over_ND = mean_density_T1D / mean_density_ND,
         violates = mean_density_T1D > mean_density_ND)

cat("=== islets per mm2 by alpha-fraction bin ===\n")
cat("Under pure survivor selection every ratio must be <= 1.\n\n")
print(as.data.frame(ineq), digits = 4)

# A bin where ND density is exactly 0 but T1D is positive gives ratio = Inf.
# That is the strongest form of falsification: islets of that composition do not
# exist in ND at all, so they cannot be ND survivors.
empty_nd <- sum(ineq$mean_density_ND == 0 & ineq$mean_density_T1D > 0, na.rm = TRUE)
if (empty_nd > 0) {
  cat(sprintf("\n%d bin(s) have NO ND islets but do have T1D islets - those\n",
              empty_nd))
  cat("compositions do not occur in ND at any density.\n")
}

nv <- sum(ineq$violates, na.rm = TRUE)
cat(sprintf("\nbins where T1D density EXCEEDS ND: %d of %d\n", nv, nrow(ineq)))
if (nv > 0) {
  cat("=> pure survivor selection is falsified. Islets of that composition are\n")
  cat("   more common per unit tissue in T1D than they ever were in ND, so they\n")
  cat("   cannot be a surviving subset of the ND population.\n")
  cat("   Largest excess:\n")
  print(as.data.frame(ineq %>% filter(violates) %>%
    arrange(desc(ratio_T1D_over_ND)) %>% head(3)), digits = 4)
} else {
  cat("=> no violation; selection remains viable on this test. Check whether the\n")
  cat("   overall islet density drop in Test 3 is large enough to supply the\n")
  cat("   T1D population from the ND alpha-rich tail.\n")
}

# Monotonicity: the density ratio is proportional to P(survive | alpha fraction),
# which should be non-decreasing if alpha-richness is what protects an islet.
r <- ineq$ratio_T1D_over_ND
r <- r[is.finite(r)]
cat(sprintf("\nratio monotone non-decreasing across bins: %s\n",
            all(diff(r) >= 0)))
cat("Non-monotone means no single survival-probability function of alpha\n")
cat("fraction can map the ND distribution onto the T1D one.\n")
```

# Test 3: absolute cell burden per mm²

```{r burden}
# Alpha cells per mm2 of tissue cannot increase under pure selection.
burden <- comp %>%
  inner_join(dens, by = c("Donor", "ImageID")) %>%
  group_by(Donor, Diagnosis) %>%
  summarise(islets = n(), alpha = sum(Count.glu), beta = sum(Count.ins),
            endocrine = sum(Count.ins + Count.glu), .groups = "drop") %>%
  inner_join(donor_area, by = c("Donor", "Diagnosis")) %>%
  mutate(islets_per_mm2 = islets / area_mm2,
         alpha_per_mm2  = alpha / area_mm2,
         beta_per_mm2   = beta / area_mm2,
         endo_per_mm2   = endocrine / area_mm2)

summ <- burden %>% group_by(Diagnosis) %>%
  summarise(donors = n(),
            across(c(islets_per_mm2, alpha_per_mm2, beta_per_mm2, endo_per_mm2),
                   median), .groups = "drop")
cat("=== per mm2 of tissue, donor medians ===\n")
print(as.data.frame(summ), digits = 4)

folds <- summ %>%
  pivot_longer(-c(Diagnosis, donors)) %>%
  pivot_wider(names_from = Diagnosis, values_from = value) %>%
  mutate(fold_T1D_over_ND = T1D / ND)
cat("\n")
print(as.data.frame(folds %>% dplyr::select(name, ND, T1D, fold_T1D_over_ND)), digits = 4)

for (v in c("alpha_per_mm2", "islets_per_mm2", "beta_per_mm2")) {
  a <- burden[[v]][burden$Diagnosis == "T1D"]
  b <- burden[[v]][burden$Diagnosis == "ND"]
  p <- suppressWarnings(wilcox.test(a, b)$p.value)
  cat(sprintf("%-16s ND %.4f vs T1D %.4f, Wilcoxon p = %.4g\n",
              v, median(b), median(a), p))
}
cat("\nalpha_per_mm2 fold > 1 => alpha cells were CREATED, not merely revealed\n")
cat("by the loss of beta cells. That is incompatible with pure selection and\n")
cat("is the single most direct evidence on the question.\n")
```

# Test 4: are alpha-rich ND islets like T1D islets?

```{r matched}
# If T1D islets are surviving alpha-rich ND islets, then ND islets in the T1D
# composition range should resemble them in size and absolute cell counts.
lo <- quantile(t1, .25); hi <- quantile(t1, .75)
cat(sprintf("matching on alpha fraction in the T1D IQR [%.3f, %.3f]\n\n", lo, hi))

mt <- comp %>%
  filter(alpha_frac >= lo, alpha_frac <= hi) %>%
  group_by(Diagnosis) %>%
  summarise(islets = n(), donors = dplyr::n_distinct(Donor),
            median_total = median(Count.ins + Count.glu),
            median_alpha = median(Count.glu),
            median_beta  = median(Count.ins),
            mean_total = mean(Count.ins + Count.glu),
            mean_alpha = mean(Count.glu), .groups = "drop")
print(as.data.frame(mt), digits = 4)
cat("\nIf the ND row has far fewer cells than the T1D row, the alpha-rich ND\n")
cat("islets are small ones and cannot be the origin of full-sized T1D islets.\n")
```

# Test 5: does composition track disease duration?

```{r duration}
# Selection at diagnosis is a one-off: alpha fraction should be flat across
# duration. Progressive conversion predicts a gradient.
t1d_only <- comp %>% filter(Diagnosis == "T1D")
if ("Disease.Duration" %in% names(t1d_only)) {
  dd <- t1d_only %>%
    group_by(Donor, Disease.Duration) %>%
    summarise(alpha_frac = mean(alpha_frac), islets = n(), .groups = "drop")
  print(as.data.frame(dd %>% arrange(Disease.Duration)), digits = 3)
  ct <- suppressWarnings(cor.test(dd$Disease.Duration, dd$alpha_frac,
                                  method = "spearman"))
  cat(sprintf("\nSpearman across %d T1D donors: rho = %.3f, p = %.4f\n",
              nrow(dd), ct$estimate, ct$p.value))
  cat("Flat is consistent with a one-off event at diagnosis; a positive\n")
  cat("gradient is consistent with progressive conversion.\n")
} else {
  cat("Disease.Duration not present in ratio_data\n")
}
```

# Sensitivity: without the size threshold

```{r sensitivity, eval = USE_NO_CUTOFF}
# The Islet.Cells >= 15 filter can delete islets that shrank below it, which is
# selection imposed by the pipeline. Repeat the two decisive tests on
# no_cutoff_data (Islet.Cells > 0), excluding Single-stain rows so Count.glu is
# meaningful.
nc <- read_any("no_cutoff_data") %>%
  filter(Stain != "Single", !is.na(Stain)) %>%
  prep_comp()

nc_area <- nc %>% distinct(Donor, Diagnosis, ImageID) %>%
  inner_join(dens, by = c("Donor", "ImageID")) %>%
  group_by(Donor, Diagnosis) %>%
  summarise(area_mm2 = sum(full_tissue_area), .groups = "drop")

nc_burden <- nc %>%
  inner_join(dens, by = c("Donor", "ImageID")) %>%
  group_by(Donor, Diagnosis) %>%
  summarise(islets = n(), alpha = sum(Count.glu), beta = sum(Count.ins),
            .groups = "drop") %>%
  inner_join(nc_area, by = c("Donor", "Diagnosis")) %>%
  mutate(islets_per_mm2 = islets / area_mm2,
         alpha_per_mm2 = alpha / area_mm2, beta_per_mm2 = beta / area_mm2)

cat("=== no size threshold: per mm2, donor medians ===\n")
print(as.data.frame(nc_burden %>% group_by(Diagnosis) %>%
  summarise(donors = n(), across(ends_with("per_mm2"), median), .groups = "drop")),
  digits = 4)
cat(sprintf("\nalpha_per_mm2 Wilcoxon p = %.4g\n",
            suppressWarnings(wilcox.test(
              nc_burden$alpha_per_mm2[nc_burden$Diagnosis == "T1D"],
              nc_burden$alpha_per_mm2[nc_burden$Diagnosis == "ND"])$p.value)))
cat("Agreement with Test 3 means the conclusion is not an artefact of the\n")
cat("size threshold.\n")
```

```{r save}
write_csv(ineq, "selection_density_inequality.csv")
write_csv(burden, "selection_burden_per_donor.csv")
cat("\nSaved selection_density_inequality.csv and selection_burden_per_donor.csv\n")
```
