---
title: "Ki-67 / CD3 / CD45 Marker Dataset Pipeline"
output: html_document
date: "`r Sys.Date()`"
---

Builds the six marker datasets consumed by `marker_count_models_v4.Rmd`:

| Output | Description |
|---|---|
| `ki67_data.csv` / `ki67_data_t1d.csv` | Ki-67+ counts per islet, with demographics |
| `cd3_data.csv` / `cd3_data_t1d.csv` | CD3+ counts per islet |
| `cd45_data.csv` / `cd45_data_t1d.csv` | CD45+ counts per islet |

Split out of `Create_unified_dataset_v8.Rmd` Parts D/E/F. Those were three
near-identical copies of the same 200-line block; here it is one function
called three times, which is most of the reduction from ~1100 lines to ~400.

**Inputs.** Run `Create_count_datasets_v1.Rmd` first. This file reads two of
its outputs and never re-reads the count CSVs:

- `Data/no_cutoff_data.rds` — demographics, `Islet.Cells`, `Count.Acinar`,
  `Stain`, `log_Islet.Cells`, with every donor-level fix already applied
- `Data/phenotyper_geometry.rds` — bounding boxes per
  `(source_file, Object.Id)`

Plus the three marker CSV folders (paths in the setup chunk).

## Structural change from Parts D/E/F

v8 de-duplicated the phenotyper data to physical islets for *matching* (D2),
then joined marker values back onto the **full** phenotyper table (D5), which
re-expanded every physical islet to one row per classifier run. A single islet
therefore contributed several rows carrying an identical outcome value but
different `Islet.Cells` — a Single-stain run only detects insulin, so its
`Islet.Cells` is not comparable to a Quad run's. Nothing in the model formulas
absorbed that.

Here the collapse happens once, at the top, and stays collapsed: one row per
physical islet throughout, keeping the classifier run that detects the most
endocrine cell types (Quad > Double > Single). Matching still runs against
*all* physical islets, including ones Part A will later drop, so a marker
object cannot be mis-assigned to a neighbouring islet just because its true
partner was excluded.

## Other changes carried over from the v8 notes

1. `total_cells` is the sum of the three phenotyper columns
   (`insulin_cells + glucagon_cells + acinar_cells`), **not** `Total.Cells` —
   direct instruction from the HALO operator to ignore `Total.Cells`/`Hema`
   regardless of what a given image's classifier settings say. `Total.Cells` is
   kept as `total_cells_raw` for reference. There is consequently no donor
   exclusion for 6413/6417: those looked anomalous only under `Total.Cells`.
2. Classifier-name variants are coalesced by regex, not literal string.
   `read.csv()` mangles headers through `make.names()`, so
   `"Ki-67 Phenotyper - Insulin Cells"` arrives as
   `"Ki.67.Phenotyper...Insulin.Cells"` — every space *and* hyphen becomes its
   own `.`. An earlier version used literal suffixes, matched nothing, and
   silently turned every hormone-composition column into `NA`. There is now a
   hard stop if any component comes back mostly missing.
3. The marker-count ceiling is `total_cells <= Islet.Cells + Count.Acinar`.
   v8 used `total_cells <= exp(log_Islet.Cells)`, i.e. `<= Islet.Cells` — but
   `Islet.Cells = Count.ins + Count.glu + Count.soma + Count.PP` excludes
   acinar cells while `total_cells` includes marker-positive acinar cells, so
   v8 was comparing a numerator against a denominator that omits part of the
   numerator. It discarded legitimate rows wherever an islet carried
   peri-islet acinar signal, preferentially at small islet sizes.
4. `Islet.Cells >= 15`, matching `full_data` in the count pipeline. v8's marker
   datasets applied only the upper bound, so they were the one place pooling
   objects with 1–14 endocrine cells in with real islets.
5. The HALO operator flagged ~10 specific 6417 image IDs as genuinely
   quad-stained (a physical-stain problem, not a settings mix-up). Not
   implemented as a filter because none were found in the export this was
   checked against — the setup chunk lists them and the run reports whether any
   are present, rather than leaving that as a note to remember.

**Percentages are deliberately not computed here.** `glu_prolif_rate`,
`ins_prolif_rate` and the `*_impossible` flags are gone. The inputs they need
are all preserved in the output (`Count.ins`, `Count.glu`, the marker
`*_cells` counts and the `pct_*_positive` columns), so they can be derived
later without rebuilding anything.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
# janitor is not needed: v8 called clean_names() on the joined frame, but here
# every column is selected and renamed explicitly.

dir.create("Data", showWarnings = FALSE)

# ---- inputs ---------------------------------------------------------------
NO_CUTOFF_RDS <- "Data/no_cutoff_data.rds"
GEOMETRY_RDS  <- "Data/phenotyper_geometry.rds"

# Check these first if anything fails - they are absolute, unlike v8's
# relative "ki67_data" which assumed the working directory was F:/Work/nPod_final_v4.
MARKER_PATHS <- list(
  ki67 = "F:/Work/nPod_final_v4/ki67_data",
  cd3  = "F:/Work/nPod_final_v4/cd3_data",
  cd45 = "F:/Work/nPod_final_v4/cd45_data"
)

# ---- parameters -----------------------------------------------------------
MIN_ISLET_CELLS  <- 15    # lower size bound; 1 reproduces v8
LOG_ISLET_MAX    <- 5.5   # upper bound: one islet, not two merged by segmentation
MATCH_RADIUS     <- 20    # matcher search radius, px
MATCH_MAX_DIST   <- 15    # kept after matching
MATCH_MIN_AREA   <- 0.5   # min bounding-box area ratio between the two objects
STAIN_PREFERENCE <- c("Quad", "Double", "Single")

# All three marker exports name their classifier columns as Ki-67 Phenotyper
# variants, including the CD3 and CD45 ones - that is the operator's naming, not
# a copy-paste error carried over from Part D. The matched columns are printed
# per marker so a naming change shows up instead of silently matching nothing.
MARKER_COL_PATTERN <- "Ki\\.?_?67"

# Flagged by the HALO operator as physically quad-stained and to be excluded if
# present. Reported rather than silently assumed absent.
QUAD_STAINED_IMAGES <- character(0)   # e.g. c("191012", "191015")
```

# Helpers

```{r helpers}
# Coalesce every column whose name matches `pattern` into one, taking whichever
# value is non-NA. Each row only has a real value in whichever classifier
# variant produced it, so there is no ambiguity about which wins.
#
# `pattern` must match make.names()-mangled headers: pass
# "Insulin[^A-Za-z]*Cells$" (any run of non-letters between the words), never a
# literal "Insulin Cells".
coalesce_by_pattern <- function(data, pattern) {
  cols <- data[grepl(pattern, names(data))]
  if (length(cols) == 0) return(rep(NA_real_, nrow(data)))
  Reduce(dplyr::coalesce, cols)
}

# Row-count ledger. dplyr::filter() drops NA rows silently, which matters here:
# total_cells is a sum of coalesced columns, so one missing component makes it
# NA and a ceiling filter would then remove that row without saying so.
filter_log <- function(df, cond, label) {
  keep  <- rlang::eval_tidy(rlang::enquo(cond), df)
  n_na  <- sum(is.na(keep))
  keep2 <- !is.na(keep) & keep
  cat(sprintf("  %-46s %8d -> %8d  (%5.2f%% dropped%s)\n",
              label, nrow(df), sum(keep2),
              100 * (1 - sum(keep2) / max(nrow(df), 1)),
              if (n_na > 0) sprintf(", %d NA", n_na) else ""))
  df[keep2, , drop = FALSE]
}

# Greedy nearest-centroid matcher, 1-to-1 within an image. Unchanged in
# behaviour from D3; it just returns keys instead of row indices.
match_one_image <- function(marker, islets, radius = MATCH_RADIUS) {
  m <- marker %>% mutate(cx = (XMin + XMax) / 2, cy = (YMin + YMax) / 2,
                         .i = row_number())
  p <- islets %>% mutate(cx = (XMin + XMax) / 2, cy = (YMin + YMax) / 2,
                         .j = row_number())

  pairs <- expand.grid(.i = m$.i, .j = p$.j) %>%
    mutate(dist = sqrt((m$cx[.i] - p$cx[.j])^2 + (m$cy[.i] - p$cy[.j])^2)) %>%
    filter(dist <= radius) %>%
    arrange(dist)
  if (nrow(pairs) == 0) return(NULL)

  used_m <- logical(nrow(m)); used_p <- logical(nrow(p))
  keep   <- logical(nrow(pairs))
  for (k in seq_len(nrow(pairs))) {
    if (!used_m[pairs$.i[k]] && !used_p[pairs$.j[k]]) {
      keep[k] <- TRUE
      used_m[pairs$.i[k]] <- TRUE
      used_p[pairs$.j[k]] <- TRUE
    }
  }
  pairs <- pairs[keep, , drop = FALSE]

  a_m <- (m$XMax[pairs$.i] - m$XMin[pairs$.i]) * (m$YMax[pairs$.i] - m$YMin[pairs$.i])
  a_p <- (p$XMax[pairs$.j] - p$XMin[pairs$.j]) * (p$YMax[pairs$.j] - p$YMin[pairs$.j])

  tibble(marker_uid = m$marker_uid[pairs$.i],
         islet_uid  = p$islet_uid[pairs$.j],
         dist_px    = pairs$dist,
         area_ratio = pmin(a_m, a_p) / pmax(a_m, a_p))
}

# Read every CSV under a folder, tolerating per-file failures.
read_halo_folder <- function(path, label) {
  if (!dir.exists(path)) stop(label, ": folder not found - ", path)
  files <- list.files(path, pattern = "\\.csv$", recursive = TRUE,
                      full.names = TRUE)
  files <- files[!grepl("Total_Summary_Results\\.csv$", files)]
  cat(sprintf("  %s: %d CSV files\n", label, length(files)))
  if (length(files) == 0) stop(label, ": no CSVs found under ", path)

  out <- lapply(files, function(f) {
    tryCatch({
      df <- read.csv(f, stringsAsFactors = FALSE)
      df$source_file <- basename(f)
      df$folder_id   <- basename(dirname(dirname(dirname(dirname(f)))))
      df
    }, error = function(e) {
      cat("  ERROR reading", f, ":", conditionMessage(e), "\n"); NULL
    })
  })
  bind_rows(out) %>%
    mutate(
      folder_id = case_when(
        folder_id == "030" ~ "HPAP030", folder_id == "041" ~ "HPAP041",
        folder_id == "048" ~ "HPAP048", folder_id == "068" ~ "HPAP068",
        folder_id == "078" ~ "HPAP078", TRUE ~ folder_id
      ),
      ImageID = str_extract(source_file, "\\d+")
    )
}
```

# Step 1: phenotyper islets

```{r phenotyper-islets}
for (f in c(NO_CUTOFF_RDS, GEOMETRY_RDS)) {
  if (!file.exists(f)) {
    stop("missing ", f, " - run Create_count_datasets_v1.Rmd first")
  }
}

no_cutoff_data     <- readRDS(NO_CUTOFF_RDS)
phenotyper_geometry <- readRDS(GEOMETRY_RDS)

# ---- demographics, keyed on the one unique phenotyper key -----------------
# (source_file, Object.Id) is the key that uniquely identifies a phenotyper
# row. v8 joined on (Donor, ImageID, Object.Id), which is NOT unique: Object.Id
# is assigned per analysis run, so the Single/Double/Quad runs of one image each
# contain an Object.Id 1, 2, 3, ... referring to different physical objects. A
# left_join on a non-unique key silently multiplies rows and can attach another
# run's Islet.Cells / Count.ins / Count.glu to a row. Part A already uses this
# key to join location_lookup.
required_ncd <- c("source_file", "Object.Id", "Donor", "ImageID", "Region",
                  "Sex", "Age", "AgeCat", "Diagnosis", "Disease.Duration",
                  "age_at_onset", "Stain", "Count.ins", "Count.glu",
                  "Count.Acinar", "Islet.Cells", "log_Islet.Cells")
missing_ncd <- setdiff(required_ncd, names(no_cutoff_data))
if (length(missing_ncd) > 0) {
  stop("no_cutoff_data.rds is missing: ", paste(missing_ncd, collapse = ", "))
}

demo <- no_cutoff_data %>%
  mutate(source_file = as.character(source_file),
         Object.Id   = as.integer(Object.Id),
         Donor       = as.character(Donor),
         ImageID_ncd = as.character(ImageID),
         Region      = as.character(Region),
         Stain       = as.character(Stain)) %>%
  dplyr::select(source_file, Object.Id, Donor, ImageID_ncd, Region, Sex, Age,
                AgeCat, Diagnosis, Disease.Duration, age_at_onset, Stain,
                Count.ins, Count.glu, Count.Acinar, Islet.Cells,
                log_Islet.Cells) %>%
  distinct()

dup_demo <- demo %>% count(source_file, Object.Id) %>% filter(n > 1)
cat("demo rows:", nrow(demo),
    "| duplicate (source_file, Object.Id) keys:", nrow(dup_demo),
    "(must be 0)\n")
if (nrow(dup_demo) > 0) {
  print(head(dup_demo, 20))
  stop("(source_file, Object.Id) is not unique in no_cutoff_data")
}

# How bad the v8 key was, for the record.
v8_dup <- no_cutoff_data %>%
  mutate(Donor = as.character(Donor), ImageID = as.character(ImageID),
         Object.Id = as.integer(Object.Id)) %>%
  distinct(Donor, ImageID, Object.Id, Islet.Cells, Count.ins, Count.glu, Region) %>%
  count(Donor, ImageID, Object.Id) %>%
  filter(n > 1)
cat("v8 key (Donor, ImageID, Object.Id): ", nrow(v8_dup), " duplicated keys",
    if (nrow(v8_dup) > 0) " <- v8 was multiplying rows here" else
      " (v8 key happened to be safe on this data)", "\n", sep = "")

# ---- collapse to one row per physical islet ------------------------------
# Stain is constant within a source_file (one algorithm per export, plus A5's
# per-image fix), so it can be looked up per file and used to rank the runs.
# Files absent from no_cutoff_data get NA and rank last; their objects are
# still available as match candidates, which is the point.
stain_by_file <- demo %>% distinct(source_file, Stain)
if (any(duplicated(stain_by_file$source_file))) {
  print(stain_by_file %>%
          filter(source_file %in% source_file[duplicated(source_file)]))
  stop("Stain is not constant within a source_file - the classifier-run ",
       "ranking used to pick one row per physical islet would be arbitrary.")
}

geom <- phenotyper_geometry %>%
  mutate(source_file = as.character(source_file),
         ImageID     = as.character(ImageID),
         Object.Id   = as.integer(Object.Id),
         islet_uid   = paste(ImageID, XMin, XMax, YMin, YMax, sep = "_")) %>%
  left_join(stain_by_file, by = "source_file") %>%
  mutate(stain_rank = match(Stain, STAIN_PREFERENCE),
         stain_rank = ifelse(is.na(stain_rank),
                             length(STAIN_PREFERENCE) + 1L, stain_rank))

phenotyper_islets <- geom %>%
  arrange(islet_uid, stain_rank, Object.Id) %>%
  distinct(islet_uid, .keep_all = TRUE)

cat(sprintf("\nphenotyper rows (all classifier runs): %d\n", nrow(geom)))
cat(sprintf("distinct physical islets:             %d  (x%.2f collapse)\n",
            nrow(phenotyper_islets),
            nrow(geom) / max(nrow(phenotyper_islets), 1)))
cat("classifier run retained per islet:\n")
print(table(phenotyper_islets$Stain, useNA = "ifany"))

# Matching is done per ImageID alone, so an ImageID shared across donors would
# let a marker object match another donor's islet.
img_donor <- geom %>% distinct(ImageID, folder_id) %>% count(ImageID) %>% filter(n > 1)
cat("\nImageIDs appearing under more than one donor folder:", nrow(img_donor),
    "(must be 0)\n")
if (nrow(img_donor) > 0) print(head(img_donor, 20))

if (length(QUAD_STAINED_IMAGES) > 0) {
  present <- intersect(QUAD_STAINED_IMAGES, unique(geom$ImageID))
  cat("operator-flagged quad-stained images present:", length(present), "/",
      length(QUAD_STAINED_IMAGES), "\n")
  if (length(present) > 0) {
    cat("  excluding:", paste(present, collapse = ", "), "\n")
    phenotyper_islets <- phenotyper_islets %>% filter(!ImageID %in% present)
  }
}

rm(phenotyper_geometry, geom, no_cutoff_data); invisible(gc())
```

# Step 2: build one marker dataset

```{r build-function}
# Pull the marker classifier's own columns and collapse the name variants.
extract_marker_counts <- function(raw, label) {
  if (!"Total.Cells" %in% names(raw)) {
    stop(label, ": no Total.Cells column in the marker export")
  }
  # v8 hard-coded `Tissue.Area..um.2.`, which is a latent hard failure:
  # read.csv() pushes "Tissue Area (um^2)" through make.names(), and whether
  # the micro sign survives as "um", ".m" or something else depends on the
  # file encoding and the locale. Match it instead.
  tissue_col <- grep("^Tissue\\.Area", names(raw), value = TRUE)
  if (length(tissue_col) == 0) {
    cat(sprintf("  NOTE: no Tissue.Area column found in %s - tissue_area_um_2 will be NA\n",
                label))
    raw$.tissue_area <- NA_real_
  } else {
    if (length(tissue_col) > 1) {
      cat("  NOTE: several Tissue.Area columns, using", tissue_col[1], "\n")
    }
    raw$.tissue_area <- suppressWarnings(as.numeric(raw[[tissue_col[1]]]))
  }

  pre <- raw %>%
    mutate(marker_uid = paste(ImageID, XMin, XMax, YMin, YMax, sep = "_")) %>%
    dplyr::select(
      ImageID, marker_uid, XMin, XMax, YMin, YMax,
      total_cells_raw  = Total.Cells,
      tissue_area_um_2 = .tissue_area,
      matches(MARKER_COL_PATTERN, ignore.case = TRUE)
    )

  fixed <- c("ImageID", "marker_uid", "XMin", "XMax", "YMin", "YMax",
             "total_cells_raw", "tissue_area_um_2")
  matched <- setdiff(names(pre), fixed)
  cat(sprintf("  classifier columns matched by MARKER_COL_PATTERN: %d\n",
              length(matched)))
  if (length(matched) == 0) {
    stop(label, ": MARKER_COL_PATTERN matched no columns - the classifier ",
         "naming has changed. Available columns:\n  ",
         paste(head(names(raw), 40), collapse = ", "))
  }
  print(matched)

  out <- pre %>%
    transmute(
      ImageID, marker_uid, XMin, XMax, YMin, YMax,
      total_cells_raw, tissue_area_um_2,
      insulin_cells         = coalesce_by_pattern(pre, "Insulin[^A-Za-z]*Cells$"),
      glucagon_cells        = coalesce_by_pattern(pre, "Glucagon[^A-Za-z]*Cells$"),
      acinar_cells          = coalesce_by_pattern(pre, "Acinar[^A-Za-z]*Cells$"),
      pct_insulin_positive  = coalesce_by_pattern(pre, "Insulin[^A-Za-z]*Positive[^A-Za-z]*Cells$"),
      pct_glucagon_positive = coalesce_by_pattern(pre, "Glucagon[^A-Za-z]*Positive[^A-Za-z]*Cells$"),
      pct_acinar_positive   = coalesce_by_pattern(pre, "Acinar[^A-Za-z]*Positive[^A-Za-z]*Cells$")
    )

  # A component that is almost entirely NA means the regex matched nothing
  # useful, not that the marker is sparse. That failure mode silently turned
  # every hormone-composition column into NA in an earlier version, so it is a
  # hard stop rather than a warning.
  comps <- c("insulin_cells", "glucagon_cells", "acinar_cells")
  na_rate <- vapply(comps, function(cc) mean(is.na(out[[cc]])), numeric(1))
  cat("  NA rate per component: ",
      paste(sprintf("%s=%.1f%%", names(na_rate), 100 * na_rate),
            collapse = ", "), "\n")
  if (any(na_rate > 0.5)) {
    stop(label, ": ", paste(names(na_rate)[na_rate > 0.5], collapse = ", "),
         " more than 50% missing - name matching has failed, not sparse data.")
  }

  all_na <- is.na(out$insulin_cells) & is.na(out$glucagon_cells) &
    is.na(out$acinar_cells)
  cat(sprintf("  rows missing all three components: %d (%.2f%%) - dropped\n",
              sum(all_na), 100 * mean(all_na)))

  out %>%
    filter(!all_na) %>%
    mutate(
      across(all_of(comps), ~ coalesce(.x, 0)),
      # total_cells is this sum, NOT Total.Cells - operator instruction to
      # ignore Total.Cells/Hema regardless of classifier settings.
      total_cells      = insulin_cells + glucagon_cells + acinar_cells,
      # Endocrine-only variant, for the sensitivity analysis on whether
      # marker+ acinar cells belong in the outcome. Defensible for CD3/CD45
      # (peri-islet infiltration is part of insulitis); for Ki-67 it mixes
      # endocrine and exocrine proliferation, which differ in baseline rate.
      total_cells_endo = insulin_cells + glucagon_cells
    )
}

build_marker_dataset <- function(label, path) {
  cat(sprintf("\n%s\n=== %s ===\n", strrep("=", 70), label))

  # ---- read ---------------------------------------------------------------
  raw <- read_halo_folder(path, label)
  cat(sprintf("  raw rows: %d, donors: %d, images: %d\n", nrow(raw),
              dplyr::n_distinct(raw$folder_id), dplyr::n_distinct(raw$ImageID)))

  # ---- marker counts ------------------------------------------------------
  mk <- extract_marker_counts(raw, label)
  if (any(duplicated(mk$marker_uid))) {
    n_dup <- sum(duplicated(mk$marker_uid))
    cat(sprintf("  WARNING: %d duplicate marker bounding boxes - keeping first\n",
                n_dup))
    mk <- mk[!duplicated(mk$marker_uid), , drop = FALSE]
  }
  rm(raw); invisible(gc())

  # ---- spatial match ------------------------------------------------------
  imgs <- intersect(unique(mk$ImageID), unique(phenotyper_islets$ImageID))
  cat(sprintf("  images present in both marker and phenotyper data: %d\n",
              length(imgs)))
  matches <- map_dfr(imgs, function(img) {
    d1 <- mk[mk$ImageID == img, , drop = FALSE]
    d2 <- phenotyper_islets[phenotyper_islets$ImageID == img, , drop = FALSE]
    if (nrow(d1) == 0 || nrow(d2) == 0) return(NULL)
    match_one_image(d1, d2)
  })
  cat(sprintf("  raw matches: %d\n", nrow(matches)))

  # True matches are both spatially close and similar in size.
  matches <- matches %>%
    filter(dist_px <= MATCH_MAX_DIST, area_ratio >= MATCH_MIN_AREA)
  cat(sprintf("  matches kept (dist <= %d, area ratio >= %.2f): %d (%.1f%% of marker objects)\n",
              MATCH_MAX_DIST, MATCH_MIN_AREA, nrow(matches),
              100 * nrow(matches) / max(nrow(mk), 1)))
  cat("  distance distribution of kept matches:\n")
  print(summary(matches$dist_px))

  # ---- assemble -----------------------------------------------------------
  # One row per matched physical islet, and it stays that way: the marker
  # values attach to the single canonical phenotyper row, not to every
  # classifier run of it.
  df <- matches %>%
    left_join(mk, by = "marker_uid", relationship = "one-to-one") %>%
    left_join(
      phenotyper_islets %>%
        dplyr::select(islet_uid, source_file, Object.Id, ImageID_ph = ImageID,
                      ph_XMin = XMin, ph_XMax = XMax,
                      ph_YMin = YMin, ph_YMax = YMax),
      by = "islet_uid", relationship = "one-to-one"
    ) %>%
    # relationship = "many-to-one" makes a silent row multiplication
    # impossible: dplyr errors instead of quietly duplicating.
    left_join(demo, by = c("source_file", "Object.Id"),
              relationship = "many-to-one")

  stopifnot(nrow(df) == nrow(matches))

  # ---- filters ------------------------------------------------------------
  cat("  attrition:\n")
  df <- df %>%
    filter_log(!is.na(Diagnosis),
               "matched to a real islet with demographics") %>%
    filter_log(log_Islet.Cells <= LOG_ISLET_MAX,
               "log_Islet.Cells <= LOG_ISLET_MAX") %>%
    # Lower size bound, absent from v8. The count pipeline applies
    # Islet.Cells >= 15 for full_data and treats 2-14 endocrine cells as a
    # separate small-cluster class, so v8's marker datasets were the only place
    # pooling scattered single endocrine cells with real islets. Those objects
    # are where the outcome is structurally near-zero, and they dominated both
    # the row count and the leverage on the islet-size slope.
    filter_log(Islet.Cells >= MIN_ISLET_CELLS,
               "Islet.Cells >= MIN_ISLET_CELLS") %>%
    # Marker+ count cannot exceed the object's cell count. The denominator has
    # to include acinar cells because the numerator does - see the header note.
    filter_log(total_cells <= Islet.Cells + Count.Acinar,
               "total_cells <= Islet.Cells + Count.Acinar")

  # ---- consistency and completeness --------------------------------------
  # The demographics join keys on (source_file, Object.Id) and never mentions
  # the image, so confirm the three independent ImageID sources agree.
  stopifnot(all(df$ImageID == df$ImageID_ph),
            all(df$ImageID == df$ImageID_ncd))

  na_demo <- vapply(c("Region", "Sex", "Age", "Islet.Cells", "log_Islet.Cells",
                      "Count.Acinar"),
                    function(v) sum(is.na(df[[v]])), integer(1))
  if (any(na_demo > 0)) {
    cat("  WARNING: NA demographics after the join (expected 0):\n")
    print(na_demo[na_demo > 0])
  }

  # ---- centring -----------------------------------------------------------
  # After every filter, so the means are those of the islets that actually end
  # up in the dataset.
  df <- df %>%
    mutate(Age_c             = Age - mean(Age, na.rm = TRUE),
           log_Islet.Cells_c = log_Islet.Cells -
                               mean(log_Islet.Cells, na.rm = TRUE)) %>%
    dplyr::select(
      Donor, ImageID, source_file, Object.Id, islet_uid, marker_uid,
      Region, Sex, Age, Age_c, AgeCat, Diagnosis, Disease.Duration, age_at_onset,
      Stain, Islet.Cells, log_Islet.Cells, log_Islet.Cells_c,
      Count.ins, Count.glu, Count.Acinar,
      total_cells, total_cells_endo, total_cells_raw,
      insulin_cells, glucagon_cells, acinar_cells,
      pct_insulin_positive, pct_glucagon_positive, pct_acinar_positive,
      tissue_area_um_2, dist_px, area_ratio,
      ph_XMin, ph_XMax, ph_YMin, ph_YMax
    )

  if (nrow(df) == 0) stop(label, ": no rows survived the filters")
  df
}
```

# Step 3: build all three

```{r build-all}
marker_data <- list()
for (tag in names(MARKER_PATHS)) {
  marker_data[[tag]] <- build_marker_dataset(tag, MARKER_PATHS[[tag]])
}
```

# Step 4: diagnostics

```{r diagnostics}
marker_diagnostics <- function(df, label) {
  cat(sprintf("\n%s\n=== %s: diagnostics ===\n", strrep("=", 70), label))

  cat("\n-- replication structure --\n")
  cat(sprintf("  rows %d, physical islets %d, rows per islet %.4f (must be 1.0000)\n",
              nrow(df), dplyr::n_distinct(df$islet_uid),
              nrow(df) / max(dplyr::n_distinct(df$islet_uid), 1)))
  cat("  classifier run retained:\n")
  print(table(df$Stain, useNA = "ifany"))

  cat("\n-- islet size --\n")
  print(summary(df$Islet.Cells))
  cat(sprintf("  below 15 endocrine cells: %d (must be 0)\n",
              sum(df$Islet.Cells < MIN_ISLET_CELLS)))

  cat("\n-- outcome --\n")
  cat(sprintf("  total_cells      mean %.3f, zeros %.1f%%, max %d\n",
              mean(df$total_cells), 100 * mean(df$total_cells == 0),
              as.integer(max(df$total_cells))))
  cat(sprintf("  total_cells_endo mean %.3f, zeros %.1f%%, max %d\n",
              mean(df$total_cells_endo), 100 * mean(df$total_cells_endo == 0),
              as.integer(max(df$total_cells_endo))))

  cat("\n-- total_cells vs total_cells_raw, per donor --\n")
  cat("  The v8 expectation was 'most donors ~0 difference'. Pooled means were\n")
  cat("  0.29 vs 0.57 (Ki-67) and 0.17 vs 0.57 (CD3) - a 49% and 70% gap, with\n")
  cat("  both markers landing on an identical 0.57. Largest 10 gaps:\n")
  per_donor <- df %>%
    group_by(Donor) %>%
    summarise(n = n(),
              mean_tc  = mean(total_cells),
              mean_raw = mean(total_cells_raw, na.rm = TRUE),
              gap      = mean_raw - mean_tc,
              .groups  = "drop") %>%
    arrange(desc(abs(gap)))
  print(as.data.frame(head(per_donor, 10)), digits = 3)
  cat(sprintf("  donors with |gap| < 0.01: %d / %d\n",
              sum(abs(per_donor$gap) < 0.01, na.rm = TRUE), nrow(per_donor)))
  # If total_cells_raw still tracks the phenotyper's own cell count, the
  # column collision the v7 notes describe is not actually resolved.
  cat(sprintf("  cor(total_cells_raw, Islet.Cells) = %.3f (near 1 => still leaking)\n",
              suppressWarnings(cor(df$total_cells_raw, df$Islet.Cells,
                                   use = "complete.obs"))))
  cat(sprintf("  cor(total_cells,     Islet.Cells) = %.3f\n",
              suppressWarnings(cor(df$total_cells, df$Islet.Cells,
                                   use = "complete.obs"))))

  cat("\n-- ceiling filter: v8 denominator vs corrected --\n")
  cat(sprintf("  would fail v8 (<= Islet.Cells):                %d (%.2f%%)\n",
              sum(df$total_cells > df$Islet.Cells),
              100 * mean(df$total_cells > df$Islet.Cells)))
  cat(sprintf("  fails corrected (<= Islet.Cells + Count.Acinar): %d (must be 0)\n",
              sum(df$total_cells > df$Islet.Cells + df$Count.Acinar)))
  cat("  The gap is legitimate rows v8 discarded for carrying acinar signal.\n")

  cat("\n-- design --\n")
  cat(sprintf("  %d donors (ND=%d, T1D=%d), %d images\n",
              dplyr::n_distinct(df$Donor),
              dplyr::n_distinct(df$Donor[df$Diagnosis == "ND"]),
              dplyr::n_distinct(df$Donor[df$Diagnosis == "T1D"]),
              dplyr::n_distinct(df$ImageID)))
  cat("  donor-level Sex x Diagnosis:\n")
  print(table(distinct(df, Donor, Diagnosis, Sex)[c("Diagnosis", "Sex")]))
  cat("  images per region:\n")
  print(table(distinct(df, ImageID, Region)$Region))

  invisible(per_donor)
}

for (tag in names(marker_data)) marker_diagnostics(marker_data[[tag]], tag)
```

# Step 5: save

```{r save}
# T1D subsets are re-centred on the T1D islets, so their Age_c /
# log_Islet.Cells_c are not comparable with the full datasets' - intended, but
# worth remembering when comparing coefficients across the two model families.
#
# Duration models use Disease.Duration_c + age_at_onset_c and never Age_c:
# current age = age at onset + disease duration, so all three together are
# rank-deficient.
make_t1d <- function(df) {
  df %>%
    filter(Diagnosis == "T1D") %>%
    mutate(
      Age_c              = Age - mean(Age, na.rm = TRUE),
      log_Islet.Cells_c  = log_Islet.Cells - mean(log_Islet.Cells, na.rm = TRUE),
      Disease.Duration_c = Disease.Duration - mean(Disease.Duration, na.rm = TRUE),
      age_at_onset       = Age - Disease.Duration,
      age_at_onset_c     = age_at_onset - mean(age_at_onset, na.rm = TRUE)
    )
}

for (tag in names(marker_data)) {
  df  <- marker_data[[tag]]
  t1d <- make_t1d(df)

  write_csv(df,  sprintf("Data/%s_data.csv", tag))
  saveRDS(df,    sprintf("Data/%s_data.rds", tag))
  write_csv(t1d, sprintf("Data/%s_data_t1d.csv", tag))
  saveRDS(t1d,   sprintf("Data/%s_data_t1d.rds", tag))

  cat(sprintf("%-5s  %7d rows / %3d donors / %4d images   t1d: %7d rows / %3d donors\n",
              tag, nrow(df), dplyr::n_distinct(df$Donor),
              dplyr::n_distinct(df$ImageID),
              nrow(t1d), dplyr::n_distinct(t1d$Donor)))
}
```

```{r handoff-check}
# marker_count_models_v4.Rmd checks for exactly these columns and stops if any
# are absent, so verify here rather than discovering it there.
REQUIRED_BY_MODELS <- c("total_cells", "total_cells_endo", "total_cells_raw",
                        "Islet.Cells", "Count.Acinar", "Stain", "islet_uid",
                        "log_Islet.Cells", "log_Islet.Cells_c",
                        "Donor", "ImageID", "Region", "Sex", "Age", "Diagnosis")
REQUIRED_T1D <- c(REQUIRED_BY_MODELS, "Disease.Duration_c", "age_at_onset_c")

ok <- TRUE
for (tag in names(marker_data)) {
  for (suffix in c("", "_t1d")) {
    f <- sprintf("Data/%s_data%s.csv", tag, suffix)
    need <- if (suffix == "") REQUIRED_BY_MODELS else REQUIRED_T1D
    nms <- names(readr::read_csv(f, n_max = 0, show_col_types = FALSE))
    miss <- setdiff(need, nms)
    cat(sprintf("%-26s %s\n", basename(f),
                if (length(miss) == 0) "OK" else
                  paste("MISSING:", paste(miss, collapse = ", "))))
    if (length(miss) > 0) ok <- FALSE
  }
}
if (!ok) stop("output is missing columns marker_count_models_v4.Rmd requires")
cat("\nNext: marker_count_models_v4.Rmd\n")
```

```{r session-info}
sessionInfo()
```
