---
title: "Rate models in original units"
output: html_document
date: "`r Sys.Date()`"
---

Turns the six rate models into numbers you can put in a sentence:

- **% of cells positive** — the rate itself, per 100 cells of the relevant type
- **marker+ cells per islet** — the same fit converted back to counts
- **observed values alongside**, computed on each model's own subset

Run after `rate_models.Rmd` in the same session.

**This also fixes a bug in that file.** Its `crude_fold_rate` column was computed
on the full dataset while the models were fitted to the restricted subsets. For
the acinar bins that is a 35% difference in islets, and the crude numerator was
counting marker-positive cells from islets where the denominator was zero —
cells with no denominator at all, which is not a rate. Everything below uses the
model's own subset.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(glmmTMB)

PER <- 100          # report per 100 cells, i.e. as a percentage
MARKERS <- c("ki67", "cd3")
BINS <- tibble(
  bin = c("ins", "glu", "aci"),
  num = c("insulin_cells", "glucagon_cells", "acinar_cells"),
  den = c("Count.ins", "Count.glu", "Count.Acinar"),
  cell = c("beta", "alpha", "acinar")
)
```

```{r collect}
# Rebuild the subsets if this is a fresh session; reuse them if not.
if (!exists("subsets")) {
  subsets <- list()
  for (tag in MARKERS) {
    d <- get(paste0(tag, "_data"))
    for (i in seq_len(nrow(BINS))) {
      keep <- !is.na(d[[BINS$den[i]]]) & d[[BINS$den[i]]] > 0
      subsets[[paste0(tag, "_", BINS$bin[i])]] <- droplevels(d[keep, , drop = FALSE])
    }
  }
}

spec <- bind_rows(lapply(MARKERS, function(tag)
  BINS %>% mutate(marker = tag, key = paste0(tag, "_", bin),
                  model = paste0(key, "_rate"))))

missing <- spec$model[!vapply(spec$model, exists, logical(1))]
if (length(missing)) stop("model objects not found: ", paste(missing, collapse = ", "),
                          "\n  run rate_models.Rmd first")
```

# Model-predicted rates

```{r predicted-rates}
# The offset is offset(log(denominator)). Setting the denominator to 100 in the
# prediction grid makes the offset log(100), so predict(type = "response")
# returns the expected number of marker+ cells PER 100 CELLS - i.e. a percentage
# directly. type = "response" also includes the zero-inflation component, so
# this is the marginal rate, not the conditional one.
predict_rate <- function(m, sub, den, per = PER, by_region = FALSE) {
  vars <- all.vars(formula(m))
  resp <- vars[1]

  grid <- expand.grid(
    Diagnosis = levels(sub$Diagnosis),
    Region = if (by_region) levels(sub$Region) else levels(sub$Region)[1],
    stringsAsFactors = FALSE, KEEP.OUT.ATTRS = FALSE)

  grid$Sex   <- levels(sub$Sex)[1]     # Female
  grid$Age_c <- 0                       # sample mean
  if ("log_Islet.Cells_c" %in% vars) grid$log_Islet.Cells_c <- 0
  grid[[den]]  <- per                   # <- makes the prediction a rate per `per`
  grid[[resp]] <- 0                     # placeholder; unused for prediction
  grid$Donor   <- levels(sub$Donor)[1]  # placeholder; re.form = NA ignores it
  grid$ImageID <- levels(sub$ImageID)[1]
  for (v in c("Diagnosis", "Region", "Sex")) {
    grid[[v]] <- factor(grid[[v]], levels = levels(sub[[v]]))
  }

  pr  <- predict(m, newdata = grid, type = "response", se.fit = TRUE,
                 re.form = NA, allow.new.levels = TRUE)
  fit <- as.numeric(pr$fit); se <- as.numeric(pr$se.fit)
  rel <- se / fit   # CI formed on the log scale so it cannot go negative

  tibble(Diagnosis = as.character(grid$Diagnosis),
         Region    = as.character(grid$Region),
         rate_pct  = fit,
         lwr_pct   = fit * exp(-1.96 * rel),
         upr_pct   = fit * exp(1.96 * rel))
}

# Observed rate on the SAME subset the model used.
observed_rate <- function(sub, num, den, per = PER) {
  sub %>%
    group_by(Diagnosis) %>%
    summarise(islets = n(), donors = dplyr::n_distinct(Donor),
              obs_rate_pct = per * sum(.data[[num]]) / sum(.data[[den]]),
              obs_cells_per_islet = mean(.data[[num]]),
              mean_denominator = mean(.data[[den]]),
              .groups = "drop") %>%
    mutate(Diagnosis = as.character(Diagnosis))
}

units_tab <- bind_rows(lapply(seq_len(nrow(spec)), function(k) {
  s <- spec[k, ]
  m <- get(s$model); sub <- subsets[[s$key]]
  predict_rate(m, sub, s$den) %>%
    left_join(observed_rate(sub, s$num, s$den), by = "Diagnosis") %>%
    mutate(marker = s$marker, cell = s$cell, model = s$model, .before = 1)
}))

cat("=== % of cells that are marker-positive ===\n")
cat("Region = Head, Sex = Female, mean Age. Includes the zero-inflation component.\n\n")
print(as.data.frame(units_tab %>%
  dplyr::select(marker, cell, Diagnosis, rate_pct, lwr_pct, upr_pct,
                obs_rate_pct, islets, donors)), digits = 3)
```

```{r effect-sizes}
# Same fit expressed three ways: as a percentage in each group, as the
# difference in percentage points, and as a ratio.
eff <- units_tab %>%
  dplyr::select(marker, cell, model, Diagnosis, rate_pct, obs_rate_pct,
                obs_cells_per_islet, mean_denominator) %>%
  pivot_wider(names_from = Diagnosis,
              values_from = c(rate_pct, obs_rate_pct, obs_cells_per_islet,
                              mean_denominator)) %>%
  mutate(
    pct_ND        = rate_pct_ND,
    pct_T1D       = rate_pct_T1D,
    diff_pp       = rate_pct_T1D - rate_pct_ND,
    ratio         = rate_pct_T1D / rate_pct_ND,
    # Reconciles the rate with the raw per-islet counts: rate x cells available.
    pred_per_islet_ND  = rate_pct_ND  / 100 * mean_denominator_ND,
    pred_per_islet_T1D = rate_pct_T1D / 100 * mean_denominator_T1D,
    obs_per_islet_ND   = obs_cells_per_islet_ND,
    obs_per_islet_T1D  = obs_cells_per_islet_T1D) %>%
  dplyr::select(marker, cell, pct_ND, pct_T1D, diff_pp, ratio,
                pred_per_islet_ND, obs_per_islet_ND,
                pred_per_islet_T1D, obs_per_islet_T1D)

cat("=== effect in original units ===\n")
cat("diff_pp is percentage points; ratio is the marginal rate ratio.\n")
cat("The per-islet columns convert the rate back to counts using each group's\n")
cat("mean cell count - pred vs obs should agree closely, and a gap means the\n")
cat("reference-level prediction is not representative of the group average.\n\n")
print(as.data.frame(eff), digits = 3)
```

```{r ratio-ci}
# Marginal rate ratio with a confidence interval. predict() gives correct CIs
# for each group's rate, but the two are correlated, so the ratio needs a joint
# calculation. emmeans does this properly; the conditional-component ratio from
# the coefficient table is the fallback and was within ~7% of the marginal in
# every model here.
ratio_ci <- function(m, sub, den, per = PER) {
  out <- tryCatch({
    at <- list(Age_c = 0)
    if ("log_Islet.Cells_c" %in% all.vars(formula(m))) at$log_Islet.Cells_c <- 0
    emm <- emmeans::emmeans(m, ~ Diagnosis, at = at, data = sub,
                            offset = log(per), component = "response")
    ct  <- as.data.frame(emmeans::contrast(emmeans::regrid(emm, transform = "log"),
                                           "revpairwise", type = "response"))
    tibble(source = "emmeans (marginal)", ratio = ct$ratio[1],
           lwr = ct$asymp.LCL[1], upr = ct$asymp.UCL[1], p = ct$p.value[1])
  }, error = function(e) NULL)

  if (is.null(out)) {
    co <- summary(m)$coefficients$cond
    e <- co["DiagnosisT1D", "Estimate"]; s <- co["DiagnosisT1D", "Std. Error"]
    out <- tibble(source = "coef table (conditional)", ratio = exp(e),
                  lwr = exp(e - 1.96 * s), upr = exp(e + 1.96 * s),
                  p = co["DiagnosisT1D", "Pr(>|z|)"])
  }
  out
}

if (requireNamespace("emmeans", quietly = TRUE)) {
  cat("=== rate ratio with CI ===\n\n")
  print(as.data.frame(bind_rows(lapply(seq_len(nrow(spec)), function(k) {
    s <- spec[k, ]
    ratio_ci(get(s$model), subsets[[s$key]], s$den) %>%
      mutate(marker = s$marker, cell = s$cell, .before = 1)
  }))), digits = 4)
} else {
  cat("install.packages('emmeans') for marginal ratio CIs\n")
}
```

```{r sentences}
# Draft wording, so the numbers and the claim cannot drift apart.
cat("=== draft sentences ===\n\n")
for (k in seq_len(nrow(eff))) {
  e <- eff[k, ]
  dir <- if (e$ratio < 1) "lower" else "higher"
  cat(sprintf(
    "%s / %s cells: %.2f%% of %s cells positive in ND vs %.2f%% in T1D\n",
    e$marker, e$cell, e$pct_ND, e$cell, e$pct_T1D))
  cat(sprintf(
    "   %.2f-fold %s in T1D (%.2f percentage points); per islet %.3f -> %.3f cells\n\n",
    if (e$ratio < 1) 1 / e$ratio else e$ratio, dir, e$diff_pp,
    e$obs_per_islet_ND, e$obs_per_islet_T1D))
}
cat("Report the CI and p from the ratio table above, not from these lines.\n")
cat("For the beta-cell rows, add the restriction: the estimate covers donors\n")
cat("with residual beta mass. Four Ki-67 and six CD3 T1D donors had no\n")
cat("insulin+ cells anywhere and are excluded by the offset.\n")
```

```{r save}
write_csv(units_tab, "rate_models_units.csv")
write_csv(eff, "rate_models_effects.csv")
cat("\nSaved rate_models_units.csv and rate_models_effects.csv\n")
```

```{r}
sessionInfo()
```

